### 2020-12-17 manuscript/main.tex
start manuscript

\documentclass[10pt,letterpaper]{article}

    % General document formatting
    \usepackage[margin=1.5in]{geometry}
    \usepackage[utf8]{inputenc}
    \usepackage{authblk}
    \usepackage{bibentry}
    \usepackage{parskip}

    \title{Potential Novel Immunotarget Revealed in Mega Analysis of Colorectal Transcriptomes}

    \author[1]{Christopher H. Dampier}
    \author[2]{David F. Grabski}
    \author[1]{Matthew Devall}
    \author[3]{Virginia Diez-Obrero}
    \author[3]{Ferran Moratalla-Navarro}
    \author[4]{David Rekosh}
    \author[4]{Marie-Louise Hammarskjold}
    \author[5]{Sara K. Rasmussen}
    \author[3]{Victor Moreno}
    \author[1]{Graham Casey}

    \affil[1]{Center for Public Health Genomics, University of Virginia, Charlottesville, Virginia, USA}
    \affil[2]{Department of Surgery, University of Virginia, Charlottesville, Virginia, USA}
    \affil[3]{Catalan Institute of Oncology, Barcelona, Spain}
    \affil[4]{Department of Microbiology, Immunology and Cancer Biology, University of Virginia, Charlottesville, Virginia, USA}
    \affil[5]{Department of Surgery, Seattle Children's, Seattle, Washington, USA}

    \makeatletter
    \renewcommand{\maketitle}{
        \begingroup
            \setlength{\parindent}{0pt}
            \begin{flushleft}
                \LARGE\textbf{\@title}
                \newline
                \newline
                \normalsize\@author
            \end{flushleft}
        \endgroup
    }
    \makeatother

\begin{document}

\maketitle

\section*{Abstract}
\textbf{Background}: The background goes here.

\textbf{Methods}: The methods go here.

\textbf{Results}: The results go here.

\textbf{Conclusions}: The conclusions go here.

\section*{Introduction}

Colorectal cancer (CRC) is a common malignancy in the United States.
Early detection and effective treatment result in a 5-year relative survival of 90\% for patients diagnosed with local disease.
However, over a fifth of new cases are diagnosed with metastatic disease, which has a 5-year relative survival of only 14\% \cite{SEER2020}.
To improve overall survival of patients with distant disease, new therapeutic approaches are needed.

Immunotherapy in CRC is a promising new treatment option for patients with distant disease.
Immune checkpoint inhibition was shown to achieve durable responses with programmed cell death protein 1 (PD-1) antibody monotherapy and combination therapy with antibodies for PD-1 and cytotoxic T-lymphocyte-associated
protein 4 (CTLA-4) \cite{Le2015, Overman2018}.
However, response to immune checkpoint inhibition was limited to the subset of tumors with mismatch repair defficiency and high microsatellite instability, which constitute approximately 4\% of all metastatic CRC cases \cite{Ganesh2019}.
Strategies to improve tumor immunogenicity and increase cytotoxic T cell infiltration are required to help extend the benefit of immune therapies to the other 96\% of metastatic CRC.
Epitopes derived from endogenous retroviral peptides expressed in cancers may offer a solution.

Human endogenous retroviruses (HERVs) are proviral elements integrated into the human genome under conditions that render them incompetent of replication or infection.
Residual HERV elements in contemporary human genomes are the consequence of retroviral infections of germ cells millions of years ago.
Since then, and like the rest of the human genome, HERV elements have accumulated point mutations and segmental deletions.
These alterations in HERV sequences disrupt assembly of functional viral machinery.
Nevertheless, some HERV sequences can still be transcribed and translated, and a subset are processed as endogenous antigens and presented on major histocompatibility complex (MHC) class I molecules \cite{Boller1997, Rooney2015}.
In healthy tissues, DNA methylation and histone modification can silence transcription of HERV genes, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and lead to translation of neoantigens \cite{Menendez2004, Wiesner2015}.

The cancer specificity of putative HERV peptides makes them promising biomarkers for and targets of immune therapies.
A major challenge to the development of HERV based therapies is the absence of consensus on how to find HERV gene products.
Proviruses are generally absent from human gene, transcript, and protein annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \cite{ERVmap2018, Treangen2011}.
Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
Furthermore, HERVs are transposable elements that have generated many similar proviral integrations throughout the human genome.
The multitude of similar sequences across the genome impairs accurate high throughout read alignment.
As a result, the genomic locations and transcriptional abundances of most HERVs are not known with confidence.

Recently, new approaches have been developed to overcome traditional limitations in HERV annotation, quantification, and protein identification \cite{Attig2017, Attig2019, ERVmap2018, Telescope2019}.
Here, we extend advances in the characterization of HERV transcriptomes to discover novel CRC specific expression patterns.
We test for differential expression of HERV transcripts in a large cohort of healthy, tumor-adjacent normal, and tumor RNA sequencing data and identify HERV transcripts over expressed in tumor samples.
We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.

\section*{Methods}

We ...

\section*{Results}

We ...

\section*{Discussion}

We ...

\bibliography{main}
\bibliographystyle{vancouver}

\end{document}

### 2020-12-20 manuscript/main.tex
documentation
work on abstract

@@
 
-    % General document formatting
+    %% general document formatting
     \usepackage[margin=1.5in]{geometry}
@@
 
+    %% redefine maketitle function to customize title formatting
+    %https://tex.stackexchange.com/questions/85343/left-align-abstract-title-and-authors
+    %https://en.wikibooks.org/wiki/LaTeX/Fonts#Sizing_text
     \makeatletter
@@
 \section*{Abstract}
-\textbf{Background}: The background goes here.
+\textbf{Background}:
+Microsatellite unstable, mismatch repair deficient metastatic cancers can be treated with immune checkpoint inhibitors, but new therapeutic approaches are needed to extend the benefits of immune modulators to the majority of patients with microsatellite stable tumors.
+Cancer specific over expression of human endogenous retroviral elements could be exploited to enhance immune infiltration of microsatellite stable colorectal cancer and render these tumors more susceptible to immune checkpoint inhibition.
+To identify novel targets for immune mediated cancer therapies, we tested for cancer specific over expression of endogenous retroviral transcripts in a large cohort of colorectal transcriptomes.
 
-\textbf{Methods}: The methods go here.
+\textbf{Methods}:
+A recently curated RNA sequencing dataset composed of 1,139 human primary colorectal tissue samples divided into three cohorts was re-processed using a novel annotation of repeat element transcripts to quantify levels of repeat element expression in colorectal transcriptomes.
+A total of ### healthy, ### tumor-adjacent normal, and ### tumor samples were profiled with short read sequencing.
+Transcripts were quantified with Salmon using the GENCODE v36 transcriptome augmented to include unannotated exons from long terminal repeat elements classified as human endogenous retroviruses (HERVs) in the Dfam database.
+Associations between transcript expression and tissue phenotype were tested using DESeq2 in the R statistical language.
 
-\textbf{Results}: The results go here.
+\textbf{Results}:
+Out of 5,025 unannotated HERV genes measured in 834 samples, including ### healthy, ### tumor-adjacent normal, and ### tumor samples, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
+Over half of the 150 HERV genes with moderate expression were in the H (26) or K (51) classes.
+Tumor specific over expression was observed in 11 of the 150 HERV genes tested.
 

### 2020-12-23 manuscript/main.tex
finish methods

@@
 
-    \author[1]{Christopher H. Dampier}
-    \author[2]{David F. Grabski}
-    \author[1]{Matthew Devall}
-    \author[3]{Virginia Diez-Obrero}
-    \author[3]{Ferran Moratalla-Navarro}
-    \author[4]{David Rekosh}
-    \author[4]{Marie-Louise Hammarskjold}
-    \author[5]{Sara K. Rasmussen}
-    \author[3]{Victor Moreno}
-    \author[1]{Graham Casey}
+    \author[1]{Christopher H. Dampier, MD}
+    \author[2]{David F. Grabski, MD, PhD}
+    \author[1]{Matthew Devall, PhD}
+    \author[3]{Virginia Diez-Obrero, BS}
+    \author[3]{Ferran Moratalla-Navarro, MS}
+    \author[4]{David Rekosh, PhD}
+    \author[4]{Marie-Louise Hammarskjold, PhD}
+    \author[5]{Sara K. Rasmussen, MD, PhD}
+    \author[3]{Victor Moreno, MD, PhD}
+    \author[1,*]{Graham Casey, PhD}
 
@@
     \affil[5]{Department of Surgery, Seattle Children's, Seattle, Washington, USA}
+    \affil[*]{Correspondence: Graham Casey, PhD, Center for Public Health Genomics, MSB Room 3238, Department of Public Health Sciences, University of Virginia, P.O.
+Box 800717, Charlottesville, VA 22908-0717, gc8r@virginia.edu}
 
@@
 
+\newpage
+\section*{Synopsis}
+1 to 3 sentences, no more than 45 words
+
+\newpage
 \section*{Abstract}
@@
 A recently curated RNA sequencing dataset composed of 1,139 human primary colorectal tissue samples divided into three cohorts was re-processed using a novel annotation of repeat element transcripts to quantify levels of repeat element expression in colorectal transcriptomes.
-A total of ### healthy, ### tumor-adjacent normal, and ### tumor samples were profiled with short read sequencing.
+A total of NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples were profiled with short read sequencing.
 Transcripts were quantified with Salmon using the GENCODE v36 transcriptome augmented to include unannotated exons from long terminal repeat elements classified as human endogenous retroviruses (HERVs) in the Dfam database.
@@
 \textbf{Results}:
-Out of 5,025 unannotated HERV genes measured in 834 samples, including ### healthy, ### tumor-adjacent normal, and ### tumor samples, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
+Out of 5,025 unannotated HERV genes measured in 834 samples, including NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
 Over half of the 150 HERV genes with moderate expression were in the H (26) or K (51) classes.
@@
 
+\newpage
 \section*{Introduction}
-
 Colorectal cancer (CRC) is a common malignancy in the United States.
@@
 \section*{Methods}
+\subsection*{Sample Selection}
+Bulk RNA sequencing (RNA-seq) data from a recently curated set of 1,139 flash frozen colorectal tissue samples were re-processed and analyzed for the presence of HERV genes \cite{Dampier2020}.
+For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 834 colon biopsies (Cohort A) was used.
+Of those biopsies, 462 were healthy tissue, 61 were tumor adjacent normal tissue, and 311 were tumors.
+For validation of discovered associations, two independent cohorts were used.
+The first (Cohort B) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
+The second (Cohort C) included single end sequencing from 275 colon biopsies of which NUMBER were healthy, NUMBER were tumor adjacent, and NUMBER were tumor.
+An important characteristic of the discovery cohort (Cohort A) and the second of the two validation cohorts (Cohort C) was the independence of all tumor and tumor adjacent samples.
+To permit valid inference from regular linear models, only one tissue phenotype was used per subject.
 
-We ...
+Samples were gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a novel cohort of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \cite{Dampier2020, Diez-Obrero2020}.
+A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al \cite{Dampier2020}.
+RNA-seq reads for all samples were stored and analyzed in FASTQ format.
+For samples downloaded in BAM format, Biobambam2 v2.0.87 \cite{Tischler2014} was used to convert reads to FASTQ format.
+
+\subsection*{Human Endogenous Retrovirus Gene Annotation}
+HERVs were identified using the genomic coordinates (GRCh38) and feature names of exons and spliced transcripts of HERV elements from a recently published human cancer transcriptome annotated with the GENCODE v24 basic gene model \cite{Frankish2018} and a model of genome-wide repeat elements \cite{Attig2019}.
+The cancer transcriptome was assembled from RNA-seq reads of 24 subjects from each of 32 cancer types obtained from The Cancer Genome Atlas (TCGA), as previously described \cite{Attig2019}.
+Names and coordinates of repeat elements were generated using RepeatMasker \cite{Smit2015} configured with nhmmer \cite{Wheeler2013} in sensitive mode using the Dfam 2.0 library (v150923) \cite{Hubley2015}, as previously described \cite{Attig2017}.
+The method relies on profile hidden Markov models representing known human repeat families to classify repeat elements throughout the genome.
+
+Previously unannotated HERV genes were selected for the current study in multiple steps.
+A complete gene transfer format (GTF) \cite{GTF} annotation of the repeat element inclusive cancer transcriptome was downloaded from Attig et al. \cite{Attig2019}.
+The annotation was parsed with a custom Python script to produce a new annotation in BED12 \cite{BED12} format.
+Transcripts were selected based on three criteria: i) overlap of at least one exon with a long terminal repeat element, ii) classification of at least one exon as HERV derived, and iii) abscence of any exon overlapping an annotated GENCODE gene.
+All transcripts labeled with the same repeat element identifier (i.e. repeat element class along with genomic coordinates) were considered isoforms of the same HERV gene.
+Genomic coordinates were extracted for all constituent exons of selected transcripts to permit inference of splice sites and exclusion of intronic sequences.
+For display purposes, repeat element identifier genomic coordinates were converted to cytobands.
+
+\subsection*{Transcript Quantification}
+Transcript quantification was performed with Salmon v1.2.1 \cite{Patro2017} in mapping-based mode with the \verb|--gcBias| and \verb|--validateMappings| flags set.
+Inputs to the Salmon \verb|quant| command were RNA-seq reads in FASTQ format and the Salmon index.
+To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34  transcript sequences \cite{Frankish2018} and the list of unannotated HERV transcript sequences.
+GENCODE sequences were downloaded from the GENCODE FTP site \cite{GENCODE-transcripts}.
+HERV sequences were extracted from the GRCh38 reference genome \cite{ENCODE-GRCh38} using the BEDTools \cite{Quinlan2010} \verb|getfasta| command with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts.
+Following the recommendation of Salmon's authors \cite{SalmonDecoys}, contig sequences from the GRCh38 reference genome were subsequently added to the target transcript sequences as decoy sequences to generate a decoy-aware transcriptome for indexing.
+
+\subsection*{Gene Expression Analysis}
+Associations between HERV gene expression and tissue phenotype were tested using the \emph{Bioconductor} suite of genomics analysis packages \cite{bioc} in the R statistical programming language \cite{R}.
+Specifically, \emph{tximport} \cite{Soneson2015} was used to extract transcript length and abundance estimates from Salmon and summarize them to the gene level for input into \emph{DESeq2} \cite{Love2014}.
+Count matrices of HERV and GENCODE genes were prefiltered to select genes with a count of at least one read in at least half of the samples in each cohort.
+Factors to normalize counts for sequencing depth and transcript length were calculated with \emph{DESeq2}.
+Latent factors of non-biological variation were estimated with \emph{sva} \cite{sva}.
+Genes of interest for association testing were selected in several steps.
+
+First, counts of all genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \cite{Ritchie2015}.
+Second, adjusted counts of all annotated genes were filtered to retain only protein coding genes.
+Third, median and standard deviation of protein coding gene expression were calculated separately for each tissue phenotype in each cohort.
+Finally, adjusted counts of HERV genes were filtered to retain only HERVs expressed within a single standard deviation of median protein coding gene expression in at least one tissue phenotype.
+
+Counts of all protein coding genes and HERVs with sufficient expression were modeled with phenotype and latent factors as explanatory variables using \emph{DESeq2}.
+Significant associations between gene count and tissue phenotype were identified by testing whether the absolute value of the effect of phenotype was at least two fold change for a given gene.
+Multiple hypothesis testing correction was performed using a false discovery rate of 5\%.
 
@@
 
+Tumor specific overexpression ... antigenic potential
+cohort A
+then cohorts B and C
+then cohort A without Attig TCGA 48 + 18 (24 COAD tumor, 24 READ tumor, 12 COAD NAT, 6 READ NAT) and GTEx 12
+then MSI vs MSS
+then tumors with most HERV overexpression
+then immunogenic peptide prediction
+
 We ...
+Of 1,001,931 transcripts composed of 3,987,323 exons 16,336 transcripts ... 35,666 exons
+
+Healthy overexpression ... functional role
+cohort A
+then cohorts B and C
+then Grabski method
 
@@
 
-We ...
+HERVs could be used as immune targets or biomarkers
+HERVs may play a role in maintaining healthy epithelium
+Limitations
+Conclusion
+
+\section*{Acknowledgements}
+
+People we want to thank. Grant support.
 
@@
 
+\section*{Figure Legends}
+
+Legs
+
+\section*{Tables}
+
+Tabs
+
 \end{document}
+
+problems with AMA style file...
+https://tex.stackexchange.com/questions/55020/updated-ama-bst-file

### 2020-12-24 manuscript/main.tex
fix formatting

@@
-\documentclass[10pt,letterpaper]{article}
+\documentclass[11pt,letterpaper]{article}
 
-    %% general document formatting
-    \usepackage[margin=1.5in]{geometry}
+    %% load packages
+    \usepackage[margin=1in]{geometry}       %% document margins
     \usepackage[utf8]{inputenc}
-    \usepackage{authblk}
-    \usepackage{bibentry}
-    \usepackage{parskip}
+    \usepackage{setspace}                     %% to get \doublespacing command
+    \usepackage{authblk}                      %% to get \affil command
+    \usepackage{bibentry}                     %% bib formatting
+    \usepackage{parskip}                      %% space between paragraphs
+    \usepackage{fontspec}                     %% for \setmainfont command, will need xelatex to compile Arial
+    \usepackage[comma,super,sort&compress]{natbib}  %% https://latex.org/forum/viewtopic.php?t=8399
+    \usepackage{graphicx}                     %% for figures
+
+    %% general document formatting commands
+    \pagestyle{plain}                         %% page numbers in middle
+    %\linespread{1.67}                        %% baseline linespread = 1.2 (1.2*1.67~2)
+    \doublespacing
+    \setmainfont{Arial}                       %% https://latex.org/forum/viewtopic.php?t=25998
+    \setcitestyle{citesep={,}}
+    \graphicspath{{figs/}}                    %% set path to figures relative to this document
 
@@
     %% redefine maketitle function to customize title formatting
-    %https://tex.stackexchange.com/questions/85343/left-align-abstract-title-and-authors
-    %https://en.wikibooks.org/wiki/LaTeX/Fonts#Sizing_text
+        %% https://tex.stackexchange.com/questions/85343/left-align-abstract-title-and-authors
+        %% https://en.wikibooks.org/wiki/LaTeX/Fonts#Sizing_text
     \makeatletter
@@
 Early detection and effective treatment result in a 5-year relative survival of 90\% for patients diagnosed with local disease.
-However, over a fifth of new cases are diagnosed with metastatic disease, which has a 5-year relative survival of only 14\% \cite{SEER2020}.
+However, over a fifth of new cases are diagnosed with metastatic disease, which has a 5-year relative survival of only 14\% \citep{SEER2020}.
 To improve overall survival of patients with distant disease, new therapeutic approaches are needed.
@@
 Immune checkpoint inhibition was shown to achieve durable responses with programmed cell death protein 1 (PD-1) antibody monotherapy and combination therapy with antibodies for PD-1 and cytotoxic T-lymphocyte-associated
-protein 4 (CTLA-4) \cite{Le2015, Overman2018}.
-However, response to immune checkpoint inhibition was limited to the subset of tumors with mismatch repair defficiency and high microsatellite instability, which constitute approximately 4\% of all metastatic CRC cases \cite{Ganesh2019}.
+protein 4 (CTLA-4) \citep{Le2015, Overman2018}.
+However, response to immune checkpoint inhibition was limited to the subset of tumors with mismatch repair defficiency and high microsatellite instability, which constitute approximately 4\% of all metastatic CRC cases \citep{Ganesh2019}.
 Strategies to improve tumor immunogenicity and increase cytotoxic T cell infiltration are required to help extend the benefit of immune therapies to the other 96\% of metastatic CRC.
@@
 These alterations in HERV sequences disrupt assembly of functional viral machinery.
-Nevertheless, some HERV sequences can still be transcribed and translated, and a subset are processed as endogenous antigens and presented on major histocompatibility complex (MHC) class I molecules \cite{Boller1997, Rooney2015}.
-In healthy tissues, DNA methylation and histone modification can silence transcription of HERV genes, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and lead to translation of neoantigens \cite{Menendez2004, Wiesner2015}.
+Nevertheless, some HERV sequences can still be transcribed and translated, and a subset are processed as endogenous antigens and presented on major histocompatibility complex (MHC) class I molecules \citep{Boller1997, Rooney2015}.
+In healthy tissues, DNA methylation and histone modification can silence transcription of HERV genes, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and lead to translation of neoantigens \citep{Menendez2004, Wiesner2015}.
 
@@
 A major challenge to the development of HERV based therapies is the absence of consensus on how to find HERV gene products.
-Proviruses are generally absent from human gene, transcript, and protein annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \cite{ERVmap2018, Treangen2011}.
+Proviruses are generally absent from human gene, transcript, and protein annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
 Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
@@
 
-Recently, new approaches have been developed to overcome traditional limitations in HERV annotation, quantification, and protein identification \cite{Attig2017, Attig2019, ERVmap2018, Telescope2019}.
+Recently, new approaches have been developed to overcome traditional limitations in HERV annotation, quantification, and protein identification \citep{Attig2017, Attig2019, ERVmap2018, Telescope2019}.
 Here, we extend advances in the characterization of HERV transcriptomes to discover novel CRC specific expression patterns.
@@
 \subsection*{Sample Selection}
-Bulk RNA sequencing (RNA-seq) data from a recently curated set of 1,139 flash frozen colorectal tissue samples were re-processed and analyzed for the presence of HERV genes \cite{Dampier2020}.
+Bulk RNA sequencing (RNA-seq) data from a recently curated set of 1,139 flash frozen colorectal tissue samples were re-processed and analyzed for the presence of HERV genes \citep{Dampier2020}.
 For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 834 colon biopsies (Cohort A) was used.
@@
 
-Samples were gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a novel cohort of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \cite{Dampier2020, Diez-Obrero2020}.
-A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al \cite{Dampier2020}.
+Samples were gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a novel cohort of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
+A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al \citep{Dampier2020}.
 RNA-seq reads for all samples were stored and analyzed in FASTQ format.
-For samples downloaded in BAM format, Biobambam2 v2.0.87 \cite{Tischler2014} was used to convert reads to FASTQ format.
+For samples downloaded in BAM format, Biobambam2 v2.0.87 \citep{Tischler2014} was used to convert reads to FASTQ format.
 
 \subsection*{Human Endogenous Retrovirus Gene Annotation}
-HERVs were identified using the genomic coordinates (GRCh38) and feature names of exons and spliced transcripts of HERV elements from a recently published human cancer transcriptome annotated with the GENCODE v24 basic gene model \cite{Frankish2018} and a model of genome-wide repeat elements \cite{Attig2019}.
-The cancer transcriptome was assembled from RNA-seq reads of 24 subjects from each of 32 cancer types obtained from The Cancer Genome Atlas (TCGA), as previously described \cite{Attig2019}.
-Names and coordinates of repeat elements were generated using RepeatMasker \cite{Smit2015} configured with nhmmer \cite{Wheeler2013} in sensitive mode using the Dfam 2.0 library (v150923) \cite{Hubley2015}, as previously described \cite{Attig2017}.
+HERVs were identified using the genomic coordinates (GRCh38) and feature names of exons and spliced transcripts of HERV elements from a recently published human cancer transcriptome annotated with the GENCODE v24 basic gene model \citep{Frankish2018} and a model of genome-wide repeat elements \citep{Attig2019}.
+The cancer transcriptome was assembled from RNA-seq reads of 24 subjects from each of 32 cancer types obtained from The Cancer Genome Atlas (TCGA), as previously described \citep{Attig2019}.
+Names and coordinates of repeat elements were generated using RepeatMasker \citep{Smit2015} configured with nhmmer \citep{Wheeler2013} in sensitive mode using the Dfam 2.0 library (v150923) \citep{Hubley2015}, as previously described \citep{Attig2017}.
 The method relies on profile hidden Markov models representing known human repeat families to classify repeat elements throughout the genome.
@@
 Previously unannotated HERV genes were selected for the current study in multiple steps.
-A complete gene transfer format (GTF) \cite{GTF} annotation of the repeat element inclusive cancer transcriptome was downloaded from Attig et al. \cite{Attig2019}.
-The annotation was parsed with a custom Python script to produce a new annotation in BED12 \cite{BED12} format.
+A complete gene transfer format (GTF) \citep{GTF} annotation of the repeat element inclusive cancer transcriptome was downloaded from Attig et al. \citep{Attig2019}.
+The annotation was parsed with a custom Python script to produce a new annotation in BED12 \citep{BED12} format.
 Transcripts were selected based on three criteria: i) overlap of at least one exon with a long terminal repeat element, ii) classification of at least one exon as HERV derived, and iii) abscence of any exon overlapping an annotated GENCODE gene.
@@
 \subsection*{Transcript Quantification}
-Transcript quantification was performed with Salmon v1.2.1 \cite{Patro2017} in mapping-based mode with the \verb|--gcBias| and \verb|--validateMappings| flags set.
+Transcript quantification was performed with Salmon v1.2.1 \citep{Patro2017} in mapping-based mode with the \verb|--gcBias| and \verb|--validateMappings| flags set.
 Inputs to the Salmon \verb|quant| command were RNA-seq reads in FASTQ format and the Salmon index.
-To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34  transcript sequences \cite{Frankish2018} and the list of unannotated HERV transcript sequences.
-GENCODE sequences were downloaded from the GENCODE FTP site \cite{GENCODE-transcripts}.
-HERV sequences were extracted from the GRCh38 reference genome \cite{ENCODE-GRCh38} using the BEDTools \cite{Quinlan2010} \verb|getfasta| command with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts.
-Following the recommendation of Salmon's authors \cite{SalmonDecoys}, contig sequences from the GRCh38 reference genome were subsequently added to the target transcript sequences as decoy sequences to generate a decoy-aware transcriptome for indexing.
+To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34  transcript sequences \citep{Frankish2018} and the list of unannotated HERV transcript sequences.
+GENCODE sequences were downloaded from the GENCODE FTP site \citep{GENCODE-transcripts}.
+HERV sequences were extracted from the GRCh38 reference genome \citep{ENCODE-GRCh38} using the BEDTools \citep{Quinlan2010} \verb|getfasta| command with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts.
+Following the recommendation of Salmon's authors \citep{SalmonDecoys}, contig sequences from the GRCh38 reference genome were subsequently added to the target transcript sequences as decoy sequences to generate a decoy-aware transcriptome for indexing.
 
 \subsection*{Gene Expression Analysis}
-Associations between HERV gene expression and tissue phenotype were tested using the \emph{Bioconductor} suite of genomics analysis packages \cite{bioc} in the R statistical programming language \cite{R}.
-Specifically, \emph{tximport} \cite{Soneson2015} was used to extract transcript length and abundance estimates from Salmon and summarize them to the gene level for input into \emph{DESeq2} \cite{Love2014}.
+Associations between HERV gene expression and tissue phenotype were tested using the \emph{Bioconductor} suite of genomics analysis packages \citep{bioc} in the R statistical programming language \citep{R}.
+Specifically, \emph{tximport} \citep{Soneson2015} was used to extract transcript length and abundance estimates from Salmon and summarize them to the gene level for input into \emph{DESeq2} \citep{Love2014}.
 Count matrices of HERV and GENCODE genes were prefiltered to select genes with a count of at least one read in at least half of the samples in each cohort.
 Factors to normalize counts for sequencing depth and transcript length were calculated with \emph{DESeq2}.
-Latent factors of non-biological variation were estimated with \emph{sva} \cite{sva}.
+Latent factors of non-biological variation were estimated with \emph{sva} \citep{sva}.
 Genes of interest for association testing were selected in several steps.
 
-First, counts of all genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \cite{Ritchie2015}.
+First, counts of all genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \citep{Ritchie2015}.
 Second, adjusted counts of all annotated genes were filtered to retain only protein coding genes.
@@
 \section*{Results}
+\subsection*{Tumor Specific Over-expression}
+To overcome the difficulty of defining the precise location of HERV sequences in the human genome, we used as a starting point a recently published, de novo assembled, HERV-aware annotation of the human cancer transcriptome \citep{Attig2019}.
+The complete annotation included over 1 million transcripts, including previously unannotated repeat sequences.
+To discover patterns of novel, cancer-specific HERV expression, we identified the subset of transcripts that were HERV-derived and without annotated exons, as detailed in the Methods.
+There were 16,336 transcripts generated from 5,025 HERV genes in the reference transcriptome of Attig et al. (Figure \ref{fig:ele_sum}A).
+
+We test for differential expression of HERV transcripts in a large cohort of healthy, tumor-adjacent normal, and tumor RNA sequencing data and identify HERV transcripts over expressed in tumor samples.
+We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.
 
@@
 
+\section*{Figures}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=0.9\textwidth]{herv-element-sum_A}
+  }
+  \caption{Summary of unannotated human endogenous retroviral sequences}
+  \label{fig:ele_sum}
+\end{figure}
+
 \end{document}

### 2020-12-24 scripts/results.R
qq plot

@@ -782,3 +782,20 @@ main()
 
 # target <- "/scratch/chd5n/test-fig.pdf"
 # ggsave(target, p, device="pdf", width=15, height=4, units="in")
+
+restabs <- load_res_tables("herv")
+y <- list()
+y[[1]] <- -log10(restabs[[1]]$pvalue)
+y[[2]] <- -log10(restabs[[2]]$pvalue)
+y[[3]] <- -log10(restabs[[3]]$pvalue)
+x <- -log10(qunif(ppoints(length(y[[1]]))))
+
+xlab <- expression("Expected -log"["10"]*"("*italic("p")*")")
+ylab <- expression("Observed -log"["10"]*"("*italic("p")*")")
+
+pdf("/scratch/chd5n/herv/results/plots/herv-qq_A.pdf", height=3, width=9)
+par(mfrow=c(1,3))
+qqplot(x, y[[1]], xlab=xlab, ylab=ylab, main="NATvHLT", col="blue")
+qqplot(x, y[[3]], xlab=xlab, ylab=ylab, main="TUMvNAT", col="blue")
+qqplot(x, y[[2]], xlab=xlab, ylab=ylab, main="TUMvHLT", col="blue")
+dev.off()

### 2020-12-25 scripts/res_corr.R
modularize results

@@ -1,0 +1,54 @@
+## to be sourced in results.R
+
+
+plot_cor <- function(x, y, label, r) {
+    df <- data.frame(herv=x, field=y)
+    ggp_title <- "Pearson Correlation"
+    ggp_subtitle <- label
+    ggp_xlab <- "L2FC from HERV Study"
+    ggp_ylab <- "L2FC from Field Effect Study"
+    txt <- paste("R =", round(r, 2))
+    ggp <- ggplot(df, aes(x=herv, y=field)) +
+        geom_point(size=0.6, stroke=0, shape=16, alpha=0.5) +
+        labs(
+            title=element_blank(),
+            subtitle=ggp_subtitle,
+            x=ggp_xlab,
+            y=ggp_ylab
+        ) +
+        annotate("text", x=(min(df$herv) + 1), y=0.95*max(df$field), label=txt,
+            size=2.5) +
+        ggp_theme_default
+    return(ggp)
+}
+
+
+calc_cor <- function(l1, l2, method="pearson") {
+    cat("Checking correlation with prior study\n")
+    r <- list()
+    p <- list()
+    for (i in seq_len(length(l1))) {
+        intsx <- intersect(l1[[i]]$gene_id, l2[[i]]$gene_id)
+        x <- l1[[i]][l1[[i]]$gene_id %in% intsx, "log2FoldChange"]
+        y <- l2[[i]][l2[[i]]$gene_id %in% intsx, "log2FoldChange"]
+        label <- sub("CRC", "TUM", unlist(strsplit(names(l1)[i], "_"))[3])
+        r[[label]] <- cor(x, y, method=method)
+        p[[label]] <- plot_cor(x, y, label, r[[label]])
+    }
+    return(setNames(list(r, p), c("cor", "fig")))
+}
+
+
+write_cor <- function(plotlist, target) {
+    p_width <- 9
+    p_height <- 3
+    fig <- cowplot::plot_grid(
+        plotlist[["NATvHLT"]], plotlist[["TUMvNAT"]], plotlist[["TUMvHLT"]],
+        nrow=1, ncol=3,
+        labels=c("A", "B", "C")
+    )
+    pdf(file=target, width=p_width, height=p_height)
+    print(fig)
+    dev.off()
+    cat("figure written to", target, "\n")
+}

### 2020-12-25 scripts/res_hervsum.R

@@ -1,0 +1,129 @@
+## to be sourced in results.R
+
+
+count_herv_ids <- function(gtf, ids) {
+    cat("Counting total herv transcripts\n")
+    cmd = paste(
+        paste("zcat", gtf),
+        "awk -v FS=\"\\t\" '!($0~/^#/) && $3 == \"transcript\" {print $0}'",
+        "wc -l",
+        sep=" | "
+    )
+    x1 <- as.numeric(system(cmd, intern=TRUE))
+    cat("Counting select herv ids\n")
+    df <- data.frame(readr::read_tsv(ids))
+    x2 <- nrow(df)
+    x3 <- length(unique(df$herv_id))
+    nms <- c("Total Transcripts", "Select Transcripts", "HERV Genes")
+    return(setNames(list(x1, x2, x3), nms))
+}
+
+
+count_herv_cats <- function(cv) {
+    nms <- base::unlist(
+        base::lapply(
+            base::lapply(
+                strsplit(cv, "[|]"), unpack_id
+            ),
+            head, 1
+        ),
+        use.names=FALSE
+    )
+    cats <- c(
+        "HERVE", "HERVH", "HERVK", "HERVL"
+    )
+    cnts <- list()
+    for (e in cats) {
+        cntr <- 0
+        for (i in nms) {
+            if (i == e) {
+                cntr <- cntr + 1
+            }
+        }
+        cnts[[e]] <- cntr
+    }
+    cnts[["Other"]] <- length(nms) - sum(unlist(cnts))
+    return(cnts)
+}
+
+
+fill_params <- function(l, params) {
+    if ("Total Transcripts" %in% names(l)) {
+        params[["ggp_title"]] <- "Number of HERV Elements Analyzed"
+        n <- format(l[["Total Transcripts"]], big.mark=",")
+        params[["ggp_sub"]] <- paste("From", n, "starting transcripts")
+        l <- l[names(l) != "Total Transcripts"]
+    } else {
+        params[["ggp_title"]] <- "Classes of HERV Genes Tested"
+        params[["ggp_sub"]] <- element_blank()
+    }
+    params[["ggp_ylab"]] <- "Counts"
+    return(setNames(list(l, params), c("l", "params")))
+}
+
+
+format_data <- function(l) {
+    df <- data.frame(base::t(data.frame(l, check.names=FALSE)))
+    df[, 2] <- rownames(df)
+    colnames(df) <- c("counts", "categories")
[…]

### 2020-12-25 scripts/res_utilities.R

@@ -1,0 +1,100 @@
+## to be sourced in results.R
+
+
+check_dir <- function(dir) {
+    ## check if intended output directory exists and create if it does not
+    d <- "/"
+    for (e in unlist(strsplit(dir, "/"))[2:length(unlist(strsplit(dir, "/")))]) {
+        d <- paste0(d, e, "/")
+        if (file.exists(d)) {
+            cat(d, "exists\n")
+        } else {
+            cat(d, "does not exist; creating now\n")
+            dir.create(d)
+        }
+    }
+}
+
+
+set_paths <- function(in_path, in_name) {
+    targets <- list()
+    for (e in in_name) {
+        target <- paste0(in_path, e)
+        n <- substr(e, 1, (nchar(e) - 4))
+        if (file.exists(target)) {
+            targets[[n]] <- target
+        } else {
+            cat("problem locating file", target, "\n")
+        }
+    }
+    return(targets)
+}
+
+
+load_res_tables <- function(study, scope="all") {
+    ## scope can be "all" or "hrv"
+    ## order genes by gene_id instead of rank so that order is consistent across
+    ## different results tables to facilitate comparisons across tables
+    cat("Loading results tables\n")
+    if (study == "herv") {
+        in_path <- Sys.getenv("table_dir")
+        if (cohort == "B") {
+            in_name <- c(
+                paste0("res_B_CRCvNAT_", scope, ".tsv")
+            )
+        } else {
+            in_name <- c(
+                paste0("res_", cohort, "_NATvHLT_", scope, ".tsv"),
+                paste0("res_", cohort, "_CRCvHLT_", scope, ".tsv"),
+                paste0("res_", cohort, "_CRCvNAT_", scope, ".tsv")
+            )
+        }
+    } else {
+        in_path <- "/nv/t74/cphgdesk/share/cphg_CaseyLab/Papers/Field_Effect/tables/"
+        in_name <- c(
+            "p-nat-hlt-ResTab.csv",
+            "p-crc-hlt-ResTab.csv",
+            "p-crc-nat-ResTab.csv"
+        )
+    }
+    targets <- set_paths(in_path, in_name)
+    tables <- list()
+    for (i in names(targets)) {
+        target <- targets[[i]]
+        if (study == "herv") {
+            df <- data.frame(readr::read_tsv(target))
[…]

### 2020-12-25 scripts/res_volcano.R

@@ -1,0 +1,149 @@
+## to be sourced in results.R
+
+
+shrink_effect <- function(dgeobj) {
+    ## apparently cannot use apeglm due to third contrast not being present in
+    ## resultsNames(dds) and apeglm not taking contrast argument; therefore,
+    ## prefer ashr with res object provided instead of coef or contrast argument
+    ##
+    ## incidentally, the shrinkage has no apparent effect on the plot
+    cat("Shrinking effect sizes for visualization\n")
+    cons <- list()
+    m <- combn(levels(colData(dgeobj$dds)$phenotype), 2)
+    for (i in seq_len(ncol(m))) {
+        cons[[i]] <- c("phenotype", rev(m[, i]))
+        names(cons) <- c(
+            names(cons)[seq_len(length(cons) - 1)],
+            paste(rev(m[, i]), collapse="v")
+        )
+    }
+    dfs <- list()
+    for (i in seq_len(length(cons))) {
+        res <- dgeobj$res[[i]][rownames(dgeobj$dds), ]
+        tmp <- lfcShrink(
+            dgeobj$dds,
+            res=res,
+            type="ashr"
+        )
+        if (all(res$gene_id == tmp$gene_id)) {
+            res$group <- "Other"
+            idx <- res$padj < 0.05 & grepl("HERV", res$gene_id) & tmp$log2FoldChange < 0
+            res[idx, "group"] <- "HERV Down"
+            idx <- res$padj < 0.05 & grepl("HERV", res$gene_id) & tmp$log2FoldChange > 0
+            res[idx, "group"] <- "HERV Up"
+            dfs[[i]] <- data.frame(
+                gene_id=res$gene_id,
+                gene_name=res$gene_name,
+                herv=grepl("HERV", res$gene_id),
+                effect=tmp$log2FoldChange,
+                pval=res$pvalue,
+                logp=-log10(res$pvalue),
+                padj=res$padj,
+                group=res$group
+            )
+        } else {
+            cat("Gene IDs are not equal\n")
+        }
+    }
+    names(dfs) <- names(cons)
+    return(dfs)
+}
+
+
+plot_volcano <- function(df, label) {
+    ## repeated calls to geom_point ensure HERV point is last point plotted
+    ggp_subtitle <- label
+    ggp_xlab <- expression("Log"["2"]*" Fold Change")
+    ggp_ylab <- expression("-log"["10"]*"("*italic("P")*")")
[…]

### 2020-12-25 scripts/results.R

@@ -29,6 +29,7 @@ library(ggplot2)
 library(cowplot)
 library(reshape2)
 library(dplyr)
+library(bacon)
 
 
 args <- commandArgs(trailingOnly=TRUE)
@@ -36,237 +37,12 @@ cohort <- args[2]
 cat(paste("Cohort:", cohort, "\n"))
 
 
-ggp_theme_default <- theme(
-    panel.background=element_rect(fill="white"),
-    panel.grid.major=element_line(color="white"),
-    panel.grid.minor=element_line(color="white"),
-    plot.background=element_rect(fill="white"),
-    plot.margin=margin(t=1, r=1, b=1, l=1, unit="lines"),
-    plot.title=element_text(size=12, face="bold", hjust=0.5),
-    plot.subtitle=element_text(size=10, face="plain", hjust=0.5),
-    axis.title.x=element_text(size=10, face="plain"),
-    axis.title.y=element_text(size=10, face="plain"),
-    axis.text.x=element_text(size=9, face="plain"),
-    axis.text.y=element_text(size=9, face="plain"),
-    # axis.ticks.x=element_none(),
-    # axis.ticks.y=element_none(),
-    axis.line.x.bottom=element_line(),
-    axis.line.y.left=element_line(),
-    legend.key=element_rect(fill="white"),
-    legend.position="top",
-    legend.title=element_blank(),
-    strip.background=element_rect(fill="black"),
-    strip.text=element_text(colour="white")
-)
-
-
-ggp_theme_col <- ggp_theme_default +
-    theme(
-        plot.margin=margin(t=0.5, r=0.5, b=0.5, l=0.5, unit="lines"),
-        plot.title=element_text(size=12, face="plain", hjust=0.5),
-        axis.text.x=element_text(size=10, face="plain", angle=45, hjust=1),
-        axis.text.y=element_text(size=10, face="plain"),
-        axis.line.x.bottom=element_blank(),
-        axis.line.y.left=element_blank(),
-        legend.position="none"
-    )
-
-
-ggp_theme_vol <- ggp_theme_default +
-    theme(
-        plot.margin=margin(t=0.25, r=0.25, b=0.25, l=0.25, unit="lines"),
-        legend.key=element_blank(),
-        legend.key.size=unit(0.001, "cm"),
-        legend.key.width=unit(0.001, "cm"),
-        legend.position=c(0.18, 0.9),
-        legend.spacing.x=unit(0.001, "cm"),
[…]

### 2020-12-25 manuscript/main.tex
minor progress on results

@@
 \subsection*{Sample Selection}
-Bulk RNA sequencing (RNA-seq) data from a recently curated set of 1,139 flash frozen colorectal tissue samples were re-processed and analyzed for the presence of HERV genes \citep{Dampier2020}.
-For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 834 colon biopsies (Cohort A) was used.
+Bulk RNA sequencing (RNA-seq) from a recently curated set of 1,139 flash frozen colorectal tissue samples was re-processed and analyzed for the presence of unannotated HERV genes \citep{Dampier2020}.
+For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 834 colon biopsies was used.
 Of those biopsies, 462 were healthy tissue, 61 were tumor adjacent normal tissue, and 311 were tumors.
+These samples are referred to as the discovery cohort.
+
 For validation of discovered associations, two independent cohorts were used.
-The first (Cohort B) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
-The second (Cohort C) included single end sequencing from 275 colon biopsies of which NUMBER were healthy, NUMBER were tumor adjacent, and NUMBER were tumor.
-An important characteristic of the discovery cohort (Cohort A) and the second of the two validation cohorts (Cohort C) was the independence of all tumor and tumor adjacent samples.
+The first (validation \#1) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
+The second (validation \#2) included single end sequencing from 275 colon biopsies of which NUMBER were healthy, NUMBER were tumor adjacent, and NUMBER were tumor.
+An important characteristic of the discovery cohort and the validation \#2 cohort was the independence of all tumor and tumor adjacent samples.
 To permit valid inference from regular linear models, only one tissue phenotype was used per subject.
 
-Samples were gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a novel cohort of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
-A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al \citep{Dampier2020}.
+Samples were gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a recently generated dataset of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
+A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al. \citep{Dampier2020}.
 RNA-seq reads for all samples were stored and analyzed in FASTQ format.
@@
 \section*{Results}
-\subsection*{Tumor Specific Over-expression}
-To overcome the difficulty of defining the precise location of HERV sequences in the human genome, we used as a starting point a recently published, de novo assembled, HERV-aware annotation of the human cancer transcriptome \citep{Attig2019}.
-The complete annotation included over 1 million transcripts, including previously unannotated repeat sequences.
-To discover patterns of novel, cancer-specific HERV expression, we identified the subset of transcripts that were HERV-derived and without annotated exons, as detailed in the Methods.
-There were 16,336 transcripts generated from 5,025 HERV genes in the reference transcriptome of Attig et al. (Figure \ref{fig:ele_sum}A).
+\subsection*{Over-Expression of Novel HERV Genes in Tumors}
+To define an inclusive set of non-canonical HERV elements against which to compare CRC transcripts, we started with a recently published model of the human cancer transcriptome assembled \emph{de novo} from TCGA samples and annotated with \emph{RepeatMasker} using the Dfam database \citep{Attig2019}.
+The complete annotation included over 1 million transcripts, 416,862 of which overlapped long terminal repeat sequences (i.e. the repeat sequences that flank HERV genes).
+To discover novel, cancer-specific HERV expression, we identified a subset of transcripts composed of exons without GENCODE annotations, at least one of which was classifed as a HERV element.
+There were 16,336 transcripts composed of 35,666 exons from 5,025 genes that met our criteria (Figure \ref{fig:ele_sum}A).
+To estimate gene abundance, RNA-seq reads from a discovery cohort of 834 colorectal samples were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
+A total of 150 unannotated HERV genes were expressed at levels within a standard deviation of canonical protein coding genes.
+Approximately half were from classes K and H, two of the better characterized classes of HERV (Figure \ref{fig:ele_sum}B).
+
+To identify HERV genes that could encode tumor specific neoantigens, we compared expression levels in tumor samples to levels in healthy samples and histologically normal tumor adjacent samples.
 
@@
 
-We ...
-Of 1,001,931 transcripts composed of 3,987,323 exons 16,336 transcripts ... 35,666 exons
+\subsection*{HERV Genes in Healthy Tissue}
 

### 2020-12-26 manuscript/main.tex
fix authors, first fig
plan for fig2

@@
 The complete annotation included over 1 million transcripts, 416,862 of which overlapped long terminal repeat sequences (i.e. the repeat sequences that flank HERV genes).
-To discover novel, cancer-specific HERV expression, we identified a subset of transcripts composed of exons without GENCODE annotations, at least one of which was classifed as a HERV element.
+We aimed to discover expression of novel, cancer-specific HERV sequences, so we selected a subset of transcripts composed of exons without GENCODE annotations, at least one of which was classifed as a HERV element.
 There were 16,336 transcripts composed of 35,666 exons from 5,025 genes that met our criteria (Figure \ref{fig:ele_sum}A).
-To estimate gene abundance, RNA-seq reads from a discovery cohort of 834 colorectal samples were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
+To estimate gene abundance in colorectal transciptomes, RNA-seq reads from a discovery cohort of 834 colorectal samples were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
 A total of 150 unannotated HERV genes were expressed at levels within a standard deviation of canonical protein coding genes.
@@
 
-To identify HERV genes that could encode tumor specific neoantigens, we compared expression levels in tumor samples to levels in healthy samples and histologically normal tumor adjacent samples.
+To identify HERV genes that could encode tumor specific neoantigens, we compared expression levels in tumor samples to levels in healthy samples and tumor adjacent, histologically normal samples.
+We have previously shown that a transcriptomic field effect in the setting of CRC can mask cancer specificity of some genes when tumor transcripts are compared to tumor adjacent normal transcripts \citep{Dampier2020}.
+Without \emph{a priori} knowledge of whether HERV genes are likely to be affected by the transcriptomic field effect, we compared tumor samples to healthy samples and tumor adjacent samples separately.
+Out of 150 HERV genes tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (Figure \ref{fig:main_find}A,B, red points).
+We intersected these two lists to identify 11 HERV genes over-expressed in tumors relative to both healthy and tumor adjacent samples (Figure \ref{fig:main_find}C).
+These 11 HERV genes could serve as tumor biomarkers and may also encode proteins unknown to the adaptive immune system that could serve as targets for immune therapy with lower likelihood of off target effects.
+
+To confirm the significance of the 11 HERV genes, we sought multiple validations of our methods and results. First, we verified the consistency of HERV vs field effect.
+Validations...
 
@@
 \bibliography{main}
-\bibliographystyle{vancouver}
+\bibliographystyle{myvancouver}
 
@@
 
-problems with AMA style file...
+problems with AMA style file
 https://tex.stackexchange.com/questions/55020/updated-ama-bst-file
+i modified vancouver.bst to get proper author lists

### 2020-12-26 scripts/results.R

@@ -29,7 +29,6 @@ library(ggplot2)
 library(cowplot)
 library(reshape2)
 library(dplyr)
-library(bacon)
 
 
 args <- commandArgs(trailingOnly=TRUE)

### 2020-12-26 manuscript/main.tex
validation

@@
 There were 16,336 transcripts composed of 35,666 exons from 5,025 genes that met our criteria (Figure \ref{fig:ele_sum}A).
-To estimate gene abundance in colorectal transciptomes, RNA-seq reads from a discovery cohort of 834 colorectal samples were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
+To estimate gene abundance in colorectal transciptomes, RNA-seq reads from a discovery cohort of 834 colorectal samples (\textbf{Supplementary Table 1}) were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
 A total of 150 unannotated HERV genes were expressed at levels within a standard deviation of canonical protein coding genes.
@@
 
-To confirm the significance of the 11 HERV genes, we sought multiple validations of our methods and results. First, we verified the consistency of HERV vs field effect.
-Validations...
+To confirm the significance of the 11 HERV genes, we sought multiple validations of our methods and results.
+First, we verified the consistency of relative protein coding gene expression between tissue phenotypes in the present study and a prior study of the same samples.
+For the comparisons between tumor samples and healthy and tumor adjacent samples, the Pearson correlations coefficients were 0.95 and 0.97, respectively (\textbf{Supplementary Figure 1}).
+The high concordance of effect sizes between this study and the prior, which employed similar methods to estimate gene abundance but did not consider HERV transcripts, indicated expansion of the reference transcriptome to include unannotated HERV sequences did not compromise estimates of relative abundance.
 
-We test for differential expression of HERV transcripts in a large cohort of healthy, tumor-adjacent normal, and tumor RNA sequencing data and identify HERV transcripts over expressed in tumor samples.
+Next, we
+
+validation #1
+validation #2
+
+restabs <- list()
+for (cohort in c("A", "B", "C")) {
+    restabs[[cohort]] <- load_res_tables("herv", scope="hrv")
+}
+
+idx <- lapply(restabs, fun <- function(x1) {
+    idx1 <- lapply(x1, fun <- function(x2) {
+        a <- x2$padj < 0.05
+        b <- x2$log2FoldChange > 0
+        c <- a & b
+        return(c)
+    })
+    return(idx1)
+})
+
+l <- list()
+for (n1 in names(restabs)) {
+    l[[n1]] <- list()
+    for (n2 in names(restabs[[n1]])) {
+        l[[n1]][[n2]] <- restabs[[n1]][[n2]][idx, "gene_id"]
+    }
+}
+
+l <- lapply(restabs, fun <- function(x) {
+    for (i in seq_len(length(x))) {
+        x[[i]]
+    }
+})
+
+tophervs <- restabs[["TUMvHLT"]][idx, "gene_id"]
+
+cohort <- "A"
+restabsA <- load_res_tables("herv", scope="hrv")
+cohort <- "B"
+restabsB <- load_res_tables("herv", scope="hrv")
+cohort <- "C"
+restabsC <- load_res_tables("herv", scope="hrv")
+
+idxA <- lapply(restabsA, fun <- function(x) {
+    a <- x$padj < 0.05
+    b <- x$log2FoldChange > 0
+    c <- a & b
+    return(c)
+})
+
+conds[[1]] <- restabs[["NATvHLT"]]$padj < 0.05
+conds[[2]] <- restabs[["NATvHLT"]]$log2FoldChange > 0
+
 We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.
 
-Tumor specific overexpression ... antigenic potential
-cohort A
 then cohorts B and C

### 2020-12-27 manuscript/main.tex
overhall to include validation of results

@@
 For validation of discovered associations, two independent cohorts were used.
-The first (validation \#1) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
-The second (validation \#2) included single end sequencing from 275 colon biopsies of which NUMBER were healthy, NUMBER were tumor adjacent, and NUMBER were tumor.
-An important characteristic of the discovery cohort and the validation \#2 cohort was the independence of all tumor and tumor adjacent samples.
+The first (independent cohort \#1) included single end sequencing from 275 colon biopsies of which NUMBER were healthy, NUMBER were tumor adjacent, and NUMBER were tumor.
+The second (independent cohort \#2) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
+An important characteristic of the discovery cohort and independent cohort \#1 was the independence of all tumor and tumor adjacent samples.
 To permit valid inference from regular linear models, only one tissue phenotype was used per subject.
@@
 
-restabs <- list()
-for (cohort in c("A", "B", "C")) {
-    restabs[[cohort]] <- load_res_tables("herv", scope="hrv")
-}
-
-idx <- lapply(restabs, fun <- function(x1) {
-    idx1 <- lapply(x1, fun <- function(x2) {
-        a <- x2$padj < 0.05
-        b <- x2$log2FoldChange > 0
-        c <- a & b
-        return(c)
-    })
-    return(idx1)
-})
-
-l <- list()
-for (n1 in names(restabs)) {
-    l[[n1]] <- list()
-    for (n2 in names(restabs[[n1]])) {
-        l[[n1]][[n2]] <- restabs[[n1]][[n2]][idx, "gene_id"]
-    }
-}
-
-l <- lapply(restabs, fun <- function(x) {
-    for (i in seq_len(length(x))) {
-        x[[i]]
-    }
-})
-
-tophervs <- restabs[["TUMvHLT"]][idx, "gene_id"]
-
-cohort <- "A"
-restabsA <- load_res_tables("herv", scope="hrv")
-cohort <- "B"
-restabsB <- load_res_tables("herv", scope="hrv")
-cohort <- "C"
-restabsC <- load_res_tables("herv", scope="hrv")
-
-idxA <- lapply(restabsA, fun <- function(x) {
-    a <- x$padj < 0.05
-    b <- x$log2FoldChange > 0
-    c <- a & b
-    return(c)
-})
-
-conds[[1]] <- restabs[["NATvHLT"]]$padj < 0.05
-conds[[2]] <- restabs[["NATvHLT"]]$log2FoldChange > 0
-
 We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.

### 2020-12-27 scripts/res_infla.R

@@ -1,0 +1,114 @@
+## to be sourced in results.R
+
+
+simple_qq <- function(restabs, target) {
+    cat("Making simple QQ\n")
+    y <- list()
+    y[[1]] <- -log10(restabs[[1]]$pvalue)
+    y[[2]] <- -log10(restabs[[2]]$pvalue)
+    y[[3]] <- -log10(restabs[[3]]$pvalue)
+    x <- -log10(qunif(ppoints(length(y[[1]]))))
+
+    xlab <- expression("Expected -log"["10"]*"("*italic("P")*")")
+    ylab <- expression("Observed -log"["10"]*"("*italic("P")*")")
+
+    pdf(file=target, height=3, width=9)
+    par(mfrow=c(1,3))
+    qqplot(x, y[[1]], xlab=xlab, ylab=ylab, main="NATvHLT", col="blue")
+    qqplot(x, y[[3]], xlab=xlab, ylab=ylab, main="TUMvNAT", col="blue")
+    qqplot(x, y[[2]], xlab=xlab, ylab=ylab, main="TUMvHLT", col="blue")
+    dev.off()
+    cat("figure written to", target, "\n")
+}
+
+
+bacon_qq <- function(tab_herv, target=NULL) {
+    cat("Running BACON\n")
+    bc <- lapply(tab_herv, fun <- function(x) {bacon(
+            effectsizes=x$log2FoldChange,
+            standarderrors=x$lfcSE,
+    )})
+    if (!is.null(target)) {
+        cat("Making BACON QQ\n")
+        pdf(file=target, height=3, width=6)
+        print(lapply(bc, fun <- function(x) {
+            plot(x, type="hist")
+        }))
+        print(lapply(bc, fun <- function(x) {
+            plot(x, type="qq")
+        }))
+        dev.off()
+        cat("figure written to", target, "\n")
+    }
+    return(bc)
+}
+
+
+bacon_correction <- function(tab_herv) {
+    tab_herv_bac <- lapply(tab_herv, fun <- function(x) {
+        pbac <- bacon::pval(
+            bacon::bacon(
+                effectsizes=x$log2FoldChange,
+                standarderrors=x$lfcSE,
+            )
+        )
+        x$pbac <- pbac
+        return(x)
+    })
+    return(tab_herv_bac)
+}
+
+
[…]

### 2020-12-27 scripts/res_qq.R

@@ -1,102 +1,0 @@
-## to be sourced in results.R
-
-
-simple_qq <- function(restabs, target) {
-    y <- list()
-    y[[1]] <- -log10(restabs[[1]]$pvalue)
-    y[[2]] <- -log10(restabs[[2]]$pvalue)
-    y[[3]] <- -log10(restabs[[3]]$pvalue)
-    x <- -log10(qunif(ppoints(length(y[[1]]))))
-
-    xlab <- expression("Expected -log"["10"]*"("*italic("P")*")")
-    ylab <- expression("Observed -log"["10"]*"("*italic("P")*")")
-
-    pdf(file=target, height=3, width=9)
-    par(mfrow=c(1,3))
-    qqplot(x, y[[1]], xlab=xlab, ylab=ylab, main="NATvHLT", col="blue")
-    qqplot(x, y[[3]], xlab=xlab, ylab=ylab, main="TUMvNAT", col="blue")
-    qqplot(x, y[[2]], xlab=xlab, ylab=ylab, main="TUMvHLT", col="blue")
-    dev.off()
-    cat("figure written to", target, "\n")
-}
-
-
-tab_herv <- load_res_tables("herv")
-bc <- lapply(tab_herv, fun <- function(x) {bacon(
-        effectsizes=x$log2FoldChange,
-        standarderrors=x$lfcSE,
-)})
-
-target_dir <- Sys.getenv("plot_dir")
-check_dir(target_dir)
-target <- paste0(target_dir, "herv-qq2_", cohort, ".pdf")
-pdf(file=target, height=3, width=6)
-print(lapply(bc, fun <- function(x) {
-    plot(x, type="hist")
-}))
-print(lapply(bc, fun <- function(x) {
-    plot(x, type="qq")
-}))
-dev.off()
-
-tab_herv2 <- lapply(tab_herv, fun <- function(x) {
-    padj2 <- bacon::pval(
-        bacon::bacon(
-            effectsizes=x$log2FoldChange,
-            standarderrors=x$lfcSE,
-        )
-    )
-    x$padj2 <- padj2
-    return(x)
-})
-
[…]

### 2020-12-27 scripts/res_topherv.R

@@ -1,34 +1,6 @@
 ## to be sourced in results.R
 
 
-parse_band <- function(e, cytomap) {
-    ## used within map_cytoband()
-    cond1 <- cytomap$chrom == e["chrom"]
-    cond2 <- cytomap$start <= as.numeric(e["start"])
-    cond3 <- cytomap$stop >= as.numeric(e["start"])
-    idx <- cond1 & cond2 & cond3
-    return(paste(
-        e["hrvnm_l"],
-        paste0(strsplit(e["chrom"], "chr")[[1]][2], cytomap[idx, "band"])
-    ))
-}
-
-
-map_cytoband <- function(tophervs) {
-    target <- paste0(Sys.getenv("ref_dir"), "annotation/ucsc/",
-        "cytoBand.txt.gz"
-    )
-    con <- gzfile(target, "rb")
-    cytomap <- readr::read_tsv(con,
-        col_names=c("chrom", "start", "stop", "band", "giemsa")
-    )
-    close(con)
-    l <- setNames(base::lapply(strsplit(tophervs, "[|]"), unpack_id), tophervs)
-    tophervs2 <- base::lapply(l, parse_band, cytomap)
-    return(tophervs2)
-}
-
-
 select_tophervs <- function(model="naive") {
     ## model indicates underlying assumption about gene expression in NAT vs HLT
     ## field or naive; field assumes NAT is intermediate between HLT and TUM;

### 2020-12-27 scripts/res_utilities.R

@@ -31,22 +31,27 @@ set_paths <- function(in_path, in_name) {
 }
 
 
-load_res_tables <- function(study, scope="all") {
+load_res_tables <- function(study, scope="all", set_cohort=NULL) {
     ## scope can be "all" or "hrv"
     ## order genes by gene_id instead of rank so that order is consistent across
     ## different results tables to facilitate comparisons across tables
     cat("Loading results tables\n")
+    if (is.null(set_cohort)) {
+        cohort_choice <- cohort
+    } else {
+        cohort_choice <- set_cohort
+    }
     if (study == "herv") {
         in_path <- Sys.getenv("table_dir")
-        if (cohort == "B") {
+        if (cohort_choice == "B") {
             in_name <- c(
                 paste0("res_B_CRCvNAT_", scope, ".tsv")
             )
         } else {
             in_name <- c(
-                paste0("res_", cohort, "_NATvHLT_", scope, ".tsv"),
-                paste0("res_", cohort, "_CRCvHLT_", scope, ".tsv"),
-                paste0("res_", cohort, "_CRCvNAT_", scope, ".tsv")
+                paste0("res_", cohort_choice, "_NATvHLT_", scope, ".tsv"),
+                paste0("res_", cohort_choice, "_CRCvHLT_", scope, ".tsv"),
+                paste0("res_", cohort_choice, "_CRCvNAT_", scope, ".tsv")
             )
         }
     } else {
@@ -97,4 +102,32 @@ unpack_id <- function(e) {
     return(setNames(c(hvnm_s, hvnm_l, chrom, start, stop),
         c("hrvnm_s", "hrvnm_l", "chrom", "start", "stop"))
     )
+}
+
+
+parse_band <- function(e, cytomap) {
+    ## used within map_cytoband()
+    cond1 <- cytomap$chrom == e["chrom"]
+    cond2 <- cytomap$start <= as.numeric(e["start"])
+    cond3 <- cytomap$stop >= as.numeric(e["start"])
+    idx <- cond1 & cond2 & cond3
+    return(paste(
+        e["hrvnm_l"],
+        paste0(strsplit(e["chrom"], "chr")[[1]][2], cytomap[idx, "band"])
+    ))
+}
+
+
+map_cytoband <- function(hervlist) {
+    target <- paste0(Sys.getenv("ref_dir"), "annotation/ucsc/",
+        "cytoBand.txt.gz"
+    )
[…]

### 2020-12-27 scripts/res_val.R

@@ -1,0 +1,196 @@
+## to be sourced in results.R
+
+
+fill_restabs <- function() {
+    cat("Loading all results for validation testing\n")
+    restabs <- list()
+    for (i in c("A", "B", "C")) {
+        restabs[[i]] <- load_res_tables("herv", scope="hrv", set_cohort=i)
+    }
+    return(restabs)
+}
+
+
+calc_cor_val <- function(restabs, cohort1, cohort2, method="pearson") {
+    cat("Checking correlations between HERV cohorts\n")
+    s1 <- paste0("res_", cohort1, "_CRCvNAT_hrv")
+    s2 <- paste0("res_", cohort2, "_CRCvNAT_hrv")
+    intx <- intersect(
+        restabs[[cohort1]][[s1]]$gene_id,
+        restabs[[cohort2]][[s2]]$gene_id
+    )
+    i <- restabs[[cohort1]][[s1]]$gene_id %in% intx
+    lfc1 <- restabs[[cohort1]][[s1]][i, "log2FoldChange"]
+    i <- restabs[[cohort2]][[s2]]$gene_id %in% intx
+    lfc2 <- restabs[[cohort2]][[s2]][i, "log2FoldChange"]
+    cc <- cor(lfc1, lfc2, method=method)
+    return(setNames(list(lfc1, lfc2, cc), c("lfc1", "lfc2", "coef")))
+}
+
+
+plot_cor_val <- function(l, labs) {
+    df <- data.frame(x=l$lfc1, y=l$lfc2)
+    ggp_title <- "Pearson Correlation"
+    ggp_subtitle <- paste("HERVs Tested =", length(l$lfc1))
+    ggp_xlab <- paste("L2FC from", labs[1])
+    ggp_ylab <- paste("L2FC from", labs[2])
+    txt <- paste("R =", round(l$coef, 2))
+    ggp <- ggplot(df, aes(x, y)) +
+        geom_point(size=1.5, stroke=0, shape=16, alpha=0.5) +
+        labs(
+            title=element_blank(),
+            subtitle=ggp_subtitle,
+            x=ggp_xlab,
+            y=ggp_ylab
+        ) +
+        annotate("text", x=(min(df$x) + 1), y=0.95*max(df$y), label=txt,
+            size=2.5) +
+        ggp_theme_default
+    return(ggp)
+}
+
+
+val_corr_analysis <- function(restabs) {
+    cat("Performing correlations tests for validation\n")
[…]

### 2020-12-27 scripts/results.R

@@ -29,11 +29,14 @@ library(ggplot2)
 library(cowplot)
 library(reshape2)
 library(dplyr)
+library(xtable)
 
 
 args <- commandArgs(trailingOnly=TRUE)
 cohort <- args[2]
+validate <- as.numeric(args[3])
 cat(paste("Cohort:", cohort, "\n"))
+cat(paste("Validate:", validate, "\n"))
 
 
 source("res_utilities.R")
@@ -42,6 +45,8 @@ source("res_corr.R")
 source("res_hervsum.R")
 source("res_volcano.R")
 source("res_topherv.R")
+source("res_infla.R")
+source("res_val.R")
 
 
 correlation_analysis <- function() {
@@ -109,16 +114,44 @@ main <- function() {
 }
 
 
-main()
+inflation_tests <- function() {
+    tab_herv <- load_res_tables("herv")
+
+    target_dir <- Sys.getenv("plot_dir")
+    check_dir(target_dir)
+    target <- paste0(target_dir, "herv-qq_", cohort, ".pdf")
+    simple_qq(tab_herv, target)
+
+    target_dir <- Sys.getenv("plot_dir")
+    check_dir(target_dir)
+    target <- paste0(target_dir, "herv-qq2_", cohort, ".pdf")
+    bacon_qq(tab_herv, target=target)
+}
 
 
+validation_tests <- function() {
+    restabs <- fill_restabs()
+
+    pl <- val_corr_analysis(restabs)
+    target_dir <- Sys.getenv("plot_dir")
+    check_dir(target_dir)
+    target <- paste0(target_dir, "herv-val-cor.pdf")
+    write_cor_val(pl, target)
+
+    l <- extract_main_hervs(restabs)
+    valdf <- build_val_df(restabs, l)
+    target_dir <- Sys.getenv("table_dir")
+    check_dir(target_dir)
+    target <- paste0(target_dir, "herv-val-table.tex")
+    write_valtable(valdf, target)
+    l2 <- stringent_val(l)
+    return(l2)
+}
 
 
-# target <- "/scratch/chd5n/test-fig.pdf"
-# ggsave(target, p, device="pdf", width=15, height=4, units="in")
-
-tab_herv <- load_res_tables("herv")
-target_dir <- Sys.getenv("plot_dir")
-check_dir(target_dir)
-target <- paste0(target_dir, "herv-qq_", cohort, ".pdf")
-simple_qq(tab_herv, target)
[…]

### 2020-12-27 scripts/results.sh

@@ -12,9 +12,9 @@
 
 ## -- requires config.sh to set environmental variables
 ## -- for use in snakemake workflow
-## -- usage: snakemake setup_dge
+## -- usage: snakemake results
 
-## -- perform: 1.5m, 2.5GB for cohort A
+## -- perform: 1-2m, 3.4GB for cohort A
 
 
 script_name=results.sh
@@ -22,6 +22,7 @@ script_name=results.sh
 
 ## command-line arguments
 cohort=$1
+validate=$2
 
 
 startup_report () {
@@ -30,6 +31,7 @@ startup_report () {
   echo Starting $script_name ...
   echo
   echo Cohort: $cohort
+  echo Validation: $validate
   echo
 }
 
@@ -56,7 +58,7 @@ shutdown_report () {
 
 run_module () {
   echo Running R script ...
-  Rscript ${script_name%.sh}.R --args $cohort
+  Rscript ${script_name%.sh}.R --args $cohort $validate
 }
 
 

### 2020-12-27 scripts/dge.R
update dge to save adj counts

@@ -49,7 +49,7 @@ load_inputs <- function(target1, target2, target3) {
 create_dds <- function(l) {
     ## create DESeq2 data set with txi and dataframe
     ## pre-filter genes prior to normalization
-    ## get size-factors and normalized counts
+    ## get normalizationFactors for normalized counts
     cat("Creating pre-filtered dds\n")
     if (cohort != "B") {
         dds <- DESeqDataSetFromTximport(l$txi, l$colData, ~ phenotype)
@@ -60,7 +60,7 @@ create_dds <- function(l) {
     sample_lim <- 1/2 * ncol(dds)
     keep <- rowSums( counts(dds) > count_lim ) >= sample_lim
     dds <- dds[keep, ]
-    dds <- DESeq(dds)
+    dds <- estimateSizeFactors(dds)
     return(dds)
 }
 
@@ -221,6 +221,14 @@ check_dir <- function(dir) {
 }
 
 
+write_adj <- function(adj, target) {
+    ## save adjusted counts
+    cat("Writing adjusted counts to file\n")
+    save(adj, file=target)
+    cat("adj written to", target, "\n")
+}
+
+
 write_dge <- function(dgeobj, target) {
     cat("Writing dgeobj to file\n")
     save(dgeobj, file=target)
@@ -270,6 +278,9 @@ main <- function() {
         dds <- run_sva(svaobj, dds)
     }
     adj <- count_adjustment(dds)
+    target <- paste0(Sys.getenv("rdata_dir"), "adj_", cohort, ".Rda")
+    check_dir(Sys.getenv("rdata_dir"))
+    write_adj(adj, target)
     dds <- filter_genes(dds, adj, inputs$hervs)
     dgeobj <- test_dge(dds, adj)
     target <- paste0(Sys.getenv("rdata_dir"), "dge_", cohort, ".Rda")

### 2020-12-27 scripts/dge.sh

@@ -12,7 +12,7 @@
 
 ## -- requires config.sh to set environmental variables
 ## -- for use in snakemake workflow
-## -- usage: snakemake setup_dge
+## -- usage: snakemake dge
 
 ## -- perform: 0.5h, 10GB for cohort A
 

### 2020-12-27 scripts/dge.R
fix way dge handles adj counts

@@ -148,8 +148,10 @@ filter_genes <- function(dds, adj, hervs) {
     test_set <- base::unique(targets)
     idx <- rownames(dds) %in% test_set
     dds <- dds[idx, ]
+    idx <- rownames(adj) %in% test_set
+    adj <- adj[idx, ]
     cat("Returning", nrow(dds), "genes in test set\n")
-    return(dds)
+    return(setNames(list(dds, adj), c("dds", "adj")))
 }
 
 
@@ -221,11 +223,11 @@ check_dir <- function(dir) {
 }
 
 
-write_adj <- function(adj, target) {
-    ## save adjusted counts
-    cat("Writing adjusted counts to file\n")
-    save(adj, file=target)
-    cat("adj written to", target, "\n")
+write_dds <- function(ddsobj, target) {
+    ## save prelim dds and adjusted counts
+    cat("Writing ddsobj to file\n")
+    save(ddsobj, file=target)
+    cat("ddsobj written to", target, "\n")
 }
 
 
@@ -278,11 +280,12 @@ main <- function() {
         dds <- run_sva(svaobj, dds)
     }
     adj <- count_adjustment(dds)
-    target <- paste0(Sys.getenv("rdata_dir"), "adj_", cohort, ".Rda")
+    ddsobj <- setNames(list(dds, adj), c("dds", "adj"))
+    target <- paste0(Sys.getenv("rdata_dir"), "dds_", cohort, ".Rda")
     check_dir(Sys.getenv("rdata_dir"))
-    write_adj(adj, target)
-    dds <- filter_genes(dds, adj, inputs$hervs)
-    dgeobj <- test_dge(dds, adj)
+    write_dds(ddsobj, target)
+    ddsobj <- filter_genes(ddsobj$dds, ddsobj$adj, inputs$hervs)
+    dgeobj <- test_dge(ddsobj$dds, ddsobj$adj)
     target <- paste0(Sys.getenv("rdata_dir"), "dge_", cohort, ".Rda")
     check_dir(Sys.getenv("rdata_dir"))
     write_dge(dgeobj, target)

### 2020-12-27 scripts/dge.sh

@@ -14,7 +14,7 @@
 ## -- for use in snakemake workflow
 ## -- usage: snakemake dge
 
-## -- perform: 0.5h, 10GB for cohort A
+## -- perform: <0.5h, <10GB for cohort A
 
 
 script_name=dge.sh

### 2020-12-28 manuscript/main.tex
compile main and supp figs and main table

@@
 We aimed to discover expression of novel, cancer-specific HERV sequences, so we selected a subset of transcripts composed of exons without GENCODE annotations, at least one of which was classifed as a HERV element.
-There were 16,336 transcripts composed of 35,666 exons from 5,025 genes that met our criteria (Figure \ref{fig:ele_sum}A).
+There were 16,336 transcripts composed of 35,666 exons from 5,025 genes that met our criteria (Figure \ref{fig:pre_sum}A).
 To estimate gene abundance in colorectal transciptomes, RNA-seq reads from a discovery cohort of 834 colorectal samples (\textbf{Supplementary Table 1}) were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
 A total of 150 unannotated HERV genes were expressed at levels within a standard deviation of canonical protein coding genes.
-Approximately half were from classes K and H, two of the better characterized classes of HERV (Figure \ref{fig:ele_sum}B).
+Approximately half were from classes K and H, two of the better characterized classes of HERV (Figure \ref{fig:pre_sum}B).
 
@@
 Without \emph{a priori} knowledge of whether HERV genes are likely to be affected by the transcriptomic field effect, we compared tumor samples to healthy samples and tumor adjacent samples separately.
-Out of 150 HERV genes tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (Figure \ref{fig:main_find}A,B, red points).
-We intersected these two lists to identify 11 HERV genes over-expressed in tumors relative to both healthy and tumor adjacent samples (Figure \ref{fig:main_find}C).
+Out of 150 HERV genes tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (Figure \ref{fig:dge_res}A,B, red points).
+We intersected these two lists to identify 11 HERV genes over-expressed in tumors relative to both healthy and tumor adjacent samples (Figure \ref{fig:dge_res}C).
 These 11 HERV genes could serve as tumor biomarkers and may also encode proteins unknown to the adaptive immune system that could serve as targets for immune therapy with lower likelihood of off target effects.
@@
 
-validation #1
-validation #2
+%validation #1
+%validation #2
 
@@
 
+\input{tabs/tab1.tex}
+
 \section*{Figures}
@@
   {
-    \includegraphics[width=0.9\textwidth]{herv-element-sum_A}
+    \includegraphics[width=0.9\textwidth]{fig1}
   }
   \caption{Summary of unannotated human endogenous retroviral sequences}
-  \label{fig:ele_sum}
+  \label{fig:pre_sum}
+\end{figure}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=0.9\textwidth]{fig2}
+  }
+  \caption{Differential expression results for unannotated HERV genes}
+  \label{fig:dge_res}
 \end{figure}

### 2020-12-28 scripts/res_figs.R

@@ -1,0 +1,218 @@
+## to be sourced in results.R
+
+
+write_cor_select <- function(plotlist, target) {
+    p_width <- 6
+    p_height <- 3
+    fig <- cowplot::plot_grid(
+        plotlist[["TUMvHLT"]], plotlist[["TUMvNAT"]],
+        nrow=1, ncol=2,
+        labels=c("A", "B")
+    )
+    pdf(file=target, width=p_width, height=p_height)
+    print(fig)
+    dev.off()
+    cat("figure written to", target, "\n")
+    return(fig)
+}
+
+
+fig_supp_cor_field <- function() {
+    cat("Making supp fig HERV vs field effect corr\n")
+    tab_herv <- load_res_tables("herv")
+    tab_field <- load_res_tables("field")
+    l <- calc_cor(tab_herv, tab_field)
+    target_dir <- Sys.getenv("plot_dir")
+    check_dir(target_dir)
+    target <- paste0(target_dir, "herv-field-cor-select_A.pdf")
+    fig <- write_cor_select(l$fig, target)
+    return(fig)
+}
+
+
+load_inputs_prelim_expr <- function(target1, target2, target3) {
+    cat("Loading inputs for prelim expr\n")
+    df <- data.frame(ensembldb::genes(EnsDb.Hsapiens.v86))
+    idx <- df$gene_biotype == "protein_coding"
+    pcg <- as.vector(df[idx, "gene_id"])
+    hv <- data.frame(readr::read_tsv(target1))
+    v <- as.vector(hv$herv_id)
+    load(target2, verbose=TRUE)
+    load(target3, verbose=TRUE)
+    return(setNames(
+        list(pcg, v, ddsobj, dgeobj),
+        c("pcg", "hervs", "ddsobj", "dgeobj")
+    ))
+}
+
+
+plot_prelim_expr <- function(pcg, hervs, ddsobj, pheno=NULL) {
+    cat("Plotting prelim expr\n")
+    ggp_title <- "Protein Coding and HERV Genes"
+    ggp_xlab <- "Expression Level (All)"
+    ggp_ylab <- "Number of Genes"
+    idx <- rownames(ddsobj$adj) %in% pcg
+    adjpcg <- ddsobj$adj[idx, ]
+    idx <- rownames(ddsobj$adj) %in% hervs
+    adjhrv <- ddsobj$adj[idx, ]
+    if (!is.null(pheno)) {
[…]

### 2020-12-28 scripts/res_val.R

@@ -137,7 +137,7 @@ build_val_df <- function(restabs, l) {
         return(vt)
     })
     valdf <- rbind(valtabs$C$res_C_CRCvHLT_hrv, valtabs$C$res_C_CRCvNAT_hrv)
-    fields <- c("gene_id", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj")
+    fields <- c("gene_id", "baseMean", "log2FoldChange", "lfcSE", "stat", "pvalue", "padj")
     valdf <- valdf[!duplicated(valdf$gene_id), fields]
     valdf$fc <- round(2 ^ valdf$log2FoldChange)
     valdf$CTL <- unlist(
@@ -173,13 +173,15 @@ build_val_df <- function(restabs, l) {
         )
     )
     valdf <- valdf[valdf$order, ]
-    fields <- c("HERV ID", "log2(FC)", "SE", "Stat", "P", "P^", "FC", "CTL", "HERV", "o")
+    fields <- c("HERV ID", "Mean", "Log2(FC)", "SE", "Stat", "P", "P Adj", "FC", "CTL", "HERV", "o")
     colnames(valdf) <- fields
-    fields <- c("HERV", "HERV ID", "FC", "log2(FC)", "SE", "Stat", "P", "P^", "CTL")
+    fields <- c("HERV", "Mean", "FC", "Log2(FC)", "SE", "Stat", "P", "P Adj", "CTL")
     valdf <- valdf[ , fields]
-    for (i in c("log2(FC)", "SE", "Stat", "P", "P^")) {
+    for (i in c("Mean", "Log2(FC)", "SE", "Stat", "P", "P Adj")) {
         valdf[ , i] <- signif(valdf[ , i], 2)
     }
+    valdf$HERV <- gsub("HERV", "", valdf$HERV)
+    valdf$HERV <- gsub("_", " ", valdf$HERV)
     return(valdf)
 }
 
@@ -189,8 +191,10 @@ write_valtable <- function(valdf, target) {
     obj <- xtable::xtable(
         valdf,
         caption="Table 1",
-        label="tab_val"
+        label="tab_val",
+        digits=c(0, 0, 0, 0, 2, 2, 2, 2, 2, 0),
+        display=c("d", "s", "f", "f", "f", "f", "f", "E", "E", "s")
     )
-    xtable::print.xtable(obj, file=target)
+    xtable::print.xtable(obj, file=target, include.rownames=FALSE)
     cat("table written to", target, "\n")
 }

### 2020-12-28 scripts/results.R

@@ -30,6 +30,7 @@ library(cowplot)
 library(reshape2)
 library(dplyr)
 library(xtable)
+library(EnsDb.Hsapiens.v86)
 
 
 args <- commandArgs(trailingOnly=TRUE)
@@ -154,4 +155,14 @@ if (validate == 1) {
     print(l2)
 } else {
     main()
+}
+
+
+### DEVS
+
+make_figures <- function() {
+    fig <- list()
+    fig[[1]] <- fig_supp_cor_field()
+    fig[[2]] <- fig_prelim_expr()
+    fig[[3]] <- fig_dge_result()
 }

### 2020-12-28 scripts/xelatex.sh

@@ -1,0 +1,21 @@
+## this script is experimental
+
+fig1=$1
+fig2=$2
+tab1=$3
+
+cp $fig1 ../manuscript/figs/fig1.pdf
+cp $fig2 ../manuscript/figs/fig2.pdf
+cp $tab1 ../manuscript/tabs/tab1.tex
+
+printf '%s' "Adding texlive to \$PATH"
+export PATH=$PATH:/home/chd5n/apps/texlive/2020/bin/x86_64-linux
+
+cd ../manuscript/
+rm main.{aux,bbl,blg,log,pdf}
+rm ._main.pdf
+
+xelatex main
+bibtex main
+xelatex main
+xelatex main

### 2020-12-28 manuscript/main.tex
make demographics tables and fix manuscript formatting

@@
     \usepackage{graphicx}                     %% for figures
+    \usepackage{caption}                      %% for \captionsetup command
+    \usepackage{placeins}                     %% for \FloatBarrier command
 
@@
     \graphicspath{{figs/}}                    %% set path to figures relative to this document
+    \captionsetup{singlelinecheck=false, justification=justified}
 
@@
 There were 16,336 transcripts composed of 35,666 exons from 5,025 genes that met our criteria (Figure \ref{fig:pre_sum}A).
-To estimate gene abundance in colorectal transciptomes, RNA-seq reads from a discovery cohort of 834 colorectal samples (\textbf{Supplementary Table 1}) were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
+To estimate gene abundance in colorectal transciptomes, RNA-seq reads from a discovery cohort of 834 colorectal samples were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
+Phenotypic and demographic characteristics of the discovery cohort are presented in Table \ref{tab:discovery}.
 A total of 150 unannotated HERV genes were expressed at levels within a standard deviation of canonical protein coding genes.
@@
 First, we verified the consistency of relative protein coding gene expression between tissue phenotypes in the present study and a prior study of the same samples.
-For the comparisons between tumor samples and healthy and tumor adjacent samples, the Pearson correlations coefficients were 0.95 and 0.97, respectively (\textbf{Supplementary Figure 1}).
+For the comparisons between tumor samples and healthy and tumor adjacent samples, the Pearson correlations coefficients were 0.95 and 0.97, respectively (Supplementary Figure \ref{fig:corrs}A,B).
 The high concordance of effect sizes between this study and the prior, which employed similar methods to estimate gene abundance but did not consider HERV transcripts, indicated expansion of the reference transcriptome to include unannotated HERV sequences did not compromise estimates of relative abundance.
@@
 
+Supplementary Table \ref{tab:ind1}
+Supplementary Table \ref{tab:ind2}
 %validation #1
@@
 
+\newpage
 \section*{Acknowledgements}
@@
 
+\newpage
 \bibliography{main}
@@
 
+\newpage
 \section*{Figure Legends}
@@
 
+\newpage
 \section*{Tables}
+\input{tabs/tab1.tex}
+\input{tabs/tab2.tex}
 
-Tabs
-
-\input{tabs/tab1.tex}
-
+\FloatBarrier
+\newpage
 \section*{Figures}
@@
 
+\FloatBarrier
+\newpage
+\section*{Supplementary Tables}
+\input{tabs/spt1.tex}
+\input{tabs/spt2.tex}
+
+\FloatBarrier
+\newpage
+\section*{Supplementary Figures}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=0.9\textwidth]{spf1}
+  }
+  \caption{Correlations of effect sizes across studies and cohorts}
+  \label{fig:corrs}
+\end{figure}
+
 \end{document}

### 2020-12-28 manuscript/tabs/spt1.tex
make demographics tables and fix manuscript formatting

\begin{table}[ht]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l|l|}
    \hline
        Independent Cohort \#1 &  &  &  &  &  &  &  &  &  &  &  \\ \hline
         &  &  &  &  &  &  &  &  &  &  &  \\ \hline
        Source & Dataset name & \# samples &  &  &  & Sex (\% female) &  &  & Age (mean, sd) &  &  \\ \hline
         &  & HLT & NAT & TUM & Total & HLT & NAT* & TUM* & HLT & NAT* & TUM* \\ \hline
        GDC & TCGA-COAD & -- & -- & 135 & 135 & -- & -- & 46.7\% & -- & -- & 69.1, 12.0 \\ \hline
        GDC & TCGA-READ & -- & -- & 60 & 60 & -- & -- & 41.7\% & -- & -- & 67.7, 10.0 \\ \hline
        SRA & MtSinai & -- & 47 & -- & 47 & -- & 61.7\% & -- & -- & 65.4, 13.3 & -- \\ \hline
        SRA & CityOfHope & -- & 2 & 8 & 10 & -- & NR & NR & -- & NR & NR \\ \hline
        SRA & Singapore & -- & 5 & 2 & 7 & -- & NR & NR & -- & NR & NR \\ \hline
        SRA & Utah & 10 & -- & -- & 10 & 40.0\% & -- & -- & 54.8, 7.3 & -- & -- \\ \hline
        SRA & ZhejiangU & -- & 2 & 2 & 4 & -- & 0.0\% & 0.0\% & -- & 58.5, 30.4 & 53.5, 2.1 \\ \hline
        SRA & TexasAM & 2 & -- & -- & 2 & 50.0\% & -- & -- & 38.5, 7.8 & -- & -- \\ \hline
        Total &  & 12 & 56 & 207 & 275 & 41.7\% & 59.2\% & 44.7\% & 52.1, 9.5 & 65.1, 13.8 & 68.5, 11.4 \\ \hline
        * missing data on subset of samples; NR signifies not reported &  &  &  &  &  &  &  &  &  &  &  \\ \hline
    \end{tabular}
    \caption{Supplementary Table 1}
    \label{tab:ind1}
\end{table}

### 2020-12-28 manuscript/tabs/tab_disc.tex
make demographics tables and fix manuscript formatting

\begin{table}[ht]
    \small
    \centering
    \caption{Phenotypic and demographic characteristics of discovery cohort}
    \label{tab:discovery}
    \begin{tabular}{ll|llll|lll|lll}
        \textbf{Source} & \textbf{Dataset} & \multicolumn{4}{l|}{\textbf{\# samples}} & \multicolumn{3}{l|}{\textbf{Sex (\% female)}} & \multicolumn{3}{l}{\textbf{Age (mean, sd)}} \\
        \cline{3-12}
         & & HLT & NAT & TUM & Total & HLT & NAT* & TUM* & HLT & NAT* & TUM* \\
        \hline
        BarcUVa & BarcUVa & 260 & -- & -- & 260 & 63 & -- & -- & 60, 7 & -- & -- \\
        GDC & TCGA-COAD & -- & 36 & 209 & 245 & -- & 53 & 45 & -- & 72, 13 & 65, 13 \\
        GDC & TCGA-READ & -- & 8 & 81 & 89 & -- & 88 & 46 & -- & 67, 18 & 63, 12 \\
        SRA & GTEx & 202 & -- & -- & 202 & 42 & -- & -- & 49, 13 & -- & -- \\
        SRA & HebeiMU & -- & 4 & 5 & 9 & -- & 75 & 60 & -- & \emph{NR} & \emph{NR} \\
        SRA & KoreaAMC & -- & 8 & 6 & 14 & -- & \emph{NR} & \emph{NR} & -- & \emph{NR} & \emph{NR} \\
        SRA & KoreaPNU & -- & 1 & 2 & 3 & -- & 0 & 100 & -- & 62, -- & 73, 4 \\
        SRA & Mayo & -- & 4 & 8 & 12 & -- & 50 & 38 & -- & 65, 16 & 66, 18 \\
        \hline
        Total &  & 462 & 61 & 311 & 834 & 54\% & 59\% & 46\% & 55, 11 & 70, 14 & 65, 13 \\
        \hline
        \multicolumn{12}{l}{* missing data on subset of samples; \emph{NR} signifies not reported}
    \end{tabular}
\end{table}

### 2020-12-28 manuscript/tabs/tab_ind2.tex
make demographics tables and fix manuscript formatting

\begin{table}[ht]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l|l|}
    \hline
        Independent Cohort \#2 &  &  &  &  &  &  &  &  &  &  &  \\ \hline
         &  &  &  &  &  &  &  &  &  &  &  \\ \hline
        Source & Dataset name & \# samples &  &  &  & Sex (\% female) &  &  & Age (mean, sd) &  &  \\ \hline
         &  & HLT & NAT & TUM & Total & HLT & NAT* & TUM* & HLT & NAT* & TUM* \\ \hline
        GDC & TCGA-COAD & -- & 3 & 3 & 6 & -- & 50.0\% & 50.0\% & -- & 62.2, 13.4 & 62.2, 13.4 \\ \hline
        GDC & TCGA-READ & -- & 1 & 1 & 2 & -- & 0.0\% & 0.0\% & -- & 50.0, -- & 50.0, -- \\ \hline
        SRA & HebeiMU & -- & 1 & 1 & 2 & -- & 0.0\% & 0.0\% & -- & NR & NR \\ \hline
        SRA & KoreaAMC & -- & 4 & 4 & 8 & -- & NR & NR & -- & NR & NR \\ \hline
        SRA & KoreaPNU & -- & 2 & 2 & 4 & -- & 50.0\% & 50.0\% & -- & 74.0, 1.2 & 74.0, 1.2 \\ \hline
        SRA & Mayo & -- & 4 & 4 & 8 & -- & 50.0\% & 50.0\% & -- & 70.3, 9.7 & 70.3, 9.7 \\ \hline
        Total &  & -- & 15 & 15 & 30 & -- & 36.4\% & 36.4\% & -- & 66.6, 11.6 & 66.6, 11.6 \\ \hline
        * missing data on subset of samples; NR signifies not reported &  &  &  &  &  &  &  &  &  &  &  \\ \hline
    \end{tabular}
    \caption{Supplementary Table 2}
    \label{tab:ind2}
\end{table}

### 2020-12-28 scripts/res_figs.R

@@ -1,13 +1,33 @@
 ## to be sourced in results.R
 
 
-write_cor_select <- function(plotlist, target) {
-    p_width <- 6
-    p_height <- 3
-    fig <- cowplot::plot_grid(
+plot_cor_select <- function(plotlist) {
+    cwp <- cowplot::plot_grid(
         plotlist[["TUMvHLT"]], plotlist[["TUMvNAT"]],
         nrow=1, ncol=2,
         labels=c("A", "B")
+    )
+    return(cwp)
+}
+
+
+plot_cor_val_select <- function(plotlist) {
+    cwp <- cowplot::plot_grid(
+        plotlist[[1]], plotlist[[2]],
+        nrow=1, ncol=2,
+        labels=c("C", "D")
+    )
+    return(cwp)
+}
+
+
+write_cor_select <- function(plotlist, target) {
+    p_width <- 6
+    p_height <- 6
+    fig <- cowplot::plot_grid(
+        plotlist=plotlist,
+        nrow=2, ncol=1,
+        labels=c(NULL, NULL)
     )
     pdf(file=target, width=p_width, height=p_height)
     print(fig)
@@ -17,15 +37,25 @@ write_cor_select <- function(plotlist, target) {
 }
 
 
-fig_supp_cor_field <- function() {
-    cat("Making supp fig HERV vs field effect corr\n")
+fig_supp_cor_field_val <- function() {
+    cat("Making supp fig HERV corrs\n")
+    cwpl <- list()
+
     tab_herv <- load_res_tables("herv")
     tab_field <- load_res_tables("field")
     l <- calc_cor(tab_herv, tab_field)
+    l$fig[["TUMvHLT"]] <- l$fig[["TUMvHLT"]] + labs(subtitle="TUM - HLT")
+    l$fig[["TUMvNAT"]] <- l$fig[["TUMvNAT"]] + labs(subtitle="TUM - NAT")
+    cwpl[[1]] <- plot_cor_select(l$fig)
+
+    restabs <- fill_restabs()
+    pl <- val_corr_analysis(restabs)
+    cwpl[[2]] <- plot_cor_val_select(pl)
+
     target_dir <- Sys.getenv("plot_dir")
     check_dir(target_dir)
-    target <- paste0(target_dir, "herv-field-cor-select_A.pdf")
-    fig <- write_cor_select(l$fig, target)
+    target <- paste0(target_dir, "herv-cor-select.pdf")
+    fig <- write_cor_select(cwpl, target)
     return(fig)
 }
 

### 2020-12-28 scripts/res_val.R

@@ -190,11 +190,14 @@ write_valtable <- function(valdf, target) {
     cat("Writing validated results to tex table\n")
     obj <- xtable::xtable(
         valdf,
-        caption="Table 1",
+        caption="Tumor-specific HERV genes validated in independent cohort",
         label="tab_val",
         digits=c(0, 0, 0, 0, 2, 2, 2, 2, 2, 0),
         display=c("d", "s", "f", "f", "f", "f", "f", "E", "E", "s")
     )
-    xtable::print.xtable(obj, file=target, include.rownames=FALSE)
+    xtable::print.xtable(obj, file=target, size="small", include.rownames=FALSE,
+        hline.after=c(0,nrow(obj)), caption.placement="top",
+        sanitize.colnames.function=function(x){paste('{\\textbf{',x,'}}', sep='')}
+    )
     cat("table written to", target, "\n")
 }

### 2020-12-28 scripts/results.R

@@ -48,6 +48,7 @@ source("res_volcano.R")
 source("res_topherv.R")
 source("res_infla.R")
 source("res_val.R")
+source("res_figs.R")
 
 
 correlation_analysis <- function() {
@@ -162,7 +163,9 @@ if (validate == 1) {
 
 make_figures <- function() {
     fig <- list()
-    fig[[1]] <- fig_supp_cor_field()
+    fig[[1]] <- fig_supp_cor_field_val()
     fig[[2]] <- fig_prelim_expr()
     fig[[3]] <- fig_dge_result()
 }
+
+make_figures()

### 2020-12-28 scripts/xelatex.sh

@@ -3,12 +3,20 @@
 fig1=$1
 fig2=$2
 tab1=$3
+tab2=$4
+spf1=$5
+spt1=$6
+spt2=$7
 
 cp $fig1 ../manuscript/figs/fig1.pdf
 cp $fig2 ../manuscript/figs/fig2.pdf
 cp $tab1 ../manuscript/tabs/tab1.tex
+cp $tab2 ../manuscript/tabs/tab2.tex
+cp $spf1 ../manuscript/figs/spf1.pdf
+cp $spt1 ../manuscript/tabs/spt1.tex
+cp $spt2 ../manuscript/tabs/spt2.tex
 
-printf '%s' "Adding texlive to \$PATH"
+printf '%s' "Adding texlive to \$PATH\n"
 export PATH=$PATH:/home/chd5n/apps/texlive/2020/bin/x86_64-linux
 
 cd ../manuscript/

### 2020-12-29 scripts/res_figs.R
final figure, heatmaps

@@ -217,6 +217,7 @@ write_dge_result <- function(plotlist, target) {
     print(fig)
     dev.off()
     cat("figure written to", target, "\n")
+    return(fig)
 }
 
 
@@ -245,4 +246,201 @@ fig_dge_result <- function() {
     target <- paste0(target_dir, "herv-dge-result.pdf")
     fig <- write_dge_result(cwpl, target)
     return(fig)
+}
+
+
+get_herv_cat <- function(hervids) {
+    maincats <- c("HERVE", "HERVH", "HERVK", "HERVL")
+    maincls <- c("E", "H", "K", "L")
+    nms <- unname(sapply(lapply(strsplit(hervids, "[|]"), unpack_id), head, 1))
+    a <- sapply(nms, fun <- function(x) {
+        if (x %in% maincats) {
+            y <- plyr::mapvalues(x, from=maincats, to=maincls, warn_missing=FALSE)
+        } else {
+            y <- "Other"
+        }
+        return(y)
+    }, USE.NAMES=FALSE)
+    return(a)
+}
+
+
+get_heat_data <- function(target) {
+    cat("Getting data ready for heatmap\n")
+    load(target, verbose=TRUE)
+
+    hrvadj <- dgeobj$adj[grepl("HERV", rownames(dgeobj$adj)), ]
+    hrvmn <- apply(hrvadj, 1, mean)
+    hrvsd <- apply(hrvadj, 1, sd)
+    hrvz <- data.frame((hrvadj - hrvmn) / hrvsd)
+    colnames(hrvz) <- colData(dgeobj$dds)$subject_id
+    hrvz <- hrvz[ , order(colData(dgeobj$dds)$phenotype)]
+    hrvz$herv_id <- rownames(hrvz)
+    hrvz$herv_cl <- get_herv_cat(hrvz$herv_id)
+    hrvz <- hrvz[order(hrvz$herv_cl, decreasing=TRUE), ]
+    hrvz$n <- 0
+    for (i in levels(as.factor(hrvz$herv_cl))) {
+        hrvz[hrvz$herv_cl == i, "n"] <- seq_len(nrow(hrvz[hrvz$herv_cl == i, ]))
+    }
+    hrvz$y <- paste(hrvz$herv_cl, hrvz$n, sep="_")
+    hrvz <- hrvz[ , !(colnames(hrvz) == "n")]
+
+    df <- melt(
+        hrvz,
+        id.vars=c("herv_id", "herv_cl", "y"),
[…]

### 2020-12-29 scripts/res_themes.R

@@ -13,8 +13,8 @@ ggp_theme_default <- theme(
     axis.title.y=element_text(size=10, face="plain"),
     axis.text.x=element_text(size=9, face="plain"),
     axis.text.y=element_text(size=9, face="plain"),
-    # axis.ticks.x=element_none(),
-    # axis.ticks.y=element_none(),
+    # axis.ticks.x=element_blank(),
+    # axis.ticks.y=element_blank(),
     axis.line.x.bottom=element_line(),
     axis.line.y.left=element_line(),
     legend.key=element_rect(fill="white"),
@@ -62,6 +62,25 @@ ggp_theme_box <- ggp_theme_default +
         legend.spacing.x=unit(0.5, "cm"),
         legend.spacing.y=unit(0.5, "cm"),
         legend.text=element_text(size=9, margin=margin(t=0.1))
+    )
+
+
+ggp_theme_heat <- ggp_theme_default +
+    theme(
+        # axis.title.x=element_text(size=10, face="plain"),
+        # axis.title.y=element_text(size=10, face="plain"),
+        axis.text.x=element_text(size=9, face="plain"),
+        axis.text.y=element_text(size=9, face="plain"),
+        axis.title.x=element_blank(),
+        axis.title.y=element_blank(),
+        # axis.text.x=element_blank(),
+        # axis.text.y=element_blank(),
+        axis.ticks.x=element_blank(),
+        axis.ticks.y=element_blank(),
+        axis.line.x.bottom=element_blank(),
+        axis.line.y.left=element_blank(),
+        legend.position="top",
+        legend.title=element_text(size=9, face="plain")
     )
 
 

### 2020-12-29 scripts/results.R

@@ -166,6 +166,7 @@ make_figures <- function() {
     fig[[1]] <- fig_supp_cor_field_val()
     fig[[2]] <- fig_prelim_expr()
     fig[[3]] <- fig_dge_result()
+    fig[[4]] <- fig_heat()
 }
 
 make_figures()

### 2020-12-29 scripts/xelatex.sh

@@ -2,17 +2,21 @@
 
 fig1=$1
 fig2=$2
-tab1=$3
-tab2=$4
-spf1=$5
-spt1=$6
-spt2=$7
+fig3=$3
+tab1=$4
+tab2=$5
+spf1=$6
+spf2=$7
+spt1=$8
+spt2=$9
 
 cp $fig1 ../manuscript/figs/fig1.pdf
 cp $fig2 ../manuscript/figs/fig2.pdf
+cp $fig3 ../manuscript/figs/fig3.pdf
 cp $tab1 ../manuscript/tabs/tab1.tex
 cp $tab2 ../manuscript/tabs/tab2.tex
 cp $spf1 ../manuscript/figs/spf1.pdf
+cp $spf2 ../manuscript/figs/spf2.pdf
 cp $spt1 ../manuscript/tabs/spt1.tex
 cp $spt2 ../manuscript/tabs/spt2.tex
 

### 2020-12-29 manuscript/main.tex
final figure, heatmaps
break out manuscript sections and supplement

--- 2020-12-28 manuscript/main.tex
@@
 \section*{Synopsis}
-1 to 3 sentences, no more than 45 words
+\input{secs/synopsis.tex}
 
@@
 \section*{Abstract}
-\textbf{Background}:
-Microsatellite unstable, mismatch repair deficient metastatic cancers can be treated with immune checkpoint inhibitors, but new therapeutic approaches are needed to extend the benefits of immune modulators to the majority of patients with microsatellite stable tumors.
-Cancer specific over expression of human endogenous retroviral elements could be exploited to enhance immune infiltration of microsatellite stable colorectal cancer and render these tumors more susceptible to immune checkpoint inhibition.
-To identify novel targets for immune mediated cancer therapies, we tested for cancer specific over expression of endogenous retroviral transcripts in a large cohort of colorectal transcriptomes.
-
-\textbf{Methods}:
-A recently curated RNA sequencing dataset composed of 1,139 human primary colorectal tissue samples divided into three cohorts was re-processed using a novel annotation of repeat element transcripts to quantify levels of repeat element expression in colorectal transcriptomes.
-A total of NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples were profiled with short read sequencing.
-Transcripts were quantified with Salmon using the GENCODE v36 transcriptome augmented to include unannotated exons from long terminal repeat elements classified as human endogenous retroviruses (HERVs) in the Dfam database.
-Associations between transcript expression and tissue phenotype were tested using DESeq2 in the R statistical language.
-
-\textbf{Results}:
-Out of 5,025 unannotated HERV genes measured in 834 samples, including NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
-Over half of the 150 HERV genes with moderate expression were in the H (26) or K (51) classes.
-Tumor specific over expression was observed in 11 of the 150 HERV genes tested.
-
-\textbf{Conclusions}: The conclusions go here.
+\input{secs/abstract.tex}
 
@@
 \section*{Introduction}
-Colorectal cancer (CRC) is a common malignancy in the United States.
-Early detection and effective treatment result in a 5-year relative survival of 90\% for patients diagnosed with local disease.
-However, over a fifth of new cases are diagnosed with metastatic disease, which has a 5-year relative survival of only 14\% \citep{SEER2020}.
-To improve overall survival of patients with distant disease, new therapeutic approaches are needed.
-
-Immunotherapy in CRC is a promising new treatment option for patients with distant disease.
-Immune checkpoint inhibition was shown to achieve durable responses with programmed cell death protein 1 (PD-1) antibody monotherapy and combination therapy with antibodies for PD-1 and cytotoxic T-lymphocyte-associated
-protein 4 (CTLA-4) \citep{Le2015, Overman2018}.
-However, response to immune checkpoint inhibition was limited to the subset of tumors with mismatch repair defficiency and high microsatellite instability, which constitute approximately 4\% of all metastatic CRC cases \citep{Ganesh2019}.
-Strategies to improve tumor immunogenicity and increase cytotoxic T cell infiltration are required to help extend the benefit of immune therapies to the other 96\% of metastatic CRC.
-Epitopes derived from endogenous retroviral peptides expressed in cancers may offer a solution.
-
-Human endogenous retroviruses (HERVs) are proviral elements integrated into the human genome under conditions that render them incompetent of replication or infection.
-Residual HERV elements in contemporary human genomes are the consequence of retroviral infections of germ cells millions of years ago.
-Since then, and like the rest of the human genome, HERV elements have accumulated point mutations and segmental deletions.
-These alterations in HERV sequences disrupt assembly of functional viral machinery.
-Nevertheless, some HERV sequences can still be transcribed and translated, and a subset are processed as endogenous antigens and presented on major histocompatibility complex (MHC) class I molecules \citep{Boller1997, Rooney2015}.
-In healthy tissues, DNA methylation and histone modification can silence transcription of HERV genes, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and lead to translation of neoantigens \citep{Menendez2004, Wiesner2015}.
-
-The cancer specificity of putative HERV peptides makes them promising biomarkers for and targets of immune therapies.
-A major challenge to the development of HERV based therapies is the absence of consensus on how to find HERV gene products.
-Proviruses are generally absent from human gene, transcript, and protein annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
-Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
-Furthermore, HERVs are transposable elements that have generated many similar proviral integrations throughout the human genome.
-The multitude of similar sequences across the genome impairs accurate high throughout read alignment.
-As a result, the genomic locations and transcriptional abundances of most HERVs are not known with confidence.
-
-Recently, new approaches have been developed to overcome traditional limitations in HERV annotation, quantification, and protein identification \citep{Attig2017, Attig2019, ERVmap2018, Telescope2019}.
-Here, we extend advances in the characterization of HERV transcriptomes to discover novel CRC specific expression patterns.
-We test for differential expression of HERV transcripts in a large cohort of healthy, tumor-adjacent normal, and tumor RNA sequencing data and identify HERV transcripts over expressed in tumor samples.
-We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.
+\input{secs/introduction.tex}
 
 \section*{Methods}
-\subsection*{Sample Selection}
-Bulk RNA sequencing (RNA-seq) from a recently curated set of 1,139 flash frozen colorectal tissue samples was re-processed and analyzed for the presence of unannotated HERV genes \citep{Dampier2020}.
-For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 834 colon biopsies was used.
-Of those biopsies, 462 were healthy tissue, 61 were tumor adjacent normal tissue, and 311 were tumors.
-These samples are referred to as the discovery cohort.
-
-For validation of discovered associations, two independent cohorts were used.
-The first (independent cohort \#1) included single end sequencing from 275 colon biopsies of which NUMBER were healthy, NUMBER were tumor adjacent, and NUMBER were tumor.
-The second (independent cohort \#2) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
-An important characteristic of the discovery cohort and independent cohort \#1 was the independence of all tumor and tumor adjacent samples.
-To permit valid inference from regular linear models, only one tissue phenotype was used per subject.
-
-Samples were gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a recently generated dataset of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
-A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al. \citep{Dampier2020}.
-RNA-seq reads for all samples were stored and analyzed in FASTQ format.
-For samples downloaded in BAM format, Biobambam2 v2.0.87 \citep{Tischler2014} was used to convert reads to FASTQ format.
-
-\subsection*{Human Endogenous Retrovirus Gene Annotation}
-HERVs were identified using the genomic coordinates (GRCh38) and feature names of exons and spliced transcripts of HERV elements from a recently published human cancer transcriptome annotated with the GENCODE v24 basic gene model \citep{Frankish2018} and a model of genome-wide repeat elements \citep{Attig2019}.
-The cancer transcriptome was assembled from RNA-seq reads of 24 subjects from each of 32 cancer types obtained from The Cancer Genome Atlas (TCGA), as previously described \citep{Attig2019}.
-Names and coordinates of repeat elements were generated using RepeatMasker \citep{Smit2015} configured with nhmmer \citep{Wheeler2013} in sensitive mode using the Dfam 2.0 library (v150923) \citep{Hubley2015}, as previously described \citep{Attig2017}.
-The method relies on profile hidden Markov models representing known human repeat families to classify repeat elements throughout the genome.
-
-Previously unannotated HERV genes were selected for the current study in multiple steps.
-A complete gene transfer format (GTF) \citep{GTF} annotation of the repeat element inclusive cancer transcriptome was downloaded from Attig et al. \citep{Attig2019}.
-The annotation was parsed with a custom Python script to produce a new annotation in BED12 \citep{BED12} format.
-Transcripts were selected based on three criteria: i) overlap of at least one exon with a long terminal repeat element, ii) classification of at least one exon as HERV derived, and iii) abscence of any exon overlapping an annotated GENCODE gene.
-All transcripts labeled with the same repeat element identifier (i.e. repeat element class along with genomic coordinates) were considered isoforms of the same HERV gene.
-Genomic coordinates were extracted for all constituent exons of selected transcripts to permit inference of splice sites and exclusion of intronic sequences.
-For display purposes, repeat element identifier genomic coordinates were converted to cytobands.
-
-\subsection*{Transcript Quantification}
-Transcript quantification was performed with Salmon v1.2.1 \citep{Patro2017} in mapping-based mode with the \verb|--gcBias| and \verb|--validateMappings| flags set.
-Inputs to the Salmon \verb|quant| command were RNA-seq reads in FASTQ format and the Salmon index.
-To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34  transcript sequences \citep{Frankish2018} and the list of unannotated HERV transcript sequences.
-GENCODE sequences were downloaded from the GENCODE FTP site \citep{GENCODE-transcripts}.
-HERV sequences were extracted from the GRCh38 reference genome \citep{ENCODE-GRCh38} using the BEDTools \citep{Quinlan2010} \verb|getfasta| command with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts.
-Following the recommendation of Salmon's authors \citep{SalmonDecoys}, contig sequences from the GRCh38 reference genome were subsequently added to the target transcript sequences as decoy sequences to generate a decoy-aware transcriptome for indexing.
-
-\subsection*{Gene Expression Analysis}
-Associations between HERV gene expression and tissue phenotype were tested using the \emph{Bioconductor} suite of genomics analysis packages \citep{bioc} in the R statistical programming language \citep{R}.
-Specifically, \emph{tximport} \citep{Soneson2015} was used to extract transcript length and abundance estimates from Salmon and summarize them to the gene level for input into \emph{DESeq2} \citep{Love2014}.
-Count matrices of HERV and GENCODE genes were prefiltered to select genes with a count of at least one read in at least half of the samples in each cohort.
-Factors to normalize counts for sequencing depth and transcript length were calculated with \emph{DESeq2}.
-Latent factors of non-biological variation were estimated with \emph{sva} \citep{sva}.
-Genes of interest for association testing were selected in several steps.
-
-First, counts of all genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \citep{Ritchie2015}.
-Second, adjusted counts of all annotated genes were filtered to retain only protein coding genes.
-Third, median and standard deviation of protein coding gene expression were calculated separately for each tissue phenotype in each cohort.
-Finally, adjusted counts of HERV genes were filtered to retain only HERVs expressed within a single standard deviation of median protein coding gene expression in at least one tissue phenotype.
-
-Counts of all protein coding genes and HERVs with sufficient expression were modeled with phenotype and latent factors as explanatory variables using \emph{DESeq2}.
-Significant associations between gene count and tissue phenotype were identified by testing whether the absolute value of the effect of phenotype was at least two fold change for a given gene.
-Multiple hypothesis testing correction was performed using a false discovery rate of 5\%.
+\input{secs/methods.tex}
 
 \section*{Results}
-\subsection*{Over-Expression of Novel HERV Genes in Tumors}
-To define an inclusive set of non-canonical HERV elements against which to compare CRC transcripts, we started with a recently published model of the human cancer transcriptome assembled \emph{de novo} from TCGA samples and annotated with \emph{RepeatMasker} using the Dfam database \citep{Attig2019}.
-The complete annotation included over 1 million transcripts, 416,862 of which overlapped long terminal repeat sequences (i.e. the repeat sequences that flank HERV genes).
-We aimed to discover expression of novel, cancer-specific HERV sequences, so we selected a subset of transcripts composed of exons without GENCODE annotations, at least one of which was classifed as a HERV element.
-There were 16,336 transcripts composed of 35,666 exons from 5,025 genes that met our criteria (Figure \ref{fig:pre_sum}A).
-To estimate gene abundance in colorectal transciptomes, RNA-seq reads from a discovery cohort of 834 colorectal samples (\textbf{Supplementary Table 1}) were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
-A total of 150 unannotated HERV genes were expressed at levels within a standard deviation of canonical protein coding genes.
-Approximately half were from classes K and H, two of the better characterized classes of HERV (Figure \ref{fig:pre_sum}B).
-
-To identify HERV genes that could encode tumor specific neoantigens, we compared expression levels in tumor samples to levels in healthy samples and tumor adjacent, histologically normal samples.
-We have previously shown that a transcriptomic field effect in the setting of CRC can mask cancer specificity of some genes when tumor transcripts are compared to tumor adjacent normal transcripts \citep{Dampier2020}.
-Without \emph{a priori} knowledge of whether HERV genes are likely to be affected by the transcriptomic field effect, we compared tumor samples to healthy samples and tumor adjacent samples separately.
-Out of 150 HERV genes tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (Figure \ref{fig:dge_res}A,B, red points).
-We intersected these two lists to identify 11 HERV genes over-expressed in tumors relative to both healthy and tumor adjacent samples (Figure \ref{fig:dge_res}C).
-These 11 HERV genes could serve as tumor biomarkers and may also encode proteins unknown to the adaptive immune system that could serve as targets for immune therapy with lower likelihood of off target effects.
-
-To confirm the significance of the 11 HERV genes, we sought multiple validations of our methods and results.
-First, we verified the consistency of relative protein coding gene expression between tissue phenotypes in the present study and a prior study of the same samples.
-For the comparisons between tumor samples and healthy and tumor adjacent samples, the Pearson correlations coefficients were 0.95 and 0.97, respectively (\textbf{Supplementary Figure 1}).
-The high concordance of effect sizes between this study and the prior, which employed similar methods to estimate gene abundance but did not consider HERV transcripts, indicated expansion of the reference transcriptome to include unannotated HERV sequences did not compromise estimates of relative abundance.
-
-Next, we
-
-%validation #1
-%validation #2
-
-We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.
-
-then cohorts B and C
-then cohort A without Attig TCGA 48 + 18 (24 COAD tumor, 24 READ tumor, 12 COAD NAT, 6 READ NAT) and GTEx 12
-then MSI vs MSS
-then tumors with most HERV overexpression
-then immunogenic peptide prediction
-
-\subsection*{HERV Genes in Healthy Tissue}
-
-Healthy overexpression ... functional role
-cohort A
-then cohorts B and C
-then Grabski method
+\input{secs/results.tex}
 
 \section*{Discussion}
+\input{secs/discussion.tex}
 
-HERVs could be used as immune targets or biomarkers
-HERVs may play a role in maintaining healthy epithelium
-Limitations
-Conclusion
-
+\newpage
 \section*{Acknowledgements}
@@
 
+\newpage
 \section*{Figure Legends}
+\input{secs/legends.tex}
 
-Legs
+\newpage
+\section*{Tables}
+\input{tabs/tab1.tex}
+\input{tabs/tab2.tex}
 
-\section*{Tables}
-
-Tabs
-
-\input{tabs/tab1.tex}
-
+\FloatBarrier
+\newpage
 \section*{Figures}
@@
 
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=0.9\textwidth]{fig3}
+  }
+  \caption{Global estimates of HERV expression across tissue phenotypes}
+  \label{fig:heat}
+\end{figure}
+
 \end{document}

### 2020-12-29 manuscript/secs/abstract.tex
break out manuscript sections and supplement

\textbf{Background}:
Microsatellite unstable, mismatch repair deficient metastatic cancers can be treated with immune checkpoint inhibitors, but new therapeutic approaches are needed to extend the benefits of immune modulators to the majority of patients with microsatellite stable tumors.
Cancer specific over expression of human endogenous retroviral elements could be exploited to enhance immune infiltration of microsatellite stable colorectal cancer and render these tumors more susceptible to immune checkpoint inhibition.
To identify novel targets for immune mediated cancer therapies, we tested for cancer specific over expression of endogenous retroviral transcripts in a large cohort of colorectal transcriptomes.

\textbf{Methods}:
A recently curated RNA sequencing dataset composed of 1,139 human primary colorectal tissue samples divided into three cohorts was re-processed using a novel annotation of repeat element transcripts to quantify levels of repeat element expression in colorectal transcriptomes.
A total of NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples were profiled with short read sequencing.
Transcripts were quantified with Salmon using the GENCODE v36 transcriptome augmented to include unannotated exons from long terminal repeat elements classified as human endogenous retroviruses (HERVs) in the Dfam database.
Associations between transcript expression and tissue phenotype were tested using DESeq2 in the R statistical language.

\textbf{Results}:
Out of 5,025 unannotated HERV genes measured in 834 samples, including NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
Over half of the 150 HERV genes with moderate expression were in the H (26) or K (51) classes.
Tumor specific over expression was observed in 11 of the 150 HERV genes tested.

\textbf{Conclusions}: The conclusions go here.

### 2020-12-29 manuscript/secs/discussion.tex
break out manuscript sections and supplement

HERVs could be used as immune targets or biomarkers
HERVs may play a role in maintaining healthy epithelium
Limitations
Conclusion

### 2020-12-29 manuscript/secs/introduction.tex
break out manuscript sections and supplement

Colorectal cancer (CRC) is a common malignancy in the United States.
Early detection and effective treatment result in a 5-year relative survival of 90\% for patients diagnosed with local disease.
However, over a fifth of new cases are diagnosed with metastatic disease, which has a 5-year relative survival of only 14\% \citep{SEER2020}.
To improve overall survival of patients with distant disease, new therapeutic approaches are needed.

Immunotherapy in CRC is a promising new treatment option for patients with distant disease.
Immune checkpoint inhibition was shown to achieve durable responses with programmed cell death protein 1 (PD-1) antibody monotherapy and combination therapy with antibodies for PD-1 and cytotoxic T-lymphocyte-associated
protein 4 (CTLA-4) \citep{Le2015, Overman2018}.
However, response to immune checkpoint inhibition was limited to the subset of tumors with mismatch repair defficiency and high microsatellite instability, which constitute approximately 4\% of all metastatic CRC cases \citep{Ganesh2019}.
Strategies to improve tumor immunogenicity and increase cytotoxic T cell infiltration are required to help extend the benefit of immune therapies to the other 96\% of metastatic CRC.
Epitopes derived from endogenous retroviral peptides expressed in cancers may offer a solution.

Human endogenous retroviruses (HERVs) are proviral elements integrated into the human genome under conditions that render them incompetent of replication or infection.
Residual HERV elements in contemporary human genomes are the consequence of retroviral infections of germ cells millions of years ago.
Since then, and like the rest of the human genome, HERV elements have accumulated point mutations and segmental deletions.
These alterations in HERV sequences disrupt assembly of functional viral machinery.
Nevertheless, some HERV sequences can still be transcribed and translated, and a subset are processed as endogenous antigens and presented on major histocompatibility complex (MHC) class I molecules \citep{Boller1997, Rooney2015}.
In healthy tissues, DNA methylation and histone modification can silence transcription of HERV genes, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and lead to translation of neoantigens \citep{Menendez2004, Wiesner2015}.

The cancer specificity of putative HERV peptides makes them promising biomarkers for and targets of immune therapies.
A major challenge to the development of HERV based therapies is the absence of consensus on how to find HERV gene products.
Proviruses are generally absent from human gene, transcript, and protein annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
Furthermore, HERVs are transposable elements that have generated many similar proviral integrations throughout the human genome.
The multitude of similar sequences across the genome impairs accurate high throughout read alignment.
As a result, the genomic locations and transcriptional abundances of most HERVs are not known with confidence.

Recently, new approaches have been developed to overcome traditional limitations in HERV annotation, quantification, and protein identification \citep{Attig2017, Attig2019, ERVmap2018, Telescope2019}.
Here, we extend advances in the characterization of HERV transcriptomes to discover novel CRC specific expression patterns.
We test for differential expression of HERV transcripts in a large cohort of healthy, tumor-adjacent normal, and tumor RNA sequencing data and identify HERV transcripts over expressed in tumor samples.
We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.

### 2020-12-29 manuscript/secs/legends.tex
break out manuscript sections and supplement

\textbf{Figure 1.} Overview of human endogenous retroviral element discovery in colorectal cancer.
A, Barplot demonstrating
B, Barplot demonstrating
C, Histogram demonstrating
D, Histogram demonstrating

\textbf{Figure 2.} Unannotated proviral elements demonstrate tissue phenotype specificity in colorectal transcriptomes.
A, Scatterplot (i.e. volcano plot) demonstrating
B, Scatterplots (i.e. volcano plot) demonstrating
C, Boxplots demonstrating

\textbf{Figure 3.} Global tendency of healthy colorectal transcriptomes to harbor relatively higher expression of unannotated proviral elements.
A, Heatmap demonstrating
B, Heatmap demonstrating

### 2020-12-29 manuscript/secs/methods.tex
break out manuscript sections and supplement

\subsection*{Sample Selection}
Bulk RNA sequencing (RNA-seq) from a recently curated set of 1,139 flash frozen colorectal tissue samples was re-processed and analyzed for the presence of unannotated HERV genes \citep{Dampier2020}.
For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 834 colon biopsies was used.
Of those biopsies, 462 were healthy tissue, 61 were tumor adjacent normal tissue, and 311 were tumors.
These samples are referred to as the discovery cohort.

For validation of discovered associations, two independent cohorts were used.
The first (independent cohort \#1) included single end sequencing from 275 colon biopsies of which NUMBER were healthy, NUMBER were tumor adjacent, and NUMBER were tumor.
The second (independent cohort \#2) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
An important characteristic of the discovery cohort and independent cohort \#1 was the independence of all tumor and tumor adjacent samples.
To permit valid inference from regular linear models, only one tissue phenotype was used per subject.

Samples were gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a recently generated dataset of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al. \citep{Dampier2020}.
RNA-seq reads for all samples were stored and analyzed in FASTQ format.
For samples downloaded in BAM format, Biobambam2 v2.0.87 \citep{Tischler2014} was used to convert reads to FASTQ format.

\subsection*{Human Endogenous Retrovirus Gene Annotation}
HERVs were identified using the genomic coordinates (GRCh38) and feature names of exons and spliced transcripts of HERV elements from a recently published human cancer transcriptome annotated with the GENCODE v24 basic gene model \citep{Frankish2018} and a model of genome-wide repeat elements \citep{Attig2019}.
The cancer transcriptome was assembled from RNA-seq reads of 24 subjects from each of 32 cancer types obtained from The Cancer Genome Atlas (TCGA), as previously described \citep{Attig2019}.
Names and coordinates of repeat elements were generated using RepeatMasker \citep{Smit2015} configured with nhmmer \citep{Wheeler2013} in sensitive mode using the Dfam 2.0 library (v150923) \citep{Hubley2015}, as previously described \citep{Attig2017}.
The method relies on profile hidden Markov models representing known human repeat families to classify repeat elements throughout the genome.

Previously unannotated HERV genes were selected for the current study in multiple steps.
A complete gene transfer format (GTF) \citep{GTF} annotation of the repeat element inclusive cancer transcriptome was downloaded from Attig et al. \citep{Attig2019}.
The annotation was parsed with a custom Python script to produce a new annotation in BED12 \citep{BED12} format.
Transcripts were selected based on three criteria: i) overlap of at least one exon with a long terminal repeat element, ii) classification of at least one exon as HERV derived, and iii) abscence of any exon overlapping an annotated GENCODE gene.
All transcripts labeled with the same repeat element identifier (i.e. repeat element class along with genomic coordinates) were considered isoforms of the same HERV gene.
Genomic coordinates were extracted for all constituent exons of selected transcripts to permit inference of splice sites and exclusion of intronic sequences.
For display purposes, repeat element identifier genomic coordinates were converted to cytobands.

\subsection*{Transcript Quantification}
Transcript quantification was performed with Salmon v1.2.1 \citep{Patro2017} in mapping-based mode with the \verb|--gcBias| and \verb|--validateMappings| flags set.
Inputs to the Salmon \verb|quant| command were RNA-seq reads in FASTQ format and the Salmon index.
To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34  transcript sequences \citep{Frankish2018} and the list of unannotated HERV transcript sequences.
GENCODE sequences were downloaded from the GENCODE FTP site \citep{GENCODE-transcripts}.
HERV sequences were extracted from the GRCh38 reference genome \citep{ENCODE-GRCh38} using the BEDTools \citep{Quinlan2010} \verb|getfasta| command with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts.
Following the recommendation of Salmon's authors \citep{SalmonDecoys}, contig sequences from the GRCh38 reference genome were subsequently added to the target transcript sequences as decoy sequences to generate a decoy-aware transcriptome for indexing.

\subsection*{Gene Expression Analysis}
Associations between HERV gene expression and tissue phenotype were tested using the \emph{Bioconductor} suite of genomics analysis packages \citep{bioc} in the R statistical programming language \citep{R}.
Specifically, \emph{tximport} \citep{Soneson2015} was used to extract transcript length and abundance estimates from Salmon and summarize them to the gene level for input into \emph{DESeq2} \citep{Love2014}.
Count matrices of HERV and GENCODE genes were prefiltered to select genes with a count of at least one read in at least half of the samples in each cohort.
Factors to normalize counts for sequencing depth and transcript length were calculated with \emph{DESeq2}.
Latent factors of non-biological variation were estimated with \emph{sva} \citep{sva}.
Genes of interest for association testing were selected in several steps.

First, counts of all genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \citep{Ritchie2015}.
Second, adjusted counts of all annotated genes were filtered to retain only protein coding genes.
Third, median and standard deviation of protein coding gene expression were calculated separately for each tissue phenotype in each cohort.
Finally, adjusted counts of HERV genes were filtered to retain only HERVs expressed within a single standard deviation of median protein coding gene expression in at least one tissue phenotype.

Counts of all protein coding genes and HERVs with sufficient expression were modeled with phenotype and latent factors as explanatory variables using \emph{DESeq2}.
Significant associations between gene count and tissue phenotype were identified by testing whether the absolute value of the effect of phenotype was at least two fold change for a given gene.
Multiple hypothesis testing correction was performed using a false discovery rate of 5\%.

### 2020-12-29 manuscript/secs/results.tex
break out manuscript sections and supplement

\subsection*{Over-Expression of Novel HERV Genes in Tumors}
To define an inclusive set of non-canonical HERV elements against which to compare CRC transcripts, we started with a recently published model of the human cancer transcriptome assembled \emph{de novo} from TCGA samples and annotated with RepeatMasker using the Dfam database \citep{Attig2019}.
The complete annotation included over 1 million transcripts, 416,862 of which overlapped long terminal repeat sequences (i.e. the repeat sequences that flank HERV genes).
We aimed to discover expression of novel, cancer-specific HERV sequences, so we selected a subset of transcripts composed of exons without GENCODE annotations, at least one of which was classifed as a HERV element.
There were 16,336 transcripts composed of 35,666 exons from 5,025 genes that met our criteria (Figure \ref{fig:pre_sum}A).
To estimate gene abundance in colorectal transciptomes, RNA-seq reads from a discovery cohort of 834 colorectal samples were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
Phenotypic and demographic characteristics of the discovery cohort are presented in Table \ref{tab:discovery}.
A total of 150 unannotated HERV genes were expressed at levels within a standard deviation of canonical protein coding genes.
Approximately half were from classes K and H, two of the better characterized classes of HERV (Figure \ref{fig:pre_sum}B).

To identify HERV genes that could encode tumor specific neoantigens, we compared expression levels in tumor samples to levels in healthy samples and tumor adjacent, histologically normal samples.
We have previously shown that a transcriptomic field effect in the setting of CRC can mask cancer specificity of some genes when tumor transcripts are compared to tumor adjacent normal transcripts \citep{Dampier2020}.
Without \emph{a priori} knowledge of whether HERV genes are likely to be affected by the transcriptomic field effect, we compared tumor samples to healthy samples and tumor adjacent samples separately.
Out of 150 HERV genes tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (Figure \ref{fig:dge_res}A,B, red points).
We intersected these two lists to identify 11 HERV genes over-expressed in tumors relative to both healthy and tumor adjacent samples (Figure \ref{fig:dge_res}C).
These 11 HERV genes could serve as tumor biomarkers and may also encode proteins unknown to the adaptive immune system that could serve as targets for immune therapy with lower likelihood of off target effects.

To confirm the significance of the 11 HERV genes, we sought multiple validations of our methods and results.
First, we verified the consistency of relative protein coding gene expression between tissue phenotypes in the present study and a prior study of the same samples.
For the comparisons between tumor samples and healthy and tumor adjacent samples, the Pearson correlations coefficients were 0.95 and 0.97, respectively (Supplementary Figure \ref{fig:corrs}A,B).
The high concordance of effect sizes between this study and the prior, which employed similar methods to estimate gene abundance but did not consider HERV transcripts, indicated expansion of the reference transcriptome to include unannotated HERV sequences did not compromise estimates of relative abundance.

Next, we

Supplementary Table \ref{tab:ind1}
Supplementary Table \ref{tab:ind2}

We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.

then cohorts B and C
then cohort A without Attig TCGA 48 + 18 (24 COAD tumor, 24 READ tumor, 12 COAD NAT, 6 READ NAT) and GTEx 12
then MSI vs MSS
then tumors with most HERV overexpression
then immunogenic peptide prediction

\subsection*{HERV Genes in Healthy Tissue}

Healthy overexpression ... functional role
cohort A
then cohorts B and C
then Grabski method

### 2020-12-29 manuscript/supp.tex
break out manuscript sections and supplement

--- 2020-12-29 manuscript/main.tex
@@
 
-    \title{Potential Novel Immunotarget Revealed in Mega Analysis of Colorectal Transcriptomes}
+    \title{Supplement to Potential Novel Immunotarget Revealed in Mega Analysis of Colorectal Transcriptomes}
 
@@
     \author[3]{Victor Moreno, MD, PhD}
-    \author[1,*]{Graham Casey, PhD}
+    \author[1]{Graham Casey, PhD}
 
@@
     \affil[5]{Department of Surgery, Seattle Children's, Seattle, Washington, USA}
-    \affil[*]{Correspondence: Graham Casey, PhD, Center for Public Health Genomics, MSB Room 3238, Department of Public Health Sciences, University of Virginia, P.O.
-Box 800717, Charlottesville, VA 22908-0717, gc8r@virginia.edu}
 
@@
 \newpage
-\section*{Synopsis}
-\input{secs/synopsis.tex}
-
-\newpage
-\section*{Abstract}
-\input{secs/abstract.tex}
-
-\newpage
-\section*{Introduction}
-\input{secs/introduction.tex}
-
-\section*{Methods}
-\input{secs/methods.tex}
-
-\section*{Results}
-\input{secs/results.tex}
-
-\section*{Discussion}
-\input{secs/discussion.tex}
-
-\newpage
-\section*{Acknowledgements}
-
-People we want to thank. Grant support.
+\section*{Supplementary Methods}
+\input{sups/section.tex}
 
@@
 \newpage
-\section*{Figure Legends}
-\input{secs/legends.tex}
+\section*{Supplementary Figure Legends}
+\input{sups/section.tex}
 
 \newpage
-\section*{Tables}
-\input{tabs/tab1.tex}
-\input{tabs/tab2.tex}
+\section*{Supplementary Tables}
+\input{tabs/spt1.tex}
+\input{tabs/spt2.tex}
 
@@
 \newpage
-\section*{Figures}
+\section*{Supplementary Figures}
 
@@
   {
-    \includegraphics[width=0.9\textwidth]{fig1}
+    \includegraphics[width=0.9\textwidth]{spf1}
   }
-  \caption{Summary of unannotated human endogenous retroviral sequences}
-  \label{fig:pre_sum}
+  \caption{Correlations of effect sizes across studies and cohorts}
+  \label{fig:corrs}
 \end{figure}
@@
   {
-    \includegraphics[width=0.9\textwidth]{fig2}
+    \includegraphics[width=0.9\textwidth]{spf2}
   }
-  \caption{Differential expression results for unannotated HERV genes}
-  \label{fig:dge_res}
-\end{figure}
-
-\begin{figure}[ht]
-  \makebox[\textwidth][c]
-  {
-    \includegraphics[width=0.9\textwidth]{fig3}
-  }
-  \caption{Global estimates of HERV expression across tissue phenotypes}
-  \label{fig:heat}
+  \caption{Boxplots re-demonstrating healthy-specific over-expression using a different HERV transcriptome}
+  \label{fig:grabski_method}
 \end{figure}
@@
 \end{document}
-
-problems with AMA style file
-https://tex.stackexchange.com/questions/55020/updated-ama-bst-file
-i modified vancouver.bst to get proper author lists

### 2020-12-29 scripts/xelatex.sh

@@ -26,8 +26,15 @@ export PATH=$PATH:/home/chd5n/apps/texlive/2020/bin/x86_64-linux
 cd ../manuscript/
 rm main.{aux,bbl,blg,log,pdf}
 rm ._main.pdf
+rm supp.{aux,bbl,blg,log,pdf}
+rm ._supp.pdf
 
 xelatex main
 bibtex main
 xelatex main
 xelatex main
+
+xelatex supp
+bibtex supp
+xelatex supp
+xelatex supp

### 2020-12-29 manuscript/secs/introduction.tex
fix intro

@@
 Early detection and effective treatment result in a 5-year relative survival of 90\% for patients diagnosed with local disease.
-However, over a fifth of new cases are diagnosed with metastatic disease, which has a 5-year relative survival of only 14\% \citep{SEER2020}.
-To improve overall survival of patients with distant disease, new therapeutic approaches are needed.
+However, early detection relies on screening strategies in which nearly a third of adults do not participate as recommended \citep{CDC2019}.
+Furthermore, over a fifth of new cases are diagnosed with metastatic disease, which has a 5-year relative survival of only 14\% \citep{SEER2020}.
+To improve detection and treatment of patients with colorectal cancer, new diagnostic and therapeutic approaches are needed.
 
-Immunotherapy in CRC is a promising new treatment option for patients with distant disease.
-Immune checkpoint inhibition was shown to achieve durable responses with programmed cell death protein 1 (PD-1) antibody monotherapy and combination therapy with antibodies for PD-1 and cytotoxic T-lymphocyte-associated
-protein 4 (CTLA-4) \citep{Le2015, Overman2018}.
-However, response to immune checkpoint inhibition was limited to the subset of tumors with mismatch repair defficiency and high microsatellite instability, which constitute approximately 4\% of all metastatic CRC cases \citep{Ganesh2019}.
-Strategies to improve tumor immunogenicity and increase cytotoxic T cell infiltration are required to help extend the benefit of immune therapies to the other 96\% of metastatic CRC.
-Epitopes derived from endogenous retroviral peptides expressed in cancers may offer a solution.
+Screening and treatment strategies have begun to rely on genomic biomarkers for prediction and prognosis, but novel genomic features with better sensitivity and specificity are under active investigation \citep{Imperiale2014, Sveen2020}.
+Among emerging treatment strategies, immune checkpoint inhibition is effective against metastatic tumors with mismatch repair defficiency and high microsatellite instability \citep{Le2015, Overman2018}.
+Unfortunately, only 4\% of metastatic CRCs display high microsatellite instability \citep{Ganesh2019}.
+Identification of additional classes of tumor with potential susceptibilty to immunotherapy and development of treatments to improve tumor immunogenicity are required to help extend the benefit of checkpoint blockade.
+Endogenous retroviral elements expressed in cancers may offer a solution.
 
@@
 These alterations in HERV sequences disrupt assembly of functional viral machinery.
-Nevertheless, some HERV sequences can still be transcribed and translated, and a subset are processed as endogenous antigens and presented on major histocompatibility complex (MHC) class I molecules \citep{Boller1997, Rooney2015}.
-In healthy tissues, DNA methylation and histone modification can silence transcription of HERV genes, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and lead to translation of neoantigens \citep{Menendez2004, Wiesner2015}.
+Nevertheless, some HERV sequences can still be transcribed and translated, with some protein products even processed as endogenous antigens and presented on major histocompatibility complex class I molecules \citep{Boller1997, Rooney2015}.
+In healthy tissues, DNA methylation and histone modification can silence transcription of HERV proviruses, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and occasionally lead to translation of neoantigens \citep{Menendez2004, Wiesner2015}.
 
-The cancer specificity of putative HERV peptides makes them promising biomarkers for and targets of immune therapies.
-A major challenge to the development of HERV based therapies is the absence of consensus on how to find HERV gene products.
-Proviruses are generally absent from human gene, transcript, and protein annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
+The HERV transcriptome is a promising source of biomarkers for cancer detection and treatment because of the putative cancer specificity of HERV mRNA expression.
+A major challenge to the development of HERVs as cancer biomarkers is the absence of consensus on how to define and identify HERV elements.
+Proviruses are generally absent from human gene and transcript annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
 Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
@@
 
-Recently, new approaches have been developed to overcome traditional limitations in HERV annotation, quantification, and protein identification \citep{Attig2017, Attig2019, ERVmap2018, Telescope2019}.
-Here, we extend advances in the characterization of HERV transcriptomes to discover novel CRC specific expression patterns.
-We test for differential expression of HERV transcripts in a large cohort of healthy, tumor-adjacent normal, and tumor RNA sequencing data and identify HERV transcripts over expressed in tumor samples.
-We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.
+Recently, new approaches have been developed to overcome traditional limitations in HERV annotation and quantification \citep{Attig2017, Attig2019, ERVmap2018, Telescope2019, Grabski2020}.
+Here, we extend advances in the characterization of HERV transcripts to discover novel CRC specific expression patterns in human colorectal transcriptomes.
+We hypothesized that HERV proviral elements would be preferentially transcribed in tumor samples compared to healthy and tumor adjacent, histologically normal samples.
+We tested this hypothesis in a large cohort of healthy, tumor adjacent, and tumor RNA sequencing data and found several HERV proviruses over-expressed in tumors but many that had greater expression in healthy tissue.

### 2020-12-30 manuscript/secs/results.tex
fix heatmap summary, fix gene vs locus, write results

@@
-\subsection*{Over-Expression of Novel HERV Genes in Tumors}
-To define an inclusive set of non-canonical HERV elements against which to compare CRC transcripts, we started with a recently published model of the human cancer transcriptome assembled \emph{de novo} from TCGA samples and annotated with RepeatMasker using the Dfam database \citep{Attig2019}.
-The complete annotation included over 1 million transcripts, 416,862 of which overlapped long terminal repeat sequences (i.e. the repeat sequences that flank HERV genes).
+\subsection*{Identification of human endogenous retroviral transcripts in colorectal transcriptomes}
+To define an inclusive set of non-canonical, transcribed HERV elements against which to compare human colorectal transcriptomes, we started with a recently published model of the human cancer transcriptome assembled \emph{de novo} from TCGA samples and annotated with RepeatMasker using the Dfam database \citep{Attig2019}.
+The complete annotation included over 1 million transcripts, 416,862 of which overlapped long terminal repeat sequences (i.e. the repeat sequences that flank HERV proviruses).
 We aimed to discover expression of novel, cancer-specific HERV sequences, so we selected a subset of transcripts composed of exons without GENCODE annotations, at least one of which was classifed as a HERV element.
-There were 16,336 transcripts composed of 35,666 exons from 5,025 genes that met our criteria (Figure \ref{fig:pre_sum}A).
-To estimate gene abundance in colorectal transciptomes, RNA-seq reads from a discovery cohort of 834 colorectal samples were quasi-mapped to a transcriptome composed of GENCODE reference transcripts combined with the 16,336 unannotated HERV transcripts using Salmon.
+There were 16,336 transcripts from 5,025 unique HERV loci that met our criteria (Figure \ref{fig:pre_sum}A).
+
+To estimate gene and proviral abundance in colorectal tissue, RNA-seq reads from a discovery cohort of 834 colorectal samples were quasi-mapped to a reference transcriptome composed of GENCODE annotated transcripts combined with the selected 16,336 unannotated HERV transcripts (see Methods).
 Phenotypic and demographic characteristics of the discovery cohort are presented in Table \ref{tab:discovery}.
-A total of 150 unannotated HERV genes were expressed at levels within a standard deviation of canonical protein coding genes.
-Approximately half were from classes K and H, two of the better characterized classes of HERV (Figure \ref{fig:pre_sum}B).
+Healthy tissue donors were younger, on average, than donors of tumor and tumor adjacent specimens.
+Transcript abundance estimates were aggregated into gene abundances for transcripts with GENCODE annotations and into locus abundances for HERV transcripts.
+We found 1,513 unique HERV loci with at least a single read detected in at least half of samples measured (Figure \ref{fig:pre_sum}A).
+These 1,513 proviruses tended to have lower expression compared to levels of mRNA from canonical protein coding genes.
+This trend was observed when all tissue phenotypes were aggregated and when tumor samples were considered separately (Figure \ref{fig:pre_sum}C,D).
 
-To identify HERV genes that could encode tumor specific neoantigens, we compared expression levels in tumor samples to levels in healthy samples and tumor adjacent, histologically normal samples.
-We have previously shown that a transcriptomic field effect in the setting of CRC can mask cancer specificity of some genes when tumor transcripts are compared to tumor adjacent normal transcripts \citep{Dampier2020}.
-Without \emph{a priori} knowledge of whether HERV genes are likely to be affected by the transcriptomic field effect, we compared tumor samples to healthy samples and tumor adjacent samples separately.
-Out of 150 HERV genes tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (Figure \ref{fig:dge_res}A,B, red points).
-We intersected these two lists to identify 11 HERV genes over-expressed in tumors relative to both healthy and tumor adjacent samples (Figure \ref{fig:dge_res}C).
-These 11 HERV genes could serve as tumor biomarkers and may also encode proteins unknown to the adaptive immune system that could serve as targets for immune therapy with lower likelihood of off target effects.
+To identify HERV proviruses more likely to succeed as biomarkers and function like protein coding genes, we filtered proviruses by expression level relative to GENCODE protein coding genes.
+Of the 1,513 unique loci detected, 150 were expressed at levels within one standard deviation of the median of canonical protein coding genes (Figure \ref{fig:pre_sum}A).
+These 150 moderately expressed proviruses were retained for association testing along with GENCODE protein coding genes.
+Approximately half were from HERV class K or H (Figure \ref{fig:pre_sum}B).
 
-To confirm the significance of the 11 HERV genes, we sought multiple validations of our methods and results.
-First, we verified the consistency of relative protein coding gene expression between tissue phenotypes in the present study and a prior study of the same samples.
-For the comparisons between tumor samples and healthy and tumor adjacent samples, the Pearson correlations coefficients were 0.95 and 0.97, respectively (Supplementary Figure \ref{fig:corrs}A,B).
-The high concordance of effect sizes between this study and the prior, which employed similar methods to estimate gene abundance but did not consider HERV transcripts, indicated expansion of the reference transcriptome to include unannotated HERV sequences did not compromise estimates of relative abundance.
+\subsection*{Proviruses associated with colorectal cancer}
+To test our hypothesis that HERV proviral elements would be preferentially transcribed in tumor samples compared to healthy and tumor adjacent samples, we regressed gene and proviral counts on tissue phenotype.
+Confounding factors (e.g.
+age) across tissue phenotypes were modeled with latent factor anaylsis (see Methods), and association tests were adjusted by including terms for the latent factors in regression models.
+We compared tumor samples to healthy samples and to tumor adjacent samples in separate association tests because we had previously found evidence of a transcriptomic field effect in the setting of CRC that can mask cancer specificity of some genes \citep{Dampier2020}.
+Without knowledge of whether HERV transcripts are likely to be affected by the transcriptomic field effect, we performed separate association tests and compared the results.
 
-Next, we
+Out of 150 HERV loci tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (\emph{P} adjusted < 0.05) (Figure \ref{fig:dge_res}A,B, red points).
+The larger number of over-expressed genes in the comparison with healthy tissue could be consistent with field effect bias but could also be due to a larger healthy tissue sample size and concomitant increased statistical power.
+We did not investigate this point further because the intersection of the tumor specific proviruses from both comparisons yielded 11 HERV loci clearly unaffected by field effect bias and expressed in a tumor specific pattern (Figure \ref{fig:dge_res}C).
+These 11 HERVs could serve as novel tumor biomarkers or as templates for translation of proteins to which immune tolerance may not have developed, so we sought to validate their tumor specificity in an independent cohort.
 
-Supplementary Table \ref{tab:ind1}
-Supplementary Table \ref{tab:ind2}
+To verify the consistency of associations between proviral expression and tissue phenotype, we applied the same methods used in the discovery cohort to an independent cohort (cohort \#1) of 275 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 207 tumor samples (Supplementary Table \textbf{1}).
+Of the 150 HERVs moderately expressed (i.e.
+expressed within a standard deviation of median protein coding expression) in the discovery cohort, 92 were also moderately expressed in the independent cohort.
+Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in the independent cohort (Supplementary Figure \textbf{2}C).
+Nominally significant (i.e.
+\emph{P} < 0.05) tumor over expression relative to either healthy or tumor adjacent samples was considered for the purpose of validation due to the small number of healthy samples.
+Seven of the 11 tumor specific HERVs identified in the discovery cohort were validated in the independent cohort (Table \ref{tab:val}).
 
-We then predict the neoantigens most likely to succeed as biomarkers and immunogenic peptides in the treatment of CRC.
+To verify accuracy of bioinformatic and statistical methods, two additional validations were performed.
+The present analysis relied on a modified reference transcriptome and an updated mapping and quantification algorithm relative to a prior study of the same samples \citep{Dampier2020}.
+Consistency of relative protein coding gene expression between phenotypes in the present analysis and the prior analysis was tested.
+Pearson correlation coefficients for comparisons between tumor samples and healthy and tumor adjacent samples were 0.95 and 0.97, respectively (Supplementary Figure \textbf{2}A,B).
+The high concordance of estimates of relative expression between analyses indicated the new bioinformatics techniques employed in the current analysis did not substantially alter estimates of transcript abundance.
 
-then cohorts B and C
-then cohort A without Attig TCGA 48 + 18 (24 COAD tumor, 24 READ tumor, 12 COAD NAT, 6 READ NAT) and GTEx 12
-then MSI vs MSS
-then tumors with most HERV overexpression
-then immunogenic peptide prediction
+Testing in the discovery cohort relied on latent factor analysis to adjust for confounding factors across tissue phenotypes (see Methods).
+An alternative approach is to compare pairs of samples matched by subject.
+This approach eliminates the possibility of comparisons between healthy and affected tissue but is expected to control for differences in age, sex, race, and other possible confounders more effectively.
+Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (Supplementary Table \textbf{2} for cohort characteristics).
+Only 48 HERV loci were tested in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples was highly concordant (Pearson \emph{R} = 0.93)(Supplementary Figure \textbf{2}D).
+The high concordance of estimates indicated the latent factor analysis employed in the discovery cohort adequately controlled the effect of confounders.
 
-\subsection*{HERV Genes in Healthy Tissue}
+%% TO DO
+%% Discovery cohort without Attig TCGA 48 + 18 (24 COAD tumor, 24 READ tumor, 12 COAD NAT, 6 READ NAT) and GTEx 12
+%% MSI vs MSS
+%% tumors with most HERV overexpression vs lymphocyte infiltration (Shirley Liu)
 
-Healthy overexpression ... functional role
-cohort A
-then cohorts B and C
+\subsection*{Proviruses are expressed in healthy colorectal tissue}
+Several HERV proviruses were expressed in tumor specific patterns, in accord with our hypothesis.
+However, a surprising finding of our unbiased survey of unannotated retroviral transcripts in colorectal transcriptomes was the unexpectedly high level of expression of HERV elements in healthy samples.
+The number of differentially expressed genes (\emph{P} adjusted < 0.05) with higher expression in healthy and tumor adjacent samples relative to tumor samples was 94 and 13, respectively.
+For comparison, the number of differentially expressed genes with lower expression in healthy and tumor adjacent samples relative to tumor samples was 28 and 14, respectively.
+This result is demonstrated in Figure \ref{fig:dge_res}A,B (blue points).
+A high number of differentially expressed genes with over expression in healthy tissue was also found in independent cohort \#1 (data not shown).
+
+To demonstrate the frequency with which healthy samples harbored relatively high proviral expression levels and to qualitatively assess whether particular classes of HERVs had different expression patterns, we computed standardized expression levels (i.e.
+Z-scores) for all samples in the discovery cohort and plotted them in two heatmaps (Figure \ref{fig:heat}A,B).
+HERVs are arranged on the y-axis and grouped by HERV class, and samples are arranged on the x-axis and grouped by tissue phenotype.
+Figure \ref{fig:heat}A shows all 150 HERVs and 834 samples tested in the discovery cohort.
+Figure \ref{fig:heat}B shows the same data in a simplified format in which HERVs are averaged by class and samples are averaged by phenotype.
+The darker shade of the healthy panels indicates higher expression levels of most HERVs.
+No particular class of HERVs
+
 then Grabski method

### 2020-12-30 scripts/res_figs.R

@@ -78,15 +78,15 @@ load_inputs_prelim_expr <- function(target1, target2, target3) {
 
 plot_prelim_expr <- function(pcg, hervs, ddsobj, pheno=NULL) {
     cat("Plotting prelim expr\n")
-    ggp_title <- "Protein Coding and HERV Genes"
-    ggp_xlab <- "Expression Level (All)"
-    ggp_ylab <- "Number of Genes"
+    ggp_title <- "GENCODE Genes and HERV Loci"
+    ggp_xlab <- "Expression Level (All Samples)"
+    ggp_ylab <- "Number of Genes/Loci"
     idx <- rownames(ddsobj$adj) %in% pcg
     adjpcg <- ddsobj$adj[idx, ]
     idx <- rownames(ddsobj$adj) %in% hervs
     adjhrv <- ddsobj$adj[idx, ]
     if (!is.null(pheno)) {
-        ggp_xlab <- "Expression Level (TUM)"
+        ggp_xlab <- "Expression Level (Tumor Samples)"
         idx <- colData(ddsobj$dds)$phenotype == pheno
         adjpcg <- adjpcg[ , idx]
         adjhrv <- adjhrv[ , idx]
@@ -135,7 +135,7 @@ run_hrvnums2 <- function(hervs, ddsobj, dgeobj) {
     x3 <- sum(rownames(ddsobj$adj) %in% hervs)
     x4 <- sum(rownames(dgeobj$adj) %in% hervs)
     labs <- c("Total Transcripts", "Selected Transcripts",
-        "Selected HERV Genes", "Detected HERV Genes", "Tested HERV Genes")
+        "Selected HERV Loci", "Detected HERV Loci", "Tested HERV Loci")
     return(setNames(list(x0, x1, x2, x3, x4), labs))
 }
 
@@ -166,7 +166,7 @@ fig_prelim_expr <- function() {
     hrvnums <- run_hrvnums2(l$hervs, l$ddsobj, l$dgeobj)
     hrvnames <- rownames(l$dgeobj$adj)[grepl("HERV", rownames(l$dgeobj$adj))]
     hrvcats <- count_herv_cats(hrvnames)
-    sumlist <- setNames(list(hrvnums, hrvcats), c("HERV Genes", "HERV Classes"))
+    sumlist <- setNames(list(hrvnums, hrvcats), c("HERV Loci", "HERV Classes"))
     pl <- plot_hrvsum(sumlist)
     pl[[1]] <- pl[[1]] +
         labs(title="HERV Elements Studied") +
[…]

### 2020-12-30 scripts/res_hervsum.R

@@ -14,7 +14,7 @@ count_herv_ids <- function(gtf, ids) {
     df <- data.frame(readr::read_tsv(ids))
     x2 <- nrow(df)
     x3 <- length(unique(df$herv_id))
-    nms <- c("Total Transcripts", "Select Transcripts", "HERV Genes")
+    nms <- c("Total Transcripts", "Select Transcripts", "HERV Loci")
     return(setNames(list(x1, x2, x3), nms))
 }
 
@@ -54,7 +54,7 @@ fill_params <- function(l, params) {
         params[["ggp_sub"]] <- paste("From", n, "starting transcripts")
         l <- l[names(l) != "Total Transcripts"]
     } else {
-        params[["ggp_title"]] <- "Classes of HERV Genes Tested"
+        params[["ggp_title"]] <- "Classes of HERV Loci Tested"
         params[["ggp_sub"]] <- element_blank()
     }
     params[["ggp_ylab"]] <- "Counts"

### 2020-12-30 scripts/res_topherv.R

@@ -63,7 +63,7 @@ format_counts <- function(tophervs, dgeobj) {
 plot_counts <- function(dfs) {
     cat("Making box plots\n")
     color_values <- c("blue", "grey", "red")
-    ggp_title <- "Expression Levels of Select HERV Genes"
+    ggp_title <- "Expression Levels of Select HERV Loci"
     ggp_ylab <- "Scaled Counts"
     lidx <- 0
     pl <- list()

### 2020-12-30 scripts/res_val.R

@@ -190,8 +190,8 @@ write_valtable <- function(valdf, target) {
     cat("Writing validated results to tex table\n")
     obj <- xtable::xtable(
         valdf,
-        caption="Tumor-specific HERV genes validated in independent cohort",
-        label="tab_val",
+        caption="Tumor specific HERV proviruses validated in independent cohort",
+        label="tab:val",
         digits=c(0, 0, 0, 0, 2, 2, 2, 2, 2, 0),
         display=c("d", "s", "f", "f", "f", "f", "f", "E", "E", "s")
     )

### 2020-12-30 scripts/results.R

@@ -68,10 +68,10 @@ herv_element_summary <- function() {
     hrvnums <- count_herv_ids(gtf, ids)
     target <- paste0(Sys.getenv("rdata_dir"), "dge_", cohort, ".Rda")
     dgeobj <- load_dgeobj(target)
-    hrvnums[["Expressed HERV Genes"]] <- sum(grepl("HERV", rownames(dgeobj$dds)))
+    hrvnums[["Expressed HERV Loci"]] <- sum(grepl("HERV", rownames(dgeobj$dds)))
     hrvnames <- rownames(dgeobj$dds)[grepl("HERV", rownames(dgeobj$dds))]
     hrvcats <- count_herv_cats(hrvnames)
-    sumlist <- setNames(list(hrvnums, hrvcats), c("HERV Genes", "HERV Classes"))
+    sumlist <- setNames(list(hrvnums, hrvcats), c("HERV Loci", "HERV Classes"))
     pl <- plot_hrvsum(sumlist)
     target_dir <- Sys.getenv("plot_dir")
     check_dir(target_dir)

### 2020-12-30 scripts/res_volcano.R
mute volcano colors with scales muted

@@ -66,9 +66,9 @@ plot_volcano <- function(df, label) {
         # geom_point(data=base::subset(df, herv==TRUE & effect<0),
         #     size=1, stroke=0, shape=16, colour="blue", alpha=0.25) +
         geom_point(data=base::subset(df, herv==TRUE & effect>0 & padj<0.05),
-            size=1.75, stroke=0, shape=16, colour="red", alpha=0.25) +
+            size=1.75, stroke=0, shape=16, colour=muted("red"), alpha=0.25) +
         geom_point(data=base::subset(df, herv==TRUE & effect<0 & padj<0.05),
-            size=1.75, stroke=0, shape=16, colour="blue", alpha=0.25) +
+            size=1.75, stroke=0, shape=16, colour=muted("blue"), alpha=0.25) +
         labs(
             title=element_blank(),
             subtitle=ggp_subtitle,

### 2020-12-30 scripts/results.R

@@ -31,6 +31,7 @@ library(reshape2)
 library(dplyr)
 library(xtable)
 library(EnsDb.Hsapiens.v86)
+library(scales)
 
 
 args <- commandArgs(trailingOnly=TRUE)

### 2020-12-30 manuscript/secs/discussion.tex
fix heatmap summary, fix gene vs locus, write results
finish results, start discussion

@@
-HERVs could be used as immune targets or biomarkers
+The identification of novel biomarkers to diagnose and guide treatment of CRC is a priority of contemporary cancer genomics research \citep{Sveen2019}.
+Biomarkers associated with immunogenic potential are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced disease.
+In this study, we performed the largest transcriptome-wide, unbiased assessment of unannotated human endogenous retroviral elements in colorectal tissue to date.
+We found several unique proviral loci with tumor specific over expression and many other loci with relatively high expression in healthy samples.
+The tumor specific HERV sequences we identified discriminate tumor from healthy tissue and histologically normal tissue adjacent to neoplastic mucosa, which highlights their possible utility in screening and surveillance.
+Furthermore, if the identified sequences are translated, they could encode neoantigens with immunogenic potential that could serve as targets of bispecific antibody therapies in addition to tumor specific biomarkers.
+
+Remarkably, one of the HERV loci with highest expression in cancer was Xp22
+
+This study also demonstrated the frequency of HERV expression in healthy colorectal tissue.
 HERVs may play a role in maintaining healthy epithelium
-Limitations
-Conclusion
+Healthy overexpression ... functional role
+This finding was contrary to our hypothesis and also undermined an alternative explanation of our
+This was an important finding because it was contrary to our hypothesis and because it provided
+First, it was contrary to our hypothesis.
+Second, it provides some
+
+Limitations of this study included a potential for batch effects from different studies and age confounder
+
+These results are important because
+This study is the largest to date to investigate transcriptome-wide effects of CRC on adjacent, histologically normal mucosa using RNA-seq.
+The results are the first to demonstrate the consistent overexpression of 20 tumor-associated genes in histologically normal tissue sampled adjacent to tumors.
+Remarkably, genes contributing to established oncogenic pathways, such as epidermal growth fact

### 2020-12-30 manuscript/secs/results.tex
finish results, start discussion

@@
 expressed within a standard deviation of median protein coding expression) in the discovery cohort, 92 were also moderately expressed in the independent cohort.
-Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in the independent cohort (Supplementary Figure \textbf{2}C).
+Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in the independent cohort (Supplementary Figure \textbf{1}C).
 Nominally significant (i.e.
@@
 Consistency of relative protein coding gene expression between phenotypes in the present analysis and the prior analysis was tested.
-Pearson correlation coefficients for comparisons between tumor samples and healthy and tumor adjacent samples were 0.95 and 0.97, respectively (Supplementary Figure \textbf{2}A,B).
+Pearson correlation coefficients for comparisons between tumor samples and healthy and tumor adjacent samples were 0.95 and 0.97, respectively (Supplementary Figure \textbf{1}A,B).
 The high concordance of estimates of relative expression between analyses indicated the new bioinformatics techniques employed in the current analysis did not substantially alter estimates of transcript abundance.
@@
 Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (Supplementary Table \textbf{2} for cohort characteristics).
-Only 48 HERV loci were tested in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples was highly concordant (Pearson \emph{R} = 0.93)(Supplementary Figure \textbf{2}D).
+Only 48 HERV loci were tested in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples was highly concordant (Pearson \emph{R} = 0.93) (Supplementary Figure \textbf{1}D).
 The high concordance of estimates indicated the latent factor analysis employed in the discovery cohort adequately controlled the effect of confounders.
@@
 Several HERV proviruses were expressed in tumor specific patterns, in accord with our hypothesis.
-However, a surprising finding of our unbiased survey of unannotated retroviral transcripts in colorectal transcriptomes was the unexpectedly high level of expression of HERV elements in healthy samples.
+However, a surprising finding of our survey of unannotated retroviral transcripts in colorectal transcriptomes was the unexpectedly high level of expression of HERV elements in healthy samples.
 The number of differentially expressed genes (\emph{P} adjusted < 0.05) with higher expression in healthy and tumor adjacent samples relative to tumor samples was 94 and 13, respectively.
@@
 
-To demonstrate the frequency with which healthy samples harbored relatively high proviral expression levels and to qualitatively assess whether particular classes of HERVs had different expression patterns, we computed standardized expression levels (i.e.
+To demonstrate the frequency with which healthy samples harbored relatively high proviral expression levels and to qualitatively assess whether particular classes of HERVs had different expression patterns across phenotypes, we computed standardized expression levels (i.e.
 Z-scores) for all samples in the discovery cohort and plotted them in two heatmaps (Figure \ref{fig:heat}A,B).
-HERVs are arranged on the y-axis and grouped by HERV class, and samples are arranged on the x-axis and grouped by tissue phenotype.
-Figure \ref{fig:heat}A shows all 150 HERVs and 834 samples tested in the discovery cohort.
-Figure \ref{fig:heat}B shows the same data in a simplified format in which HERVs are averaged by class and samples are averaged by phenotype.
-The darker shade of the healthy panels indicates higher expression levels of most HERVs.
-No particular class of HERVs
+In the heatmaps, HERVs were listed on the y-axis and grouped by class, and samples were listed on the x-axis and grouped by tissue phenotype.
+Figure \ref{fig:heat}A shows all 150 HERVs and 833 samples tested in the discovery cohort, and Figure \ref{fig:heat}B shows the same data summarized by medians for each class in each phenotype.
+The darker shade of the healthy panels indicated higher expression levels of most HERVs classes.
+Subsets of HERV loci from all classes demonstrated higher expression in tumors, but these were exceptions.
 
-then Grabski method
+To verify this surprising result, we tested for differential expression in the discovery cohort using an alternative HERV transcriptome recently compiled through manual curation of published HERV sequences \citep{Grabski2020}.
+The alternative transcriptome included 91 HERV sequences, all from class K.
+All but one of the six differentially expressed HERV transcripts were expressed at higher levels in healthy samples (Supplementary Figure \textbf{2}).

### 2020-12-30 manuscript/secs/discussion.tex
forging ahead through discussion

@@
 Biomarkers associated with immunogenic potential are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced disease.
-In this study, we performed the largest transcriptome-wide, unbiased assessment of unannotated human endogenous retroviral elements in colorectal tissue to date.
+In this study, we performed the largest unbiased assessment of transcribed but unannotated human endogenous retroviral elements in colorectal tissue to date.
 We found several unique proviral loci with tumor specific over expression and many other loci with relatively high expression in healthy samples.
-The tumor specific HERV sequences we identified discriminate tumor from healthy tissue and histologically normal tissue adjacent to neoplastic mucosa, which highlights their possible utility in screening and surveillance.
-Furthermore, if the identified sequences are translated, they could encode neoantigens with immunogenic potential that could serve as targets of bispecific antibody therapies in addition to tumor specific biomarkers.
+The tumor specific HERV sequences identified can be used to distinguish tumor from healthy and normal tumor adjacent tissue, highlighting their possible utility in screening and surveillance.
+If the identified sequences are translated, they could also encode neoantigens with immunogenic potential.
 
-Remarkably, one of the HERV loci with highest expression in cancer was Xp22
+Three of the seven validated HERV proviruses with tumor specific expression were of class H.
+Over expression of HERV-H proviruses in CRC has been demonstrated previously with microarray \citep{Perot2012} and quantitative real time polymerase chain reaction \citep{Perot2015}.
+The prominence of HERV-H transcripts in RNA-seq from CRC samples of The Cancer Genome Atlas, from which many of the samples in the current study were generated, has also been previously reported \citep{Desai2017, Attig2019}.
+Therefore, our findings are generally consistent with prior work.
+However, we provide new evidence of the reliability of tumor specific over expression by jointly analyzing a large number of samples from different studies.
+Furthermore, our analysis was not restricted to a particular class of provirus.
+The unbiased nature of our assessment allows us to contextualize the utility of measuring transcripts from three specific HERV-H loci relative to other HERV-H loci and loci from other HERV classes.
 
-This study also demonstrated the frequency of HERV expression in healthy colorectal tissue.
-HERVs may play a role in maintaining healthy epithelium
-Healthy overexpression ... functional role
-This finding was contrary to our hypothesis and also undermined an alternative explanation of our
-This was an important finding because it was contrary to our hypothesis and because it provided
-First, it was contrary to our hypothesis.
-Second, it provides some
+The provirus at the Xp22.32 locus appears to be the same HERV-H sequence (Xp22.3) found to be over expressed in multiple gastrointestinal cancers, including CRC, in a previous study \citep{Wentzensen2007}.
+Remarkably, this sequence was subsequently found to encode an \emph{env} protein with at least three immunogenic peptides \citep{Mullins2012}.
+The peptides were successfully used to sensitize human T cells that selectively attacked and killed CRC cell lines expressing the HERV Xp22.3 (Xp22.32) sequence.
+Other HERV loci identified in our analysis may encode peptides with similar immunogenic potential.
+The prediction of protein sequences implied by other tumor specific proviral transcripts and the testing of putative epitopes for immunogenic potential is an important goal of our future work.
+
+This study also demonstrated the surprising frequency of HERV expression in healthy colorectal tissue.
+Interpretation of this result is challenging.
+HERVs have been shown to have rare functional roles in healthy tissue \citep{Rote2004} and could conceivably have a role in maintaining healthy colorectal epithelium, but this is speculative.
+The over expression in healthy tissues could also be a result of the age distribution across tissue phenotypes.
+Latent factor analysis is expected to adjust for multiple confounders, including subject age, but it is possible the adjustment was insufficient.
+Although surprising, the unexpectedly high levels of HERV expression in healthy samples was also reassuring.
+The set of reference sequences used for HERV detection in this analysis was assembled from cancer transcriptomes.
+We expected to have difficulty distinguishing expression due to epigenetic de-repression of repeat sequences from an artifact.
 

### 2020-12-31 manuscript/secs/discussion.tex
fix supp tabs, finish discussion

--- 2020-12-29 manuscript/secs/discussion.tex
@@
-HERVs could be used as immune targets or biomarkers
-HERVs may play a role in maintaining healthy epithelium
-Limitations
-Conclusion
+In this study, we performed the largest unbiased assessment of transcribed but unannotated human endogenous retroviral elements in colorectal tissue to date.
+We found several unique proviral loci with tumor specific over expression and many other loci with relatively high expression in healthy samples.
+Human endogenous retroviral elements are normally transcriptionally repressed repetitive sequences that are often de-repressed in epigenetically altered cancer cells \citep{Babaian2016}.
+HERV transcripts have been shown to engage pattern recognition receptors and activate interferon responses in cancer cell lines \citep{Chiappinelli2015, Roulois2015} and have been associated with tumor infiltration by T regulatory cells \citep{Desai2017}, response to PD-L1 blockade in urothelial carcinoma, and survival in subjects with CRC \citep{Solovyov2018}.
+Our results are important because the tumor specific HERV sequences we identified could be used to distinguish tumor from non-tumor tissue for screening and surveillance and to predict or even modify response to immunotherapies.
+
+Three of the seven HERV proviruses we validated as tumor specific were from group H.
+Over expression of HERV-H proviruses in CRC has been demonstrated previously with microarray \citep{Perot2012} and quantitative real time polymerase chain reaction \citep{Perot2015}.
+The prominence of HERV-H transcripts in RNA-seq from CRC samples of The Cancer Genome Atlas, from which many of the samples in the current study were generated, has also been previously reported \citep{Desai2017, Attig2019}.
+Therefore, our findings are generally consistent with prior work.
+However, we provide new evidence of the reliability of tumor specific over expression by jointly analyzing a large number of samples from different studies.
+Furthermore, our analysis was not restricted to a particular group of proviruses.
+The unbiased nature of our assessment provides important context for the community of researchers selecting genomic features to develop as biomarkers and immunotargets for CRC.
+
+We note that the tumor specific provirus we mapped to the high resolution cytoband Xp22.32 appears to be the same HERV-H provirus previously mapped to the lower resolution cytoband Xp22.3 and found to be over expressed in CRC \citep{Wentzensen2007}.
+Remarkably, this sequence was subsequently found to encode an \emph{env} protein including at least three immunogenic epitopes \citep{Mullins2012}.
+The epitopes were successfully used to sensitize human cytotoxic T cells in culture, and the sensitized lymphocytes were shown to selectively lyse CRC cell lines expressing the HERV Xp22.3 provirus.
+Other HERV loci identified in our analysis may encode peptides with similar immunogenic potential.
+Identification of protein sequences that could be translated from tumor specific proviral transcripts is therefore an important goal of our future work.
+
+This study also demonstrated the surprising frequency of HERV expression in healthy colorectal tissue.
+Interpretation of this result is challenging.
+HERVs have been shown to have rare functional roles in healthy cells \citep{Rote2004} and could conceivably have a role in maintaining healthy colorectal epithelium, but this is speculative.
+Over expression in our healthy samples could also result from technical (i.e. sequencing protocols) or demographic (i.e. age) differences between phenotypes.
+Latent factor analysis is expected to adjust for multiple confounders, including subject age and sequencing artifacts, but the adjustment may be insufficient.
+Although surprising, the detection of high levels of HERV expression in healthy samples reduced the likelihood that using a reference HERV transcriptome assembled exclusively from contigs of cancer cells had biased our results.
+
+The major limitation of this study was its reliance on complex bioinformatic and statistical methods to aggregate heterogeneous RNA-seq datasets and quantify poorly annotated transcripts.
+Specifically, the definition and quantification of HERV elements was of critical importance.
+We used a published, hidden Markov model based annotation of a \emph{de novo} assembled cancer transcriptome to define and locate the HERV transcripts and loci we studied \citep{Attig2019}.
+We used a published, alignment-free, k-mer matching quantification algorithm to probabilistically estimate transcript abundance of repetitive sequences that may otherwise have been excluded from analysis due to multiple possible mappings \citep{Patro2017}.
+We used state of the art statistical packages designed specifically for high dimensional genomics analyses to perform association tests \citep{Love2014, sva}.
+We also pursued multiple validations of our findings and methods to ensure reproducibility.
+
+The aim of this work was to prioritize HERV transcripts for potential development as novel biomarkers or immunotargets in the management of CRC.
+Markers and targets for immune modulation are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced CRC.
+The seven proviruses we identified were consistently and reliably over expression in colorectal tumors and warrant future investigation.

### 2020-12-31 manuscript/tabs/tab_ind1.tex
fix supp tabs, finish discussion

\begin{table}[ht]
    \small
    \centering
    \caption{Phenotypic and demographic characteristics of independent cohort \#1}
    \label{tab:ind1}
    \begin{tabular}{ll|llll|lll|lll}
        \textbf{Source} & \textbf{Dataset} & \multicolumn{4}{l|}{\textbf{\# samples}} & \multicolumn{3}{l|}{\textbf{Sex (\% female)}} & \multicolumn{3}{l}{\textbf{Age (mean, sd)}} \\
        \cline{3-12}
         & & HLT & NAT & TUM & Total & HLT & NAT* & TUM* & HLT & NAT* & TUM* \\
        \hline
        GDC & TCGA-COAD & -- & -- & 135 & 135 & -- & -- & 47 & -- & -- & 69, 12 \\
        GDC & TCGA-READ & -- & -- & 60 & 60 & -- & -- & 42 & -- & -- & 68, 10 \\
        SRA & MtSinai & -- & 47 & -- & 47 & -- & 62 & -- & -- & 65, 13 & -- \\
        SRA & CityOfHope & -- & 2 & 8 & 10 & -- & \emph{NR} & \emph{NR} & -- & \emph{NR} & \emph{NR} \\
        SRA & Singapore & -- & 5 & 2 & 7 & -- & \emph{NR} & \emph{NR} & -- & \emph{NR} & \emph{NR} \\
        SRA & Utah & 10 & -- & -- & 10 & 40 & -- & -- & 55, 7 & -- & -- \\
        SRA & ZhejiangU & -- & 2 & 2 & 4 & -- & 0 & 0 & -- & 59, 30 & 54, 2 \\
        SRA & TexasAM & 2 & -- & -- & 2 & 50 & -- & -- & 39, 8 & -- & -- \\
        \hline
        Total &  & 12 & 56 & 207 & 275 & 42\% & 59\% & 45\% & 52, 10 & 65, 14 & 69, 11 \\
        \hline
        \multicolumn{12}{l}{* missing data on subset of samples; \emph{NR} signifies not reported} \\
    \end{tabular}
\end{table}

### 2020-12-31 manuscript/tabs/tab_ind2.tex
fix supp tabs, finish discussion

@@
 \begin{table}[ht]
+    \small
     \centering
-    \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l|l|}
+    \caption{Phenotypic and demographic characteristics of independent cohort \#2}
+    \label{tab:ind2}
+    \begin{tabular}{ll|llll|lll|lll}
+        \textbf{Source} & \textbf{Dataset} & \multicolumn{4}{l|}{\textbf{\# samples}} & \multicolumn{3}{l|}{\textbf{Sex (\% female)}} & \multicolumn{3}{l}{\textbf{Age (mean, sd)}} \\
+        \cline{3-12}
+         & & HLT & NAT & TUM & Total & HLT & NAT* & TUM* & HLT & NAT* & TUM* \\
         \hline
-        Independent Cohort \#2 &  &  &  &  &  &  &  &  &  &  &  \\ \hline
-         &  &  &  &  &  &  &  &  &  &  &  \\ \hline
-        Source & Dataset name & \# samples &  &  &  & Sex (\% female) &  &  & Age (mean, sd) &  &  \\ \hline
-         &  & HLT & NAT & TUM & Total & HLT & NAT* & TUM* & HLT & NAT* & TUM* \\ \hline
-        GDC & TCGA-COAD & -- & 3 & 3 & 6 & -- & 50.0\% & 50.0\% & -- & 62.2, 13.4 & 62.2, 13.4 \\ \hline
-        GDC & TCGA-READ & -- & 1 & 1 & 2 & -- & 0.0\% & 0.0\% & -- & 50.0, -- & 50.0, -- \\ \hline
-        SRA & HebeiMU & -- & 1 & 1 & 2 & -- & 0.0\% & 0.0\% & -- & NR & NR \\ \hline
-        SRA & KoreaAMC & -- & 4 & 4 & 8 & -- & NR & NR & -- & NR & NR \\ \hline
-        SRA & KoreaPNU & -- & 2 & 2 & 4 & -- & 50.0\% & 50.0\% & -- & 74.0, 1.2 & 74.0, 1.2 \\ \hline
-        SRA & Mayo & -- & 4 & 4 & 8 & -- & 50.0\% & 50.0\% & -- & 70.3, 9.7 & 70.3, 9.7 \\ \hline
-        Total &  & -- & 15 & 15 & 30 & -- & 36.4\% & 36.4\% & -- & 66.6, 11.6 & 66.6, 11.6 \\ \hline
-        * missing data on subset of samples; NR signifies not reported &  &  &  &  &  &  &  &  &  &  &  \\ \hline
+        GDC & TCGA-COAD & -- & 3 & 3 & 6 & -- & 50 & 50 & -- & 62, 13 & 62, 13 \\
+        GDC & TCGA-READ & -- & 1 & 1 & 2 & -- & 0 & 0 & -- & 50, -- & 50, -- \\
+        SRA & HebeiMU & -- & 1 & 1 & 2 & -- & 0 & 0 & -- & \emph{NR} & \emph{NR} \\
+        SRA & KoreaAMC & -- & 4 & 4 & 8 & -- & \emph{NR} & \emph{NR} & -- & \emph{NR} & \emph{NR} \\
+        SRA & KoreaPNU & -- & 2 & 2 & 4 & -- & 50 & 50 & -- & 74, 1 & 74, 1 \\
+        SRA & Mayo & -- & 4 & 4 & 8 & -- & 50 & 50 & -- & 70, 10 & 70, 10 \\
+        \hline
+        Total &  & -- & 15 & 15 & 30 & -- & 36\% & 36\% & -- & 67, 12 & 67, 12 \\
+        \hline
+        \multicolumn{12}{l}{* missing data on subset of samples; \emph{NR} signifies not reported} \\
     \end{tabular}
-    \caption{Supplementary Table 2}
-    \label{tab:ind2}
 \end{table}

### 2020-12-31 manuscript/secs/abstract.tex
fix final results section

@@
 \textbf{Background}:
+Identification of novel genomic features with potential utility as biomarkers or targets for detection and treatment of colorectal cancer is a priority of cancer researchers.
+Markers and targets for immune modulation are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced CRC.
+
 Microsatellite unstable, mismatch repair deficient metastatic cancers can be treated with immune checkpoint inhibitors, but new therapeutic approaches are needed to extend the benefits of immune modulators to the majority of patients with microsatellite stable tumors.

### 2020-12-31 manuscript/secs/acknowledge.tex
fix final results section

The results shown here are in part based on data generated by The Cancer Genome Atlas (TCGA) Research Network: https://www.cancer.gov/tcga.
The results are also based on data generated by the Genotype-Tissue Expression (GTEx) Project.
GTEx was supported by the Common Fund of the Office of the Director of the National Institutes of Health and by NCI, NHGRI, NHLBI, NIDA, NIMH, and NINDS.
The data used for the analyses described in this manuscript were obtained from dbGaP accession number phs000424.v7.p2 on2020-07-15.
This research was supported by the Genomics Core Facility of the CWRU School of Medicine's Genetics and Genome Sciences Department.
Funding for this work was provided by the following NIH Grants: R01 NIH/NCI CA143237 and CA201407 (G.C.), T32 5T32CA163177-07 (Craig Slingluff, MD).
The authors are grateful to Dr Craig Slingluff from the Department of Surgery at the University of Virginia, whose T32 training grant supported CHD and DFG.

### 2020-12-31 manuscript/secs/results.tex
fix final results section

--- 2020-12-30 manuscript/secs/results.tex
@@
 
-To verify the consistency of associations between proviral expression and tissue phenotype, we applied the same methods used in the discovery cohort to an independent cohort (cohort \#1) of 275 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 207 tumor samples (Supplementary Table \textbf{1}).
+To verify the consistency of associations between proviral expression and tissue phenotype, we applied the same methods used in the discovery cohort to an independent cohort (cohort \#1) of 275 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 207 tumor samples (Supplementary Table 1).
 Of the 150 HERVs moderately expressed (i.e.
 expressed within a standard deviation of median protein coding expression) in the discovery cohort, 92 were also moderately expressed in the independent cohort.
-Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in the independent cohort (Supplementary Figure \textbf{2}C).
+Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in the independent cohort (Supplementary Figure 1C).
 Nominally significant (i.e.
@@
 Consistency of relative protein coding gene expression between phenotypes in the present analysis and the prior analysis was tested.
-Pearson correlation coefficients for comparisons between tumor samples and healthy and tumor adjacent samples were 0.95 and 0.97, respectively (Supplementary Figure \textbf{2}A,B).
+Pearson correlation coefficients for comparisons between tumor samples and healthy and tumor adjacent samples were 0.95 and 0.97, respectively (Supplementary Figure 1A,B).
 The high concordance of estimates of relative expression between analyses indicated the new bioinformatics techniques employed in the current analysis did not substantially alter estimates of transcript abundance.
@@
 This approach eliminates the possibility of comparisons between healthy and affected tissue but is expected to control for differences in age, sex, race, and other possible confounders more effectively.
-Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (Supplementary Table \textbf{2} for cohort characteristics).
-Only 48 HERV loci were tested in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples was highly concordant (Pearson \emph{R} = 0.93)(Supplementary Figure \textbf{2}D).
+Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (see Supplementary Table 2 for cohort characteristics).
+Only 48 HERV loci were tested in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples was highly concordant (Pearson \emph{R} = 0.93) (Supplementary Figure 1).
 The high concordance of estimates indicated the latent factor analysis employed in the discovery cohort adequately controlled the effect of confounders.
@@
 
-To demonstrate the frequency with which healthy samples harbored relatively high proviral expression levels and to qualitatively assess whether particular classes of HERVs had different expression patterns, we computed standardized expression levels (i.e.
-Z-scores) for all samples in the discovery cohort and plotted them in two heatmaps (Figure \ref{fig:heat}A,B).
-HERVs are arranged on the y-axis and grouped by HERV class, and samples are arranged on the x-axis and grouped by tissue phenotype.
-Figure \ref{fig:heat}A shows all 150 HERVs and 834 samples tested in the discovery cohort.
-Figure \ref{fig:heat}B shows the same data in a simplified format in which HERVs are averaged by class and samples are averaged by phenotype.
-The darker shade of the healthy panels indicates higher expression levels of most HERVs.
-No particular class of HERVs
-
-then Grabski method
+To demonstrate the frequency with which healthy samples harbored relatively high proviral expression levels and to qualitatively assess whether particular groups of HERVs had different expression patterns across phenotypes, we computed standardized expression levels (i.e.
+Z-scores) for all samples in the discovery cohort and plotted them in a heatmap (Figure \ref{fig:heat}).
+The darker shade of the healthy panels indicated higher expression levels of most HERVs across all groups.
+Subsets of HERV loci from each group demonstrated higher expression in tumors, but these were exceptions.
+To verify this result, we tested for differential expression in the discovery cohort using an alternative HERV transcriptome recently compiled by manual curation of published HERV sequences \citep{Grabski2020}.
+The alternative transcriptome included 91 HERV-K (HML-2) sequences.
+All but one of the six differentially expressed HERV transcripts were higher in healthy samples, indicating our finding was not due to our definition of HERVs (Supplementary Figure 2).

### 2020-12-31 manuscript/main.tex
forging ahead through discussion
fix final results section
finish text, redo figures

@@
 
-    \title{Potential Novel Immunotarget Revealed in Mega Analysis of Colorectal Transcriptomes}
+    %\title{Potential Novel Immunotarget Revealed in Mega Analysis of Colorectal Transcriptomes}
+    \title{Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer}
 
@@
 \section*{Acknowledgements}
-
-People we want to thank. Grant support.
+\input{secs/acknowledge.tex}
 
@@
   }
-  \caption{Summary of unannotated human endogenous retroviral sequences}
+  \caption{Summary of unannotated human endogenous retroviral sequences found in colorectal transcriptomes}
   \label{fig:pre_sum}
@@
   }
-  \caption{Differential expression results for unannotated HERV genes}
+  \caption{Differential expression of unannotated human endogenous retroviral loci}
   \label{fig:dge_res}
@@
   }
-  \caption{Global estimates of HERV expression across tissue phenotypes}
+  \caption{Global estimates of proviral expression across tissue phenotypes}
   \label{fig:heat}

### 2020-12-31 manuscript/secs/abstract.tex
finish text, redo figures

--- 2020-12-29 manuscript/secs/abstract.tex
@@
 \textbf{Background}:
-Microsatellite unstable, mismatch repair deficient metastatic cancers can be treated with immune checkpoint inhibitors, but new therapeutic approaches are needed to extend the benefits of immune modulators to the majority of patients with microsatellite stable tumors.
-Cancer specific over expression of human endogenous retroviral elements could be exploited to enhance immune infiltration of microsatellite stable colorectal cancer and render these tumors more susceptible to immune checkpoint inhibition.
-To identify novel targets for immune mediated cancer therapies, we tested for cancer specific over expression of endogenous retroviral transcripts in a large cohort of colorectal transcriptomes.
+Colorectal cancer is the second leading cause of cancer death in the United States.
+Identification of novel genomic features with potential utility as diagnostic or prognostic biomarkers or therapeutic targets could help improve detection and treatment of patients with colorectal cancer.
+Features associated with immune modulation are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced disease.
+Endogenous retroviral elements expressed in tumors could be exploited for diagnostic or therapeutic use.
+To identify novel markers and targets for immune mediated cancer therapies, we tested for cancer specific over expression of non-canonical endogenous retroviral transcripts in a large cohort of colorectal transcriptomes.
 
 \textbf{Methods}:
-A recently curated RNA sequencing dataset composed of 1,139 human primary colorectal tissue samples divided into three cohorts was re-processed using a novel annotation of repeat element transcripts to quantify levels of repeat element expression in colorectal transcriptomes.
-A total of NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples were profiled with short read sequencing.
-Transcripts were quantified with Salmon using the GENCODE v36 transcriptome augmented to include unannotated exons from long terminal repeat elements classified as human endogenous retroviruses (HERVs) in the Dfam database.
-Associations between transcript expression and tissue phenotype were tested using DESeq2 in the R statistical language.
+A curated RNA sequencing dataset composed of 1,139 human primary colorectal tissue samples divided into three independent cohorts was re-processed using a novel annotation of repeat element transcripts to quantify levels of human endogenous retroviral mRNA in colorectal transcriptomes.
+A total of 474 healthy, 133 tumor adjacent normal, and 532 tumor samples were profiled with short read sequencing.
+Transcripts were quantified with Salmon using the GENCODE v36 transcriptome augmented to include unannotated exons from long terminal repeat elements classified as human endogenous retroviruses in the Dfam database.
+Associations between transcript expression and tissue phenotype were tested using \emph{DESeq2} in the R statistical programming language.
 
 \textbf{Results}:
-Out of 5,025 unannotated HERV genes measured in 834 samples, including NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
-Over half of the 150 HERV genes with moderate expression were in the H (26) or K (51) classes.
-Tumor specific over expression was observed in 11 of the 150 HERV genes tested.
+Out of 5,025 unannotated proviral loci queried in our discovery analysis, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
+Over half of the 150 loci were from two groups of endogenous retrovirus (51 from K, 26 from H).
+Tumor specific over expression was observed in 11 of the 150 loci tested (\emph{P} adjusted < 0.05), and seven were validated in an independent cohort.
+Widespread proviral expression was also found in healthy tissues.
 
-\textbf{Conclusions}: The conclusions go here.
+\textbf{Conclusions}:
+In the largest unbiased assessment of non-canonical human endogenous retroviral transcript expression in colorectal transcriptomes, we found relatively few loci expressed at levels similar to canonical protein coding genes.
+Nevertheless, we identified seven proviruses with tumor specific expression patterns.
+These proviruses have potential as novel biomarkers or immunotargets and should be prioritized in future investigations.

### 2020-12-31 manuscript/secs/discussion.tex
finish text, redo figures

@@
 Human endogenous retroviral elements are normally transcriptionally repressed repetitive sequences that are often de-repressed in epigenetically altered cancer cells \citep{Babaian2016}.
-HERV transcripts have been shown to engage pattern recognition receptors and activate interferon responses in cancer cell lines \citep{Chiappinelli2015, Roulois2015} and have been associated with tumor infiltration by T regulatory cells \citep{Desai2017}, response to PD-L1 blockade in urothelial carcinoma, and survival in subjects with CRC \citep{Solovyov2018}.
+HERV transcripts have been shown to engage pattern recognition receptors and activate interferon responses in cancer cell lines \citep{Chiappinelli2015, Roulois2015} and have been associated with tumor infiltration by T regulatory cells \citep{Desai2017}, response to immune checkpoint blockade in urothelial carcinoma, and survival in subjects with CRC \citep{Solovyov2018}.
 Our results are important because the tumor specific HERV sequences we identified could be used to distinguish tumor from non-tumor tissue for screening and surveillance and to predict or even modify response to immunotherapies.
@@
 Over expression of HERV-H proviruses in CRC has been demonstrated previously with microarray \citep{Perot2012} and quantitative real time polymerase chain reaction \citep{Perot2015}.
-The prominence of HERV-H transcripts in RNA-seq from CRC samples of The Cancer Genome Atlas, from which many of the samples in the current study were generated, has also been previously reported \citep{Desai2017, Attig2019}.
+The prominence of HERV-H transcripts in RNA-seq from TCGA CRC samples has also been previously reported \citep{Desai2017, Attig2019}.
 Therefore, our findings are generally consistent with prior work.
@@
 
-This study also demonstrated the surprising frequency of HERV expression in healthy colorectal tissue.
+This study also demonstrated widespread expression of a subset of HERVs in healthy colorectal tissue.
 Interpretation of this result is challenging.
@@
 
-The major limitation of this study was its reliance on complex bioinformatic and statistical methods to aggregate heterogeneous RNA-seq datasets and quantify poorly annotated transcripts.
-Specifically, the definition and quantification of HERV elements was of critical importance.
+The major limitation of this study was its reliance on heterogeneous RNA-seq datasets and quantifications of short reads of repetitive sequences that are infamously challenging to estimate accurately.
+To address these limitations, we built a complex bioinformatic pipeline based on state of the art software.
+Specifically, our definition of HERV elements was of critical importance.
 We used a published, hidden Markov model based annotation of a \emph{de novo} assembled cancer transcriptome to define and locate the HERV transcripts and loci we studied \citep{Attig2019}.
 We used a published, alignment-free, k-mer matching quantification algorithm to probabilistically estimate transcript abundance of repetitive sequences that may otherwise have been excluded from analysis due to multiple possible mappings \citep{Patro2017}.
-We used state of the art statistical packages designed specifically for high dimensional genomics analyses to perform association tests \citep{Love2014, sva}.
+We used actively developed and updated statistical packages designed specifically for high dimensional genomics analyses to perform association tests \citep{Love2014, sva}.
 We also pursued multiple validations of our findings and methods to ensure reproducibility.
@@
 The aim of this work was to prioritize HERV transcripts for potential development as novel biomarkers or immunotargets in the management of CRC.
-Markers and targets for immune modulation are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced CRC.
-The seven proviruses we identified were consistently and reliably over expression in colorectal tumors and warrant future investigation.
+Markers and targets for immune modulation are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced disease.
+The seven proviruses we identified were consistently and reliably over expressed in colorectal tumors and warrant future investigation.

### 2020-12-31 manuscript/secs/introduction.tex
finish text, redo figures

@@
 
-Screening and treatment strategies have begun to rely on genomic biomarkers for prediction and prognosis, but novel genomic features with better sensitivity and specificity are under active investigation \citep{Imperiale2014, Sveen2020}.
+Screening and treatment strategies have begun to rely on genomic biomarkers for prediction and prognosis, and novel genomic features with better sensitivity and specificity are under active investigation \citep{Imperiale2014, Sveen2020}.
 Among emerging treatment strategies, immune checkpoint inhibition is effective against metastatic tumors with mismatch repair defficiency and high microsatellite instability \citep{Le2015, Overman2018}.
@@
 
-Human endogenous retroviruses (HERVs) are proviral elements integrated into the human genome under conditions that render them incompetent of replication or infection.
+Human endogenous retroviruses (HERVs) are formerly transposable, proviral elements previously integrated into the human genome and now incompetent of transposition, replication, or infection \citep{Babaian2016}.
 Residual HERV elements in contemporary human genomes are the consequence of retroviral infections of germ cells millions of years ago.
-Since then, and like the rest of the human genome, HERV elements have accumulated point mutations and segmental deletions.
+Since then, like the rest of the human genome, HERV elements have accumulated point mutations and segmental deletions.
 These alterations in HERV sequences disrupt assembly of functional viral machinery.
-Nevertheless, some HERV sequences can still be transcribed and translated, with some protein products even processed as endogenous antigens and presented on major histocompatibility complex class I molecules \citep{Boller1997, Rooney2015}.
-In healthy tissues, DNA methylation and histone modification can silence transcription of HERV proviruses, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and occasionally lead to translation of neoantigens \citep{Menendez2004, Wiesner2015}.
+Nevertheless, some HERV sequences can be transcribed and translated, and some protein products can be processed as endogenous antigens and presented on major histocompatibility complex class I molecules \citep{Boller1997, Mullins2012, Rooney2015}.
+In healthy tissues, DNA methylation and histone modification silence HERV transcription, but epigenetic modifications in the context of cancer can de-repress HERV transcription and occasionally lead to translation of neoantigens \citep{Babaian2016, Menendez2004, Wiesner2015}.
 
-The HERV transcriptome is a promising source of biomarkers for cancer detection and treatment because of the putative cancer specificity of HERV mRNA expression.
+The HERV transcriptome is a promising source of novel genomic features for cancer detection and treatment because of the putative cancer specificity of HERV mRNA expression and its association with immune modulation \citep{Chiappinelli2015, Roulois2015, Desai2017, Solovyov2018}.
 A major challenge to the development of HERVs as cancer biomarkers is the absence of consensus on how to define and identify HERV elements.
@@
 Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
-Furthermore, HERVs are transposable elements that have generated many similar proviral integrations throughout the human genome.
+Furthermore, HERVs were originally transposable elements that have generated many similar proviral integrations throughout the human genome.
 The multitude of similar sequences across the genome impairs accurate high throughout read alignment.
@@
 We hypothesized that HERV proviral elements would be preferentially transcribed in tumor samples compared to healthy and tumor adjacent, histologically normal samples.
-We tested this hypothesis in a large cohort of healthy, tumor adjacent, and tumor RNA sequencing data and found several HERV proviruses over-expressed in tumors but many that had greater expression in healthy tissue.
+We tested this hypothesis in a curated dataset of healthy, tumor adjacent, and tumor RNA sequencing (RNA-seq) reads and found several HERV proviruses over expressed in tumors but many that had greater expression in healthy tissue.

### 2020-12-31 manuscript/secs/methods.tex
finish text, redo figures

@@
-\subsection*{Sample Selection}
-Bulk RNA sequencing (RNA-seq) from a recently curated set of 1,139 flash frozen colorectal tissue samples was re-processed and analyzed for the presence of unannotated HERV genes \citep{Dampier2020}.
-For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 834 colon biopsies was used.
-Of those biopsies, 462 were healthy tissue, 61 were tumor adjacent normal tissue, and 311 were tumors.
-These samples are referred to as the discovery cohort.
+\subsection*{Sample selection}
+Bulk RNA-seq from a curated set of 1,139 flash frozen colorectal tissue samples was re-processed and analyzed for the presence of unannotated HERV transcripts \citep{Dampier2020}.
+For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 833 colon biopsies was used.
+Of those biopsies, 462 were from healthy, 62 were from tumor adjacent normal, and 309 were from tumor tissue.
+This set of samples is referred to as the discovery cohort.
 
 For validation of discovered associations, two independent cohorts were used.
-The first (independent cohort \#1) included single end sequencing from 275 colon biopsies of which NUMBER were healthy, NUMBER were tumor adjacent, and NUMBER were tumor.
+The first (independent cohort \#1) included single end sequencing from 276 colon biopsies of which 12 were healthy, 56 were tumor adjacent, and 208 were tumor.
 The second (independent cohort \#2) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
@@
 
-Samples were gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a recently generated dataset of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
+Samples were originally gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a recently generated dataset of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
 A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al. \citep{Dampier2020}.
@@
 
-\subsection*{Human Endogenous Retrovirus Gene Annotation}
+\subsection*{Human endogenous retrovirus annotation}
 HERVs were identified using the genomic coordinates (GRCh38) and feature names of exons and spliced transcripts of HERV elements from a recently published human cancer transcriptome annotated with the GENCODE v24 basic gene model \citep{Frankish2018} and a model of genome-wide repeat elements \citep{Attig2019}.
@@
 Transcripts were selected based on three criteria: i) overlap of at least one exon with a long terminal repeat element, ii) classification of at least one exon as HERV derived, and iii) abscence of any exon overlapping an annotated GENCODE gene.
-All transcripts labeled with the same repeat element identifier (i.e. repeat element class along with genomic coordinates) were considered isoforms of the same HERV gene.
+All transcripts labeled with the same repeat element identifier (i.e. repeat element class along with genomic coordinates) were considered isoforms of the same HERV locus.
 Genomic coordinates were extracted for all constituent exons of selected transcripts to permit inference of splice sites and exclusion of intronic sequences.
-For display purposes, repeat element identifier genomic coordinates were converted to cytobands.
+For display purposes, repeat element identifier genomic coordinates were converted to cytobands based on the cytoBand table downloaded from the UCSC Genome Browser \citep{Furey2003, Navarro2020}.
 
-\subsection*{Transcript Quantification}
+\subsection*{Transcript quantification}
 Transcript quantification was performed with Salmon v1.2.1 \citep{Patro2017} in mapping-based mode with the \verb|--gcBias| and \verb|--validateMappings| flags set.
-Inputs to the Salmon \verb|quant| command were RNA-seq reads in FASTQ format and the Salmon index.
-To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34  transcript sequences \citep{Frankish2018} and the list of unannotated HERV transcript sequences.
+Inputs to the Salmon \verb|quant| command were RNA-seq reads in FASTQ format and a custom Salmon index.
+To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34  transcript sequences \citep{Frankish2018} and the list of selected HERV transcript sequences.
 GENCODE sequences were downloaded from the GENCODE FTP site \citep{GENCODE-transcripts}.
-HERV sequences were extracted from the GRCh38 reference genome \citep{ENCODE-GRCh38} using the BEDTools \citep{Quinlan2010} \verb|getfasta| command with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts.
+HERV sequences were extracted from the GRCh38 reference genome \citep{ENCODE-GRCh38} using the BEDTools \verb|getfasta| command \citep{Quinlan2010} with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts mentioned above.
 Following the recommendation of Salmon's authors \citep{SalmonDecoys}, contig sequences from the GRCh38 reference genome were subsequently added to the target transcript sequences as decoy sequences to generate a decoy-aware transcriptome for indexing.
 
-\subsection*{Gene Expression Analysis}
-Associations between HERV gene expression and tissue phenotype were tested using the \emph{Bioconductor} suite of genomics analysis packages \citep{bioc} in the R statistical programming language \citep{R}.
-Specifically, \emph{tximport} \citep{Soneson2015} was used to extract transcript length and abundance estimates from Salmon and summarize them to the gene level for input into \emph{DESeq2} \citep{Love2014}.
-Count matrices of HERV and GENCODE genes were prefiltered to select genes with a count of at least one read in at least half of the samples in each cohort.
+\subsection*{Proviral locus and protein coding gene expression analysis}
+Associations between HERV locus expression and tissue phenotype were tested using the \emph{Bioconductor} suite of genomics analysis packages \citep{bioc} in the R statistical programming language \citep{R}.
+Specifically, \emph{tximport} \citep{Soneson2015} was used to extract transcript length and abundance estimates from Salmon output and summarize them by HERV locus and GENCODE gene for input into \emph{DESeq2} \citep{Love2014}.
+Count matrices of HERV loci and GENCODE genes were prefiltered to select loci and genes with a count of at least one read in at least half of samples in each cohort.
 Factors to normalize counts for sequencing depth and transcript length were calculated with \emph{DESeq2}.
-Latent factors of non-biological variation were estimated with \emph{sva} \citep{sva}.
-Genes of interest for association testing were selected in several steps.
+Latent factors of non-biological variation were estimated with \emph{sva} \citep{sva} as described in the Supplementary Methods.
+Loci and genes of interest for association testing were selected in several steps.
 
-First, counts of all genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \citep{Ritchie2015}.
+First, counts of all loci and genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \citep{Ritchie2015}.
 Second, adjusted counts of all annotated genes were filtered to retain only protein coding genes.
-Third, median and standard deviation of protein coding gene expression were calculated separately for each tissue phenotype in each cohort.
-Finally, adjusted counts of HERV genes were filtered to retain only HERVs expressed within a single standard deviation of median protein coding gene expression in at least one tissue phenotype.
+Third, the median and standard deviation of protein coding gene expression were calculated separately for each tissue phenotype in each cohort.
+Finally, adjusted counts of HERV loci were filtered to retain only HERVs expressed within a single standard deviation of median protein coding gene expression in at least one tissue phenotype.
 
-Counts of all protein coding genes and HERVs with sufficient expression were modeled with phenotype and latent factors as explanatory variables using \emph{DESeq2}.
+Counts of all HERV loci and protein coding genes with sufficient expression were modeled with phenotype and latent factors as explanatory variables using \emph{DESeq2}.
 Significant associations between gene count and tissue phenotype were identified by testing whether the absolute value of the effect of phenotype was at least two fold change for a given gene.
 Multiple hypothesis testing correction was performed using a false discovery rate of 5\%.
+
+All figures were produced with \emph{ggplot2} \citep{Wickham2016} as described in the Supplementary Methods.

### 2020-12-31 manuscript/secs/results.tex
finish text, redo figures

@@
-\subsection*{Identification of human endogenous retroviral transcripts in colorectal transcriptomes}
+\subsection*{Identification of non-canonical human endogenous retroviral transcripts in colorectal transcriptomes}
 To define an inclusive set of non-canonical, transcribed HERV elements against which to compare human colorectal transcriptomes, we started with a recently published model of the human cancer transcriptome assembled \emph{de novo} from TCGA samples and annotated with RepeatMasker using the Dfam database \citep{Attig2019}.
@@
 
-To estimate gene and proviral abundance in colorectal tissue, RNA-seq reads from a discovery cohort of 834 colorectal samples were quasi-mapped to a reference transcriptome composed of GENCODE annotated transcripts combined with the selected 16,336 unannotated HERV transcripts (see Methods).
+To estimate gene and proviral abundance in colorectal tissue, RNA-seq reads from a discovery cohort of 833 colorectal samples were quasi-mapped to a reference transcriptome composed of GENCODE annotated transcripts combined with the selected 16,336 unannotated HERV transcripts (see Methods).
 Phenotypic and demographic characteristics of the discovery cohort are presented in Table \ref{tab:discovery}.
@@
 
-To identify HERV proviruses more likely to succeed as biomarkers and function like protein coding genes, we filtered proviruses by expression level relative to GENCODE protein coding genes.
+To identify HERV proviruses with potential to function as protein coding genes and succeed as biomarkers or immunotargets, we filtered HERV loci by expression level relative to GENCODE protein coding genes.
 Of the 1,513 unique loci detected, 150 were expressed at levels within one standard deviation of the median of canonical protein coding genes (Figure \ref{fig:pre_sum}A).
 These 150 moderately expressed proviruses were retained for association testing along with GENCODE protein coding genes.
-Approximately half were from HERV class K or H (Figure \ref{fig:pre_sum}B).
+Approximately half were from HERV group K or H (Figure \ref{fig:pre_sum}B).
 
@@
 We did not investigate this point further because the intersection of the tumor specific proviruses from both comparisons yielded 11 HERV loci clearly unaffected by field effect bias and expressed in a tumor specific pattern (Figure \ref{fig:dge_res}C).
-These 11 HERVs could serve as novel tumor biomarkers or as templates for translation of proteins to which immune tolerance may not have developed, so we sought to validate their tumor specificity in an independent cohort.
+These 11 HERVs could serve as novel tumor biomarkers or as templates for translation of immunogenic peptides, so we sought to validate their tumor specificity in an independent cohort.
 
-To verify the consistency of associations between proviral expression and tissue phenotype, we applied the same methods used in the discovery cohort to an independent cohort (cohort \#1) of 275 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 207 tumor samples (Supplementary Table 1).
+To verify the consistency of associations between proviral expression and tissue phenotype, we applied the same methods used in the discovery cohort to an independent cohort (cohort \#1) of 276 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 208 tumor samples (Supplementary Table 1).
 Of the 150 HERVs moderately expressed (i.e.
@@
 Nominally significant (i.e.
-\emph{P} < 0.05) tumor over expression relative to either healthy or tumor adjacent samples was considered for the purpose of validation due to the small number of healthy samples.
+\emph{P} < 0.05) tumor over-expression relative to either healthy or tumor adjacent samples was considered for the purpose of validation due to the small number of healthy samples.
 Seven of the 11 tumor specific HERVs identified in the discovery cohort were validated in the independent cohort (Table \ref{tab:val}).
@@
 Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (see Supplementary Table 2 for cohort characteristics).
-Only 48 HERV loci were tested in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples was highly concordant (Pearson \emph{R} = 0.93) (Supplementary Figure 1).
+Only 48 HERV loci were moderately expressed in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples was highly concordant (Pearson \emph{R} = 0.93) (Supplementary Figure 1).
 The high concordance of estimates indicated the latent factor analysis employed in the discovery cohort adequately controlled the effect of confounders.
@@
 Subsets of HERV loci from each group demonstrated higher expression in tumors, but these were exceptions.
-To verify this result, we tested for differential expression in the discovery cohort using an alternative HERV transcriptome recently compiled by manual curation of published HERV sequences \citep{Grabski2020}.
-The alternative transcriptome included 91 HERV-K (HML-2) sequences.
-All but one of the six differentially expressed HERV transcripts were higher in healthy samples, indicating our finding was not due to our definition of HERVs (Supplementary Figure 2).
+
+To verify this surprising result, we tested for differential expression in the discovery cohort using an alternative HERV transcriptome recently compiled by manual curation of published but nevertheless poorly annotated HERV-K (HML-2) sequences \citep{Grabski2020}.
+The alternative transcriptome included 91 proviruses.
+A total of six were differentially expressed between healthy and tumor samples (\emph{P} adjusted < 0.05).
+All but one of the six were higher in healthy samples, indicating healthy specific over expression was not an artifact of our customized HERV transcriptome (Supplementary Figure 2).

### 2020-12-31 manuscript/supp.tex
finish text, redo figures

@@
 
-    \title{Supplement to Potential Novel Immunotarget Revealed in Mega Analysis of Colorectal Transcriptomes}
+    \title{Supplement to Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer}
 
@@
 \section*{Supplementary Methods}
-\input{sups/section.tex}
+\input{sups/supmethods.tex}
 
@@
 \section*{Supplementary Figure Legends}
-\input{sups/section.tex}
+\input{sups/suplegends.tex}
 
@@
   }
-  \caption{Boxplots re-demonstrating healthy-specific over-expression using a different HERV transcriptome}
+  \caption{Boxplots re-demonstrating healthy specific over expression using a novel HERV-K transcriptome}
   \label{fig:grabski_method}

### 2020-12-31 manuscript/sups/supmethods.tex
finish text, redo figures

\subsection*{Latent factor analysis}
To estimate the confounding influences of latent factors, including demographic variables with missing data, surrogate variable analysis was performed using the \emph{sva} package \citep{sva}.
First, the number of significant latent factors was estimated with the \verb|num.sv| function.
For this estimation, the Buja Eyuboglu method was used because it estimated a lower, more tractable number of latent factors.
The estimation was performed using a full model with two terms, one for the intercept and one for tissue phenotype.
Second, the latent factors (i.e. surrogate variables) themselves were estimated with the \verb|svaseq| function.
For this estimation, the same full model with intercept and phenotype terms was used, and the null model was set to include an intercept term alone.
The surrogate variables returned from this process included latent factors affecting every sample and were included as variables in the sample annotation matrix along with tissue phenotype.

\subsection*{Data vizualization}
Results of high dimensional data analysis were displayed in figures generated with the \emph{ggplot2} package \citep{Wickham2016}.
Multipanel figures were assembled using the \emph{cowplot} package \cite{Wilke2020}.

Expression levels used for vizualization were based on normalized, transformed, and adjusted counts.
Normalization was performed with the \emph{DESeq2} package \citep{Love2014} using the \verb|estimateSizeFactors| and then \verb|counts| functions.
The \verb|counts| function was used with the \verb|normalized| argument set to TRUE.
Transformation was performed with \emph{DESeq2} using the variance stabilizing transformation on normalized counts.
The variance stabilizing transformation includes a log2 transformation.
Finally, adjustment of normalized and transformed counts for latent factors was performed with \emph{limma} using the \verb|remoteBatchEffect| function.
The normalized, transformed, and adjusted counts were used in histograms to represent expression levels of GENCODE genes and HERV loci.
These counts were also overlaid atop boxplots to represent expression levels of tumor specific HERV loci.
Lastly, the same counts were used to calculate Z-scores to represent expression levels of all HERV loci tested.

Effect sizes (i.e. $log2(fold change)$) used for vizualization of correlations between different analyses were calculated by \emph{DESeq2} regression models and used without shrinkage.
Effect sizes used for volcano plots (i.e. scatter plots of $log2(fold change)$ versus $-log10(\emph{P})$) were shrunken with the \emph{ashr} package \citep{Stephens2020}.

### 2020-12-31 scripts/res_figs.R

@@ -166,7 +166,7 @@ fig_prelim_expr <- function() {
     hrvnums <- run_hrvnums2(l$hervs, l$ddsobj, l$dgeobj)
     hrvnames <- rownames(l$dgeobj$adj)[grepl("HERV", rownames(l$dgeobj$adj))]
     hrvcats <- count_herv_cats(hrvnames)
-    sumlist <- setNames(list(hrvnums, hrvcats), c("HERV Loci", "HERV Classes"))
+    sumlist <- setNames(list(hrvnums, hrvcats), c("HERV Loci", "HERV Groups"))
     pl <- plot_hrvsum(sumlist)
     pl[[1]] <- pl[[1]] +
         labs(title="HERV Elements Studied") +
@@ -376,6 +376,7 @@ plot_heat_big <- function(df) {
 
 
 plot_heat_small <- function(df) {
+    ## this turned out to be too problematic to interpret and is deprectated
     df <- df %>% dplyr::group_by(herv_cl, pheno) %>%
         dplyr::summarise(Z=median(Z), n=n())
     ggp <- ggplot(df, aes(x=pheno, y=herv_cl, fill=Z)) +
@@ -388,14 +389,15 @@ plot_heat_small <- function(df) {
 
 
 write_heat <- function(plotlist, target) {
-    p_width <- 7
-    p_height <- 3.5
+    p_width <- 6   ## 7 for A,B
+    p_height <- 3  ## 3.5 for A,B
     fig <- cowplot::plot_grid(
         plotlist=plotlist,
         nrow=1,
-        ncol=2,
-        labels=c("A", "B"),
-        rel_widths=c(1, 0.6)
+        ncol=1,
+        # labels=c("A", "B"),
+        labels=c(NULL)
+        # rel_widths=c(1, 0.6)
     )
     pdf(file=target, width=p_width, height=p_height)
     print(fig)

### 2020-12-31 scripts/res_hervsum.R

@@ -54,7 +54,7 @@ fill_params <- function(l, params) {
         params[["ggp_sub"]] <- paste("From", n, "starting transcripts")
         l <- l[names(l) != "Total Transcripts"]
     } else {
-        params[["ggp_title"]] <- "Classes of HERV Loci Tested"
+        params[["ggp_title"]] <- "Groups of HERV Loci Tested"
         params[["ggp_sub"]] <- element_blank()
     }
     params[["ggp_ylab"]] <- "Counts"

### 2020-12-31 scripts/res_topherv.R

@@ -62,9 +62,9 @@ format_counts <- function(tophervs, dgeobj) {
 
 plot_counts <- function(dfs) {
     cat("Making box plots\n")
-    color_values <- c("blue", "grey", "red")
-    ggp_title <- "Expression Levels of Select HERV Loci"
-    ggp_ylab <- "Scaled Counts"
+    # color_values <- c("blue", "grey", "red")
+    ggp_title <- "Tumor Specific HERV Loci"
+    ggp_ylab <- "Expression Level"
     lidx <- 0
     pl <- list()
     for (df in dfs) {
@@ -73,7 +73,8 @@ plot_counts <- function(dfs) {
             geom_boxplot(outlier.size=-1, position=position_dodge2()) +
             geom_point(size=0.25, alpha=0.5, position=position_jitterdodge()) +
             labs(title=ggp_title, x=element_blank(), y=ggp_ylab) +
-            scale_colour_manual(values=color_values) +
+            # scale_colour_manual(values=color_values) +
+            scale_colour_grey(start=0.8, end=0.2) +
             ggp_theme_box
         if (cohort == "A" & length(levels(df$herv_id)) < 6) {
             ggp <- ggp +

### 2020-12-31 scripts/results.R

@@ -72,7 +72,7 @@ herv_element_summary <- function() {
     hrvnums[["Expressed HERV Loci"]] <- sum(grepl("HERV", rownames(dgeobj$dds)))
     hrvnames <- rownames(dgeobj$dds)[grepl("HERV", rownames(dgeobj$dds))]
     hrvcats <- count_herv_cats(hrvnames)
-    sumlist <- setNames(list(hrvnums, hrvcats), c("HERV Loci", "HERV Classes"))
+    sumlist <- setNames(list(hrvnums, hrvcats), c("HERV Loci", "HERV Groups"))
     pl <- plot_hrvsum(sumlist)
     target_dir <- Sys.getenv("plot_dir")
     check_dir(target_dir)
@@ -152,17 +152,8 @@ validation_tests <- function() {
 }
 
 
-if (validate == 1) {
-    l2 <- validation_tests()
-    print(l2)
-} else {
-    main()
-}
-
-
-### DEVS
-
 make_figures <- function() {
+    cat("Making figures for manuscript\n")
     fig <- list()
     fig[[1]] <- fig_supp_cor_field_val()
     fig[[2]] <- fig_prelim_expr()
@@ -170,4 +161,17 @@ make_figures <- function() {
     fig[[4]] <- fig_heat()
 }
 
-make_figures()
+
+if (validate == 1) {
+    l2 <- validation_tests()
+    print(l2)
+    make_figures()
+} else {
+    main()
+}
+
+
+### DEVS
+devs <- function() {
+    ## currently nothing under development
+}

### 2020-12-31 scripts/results.sh

@@ -3,7 +3,7 @@
 #SBATCH --ntasks=1
 #SBATCH --ntasks-per-node=1
 #SBATCH --cpus-per-task=1
-#SBATCH --mem=05000
+#SBATCH --mem=10000
 #SBATCH --time=00:30:00
 #SBATCH --partition=standard
 #SBATCH --account=chd5n_alloc
@@ -14,7 +14,7 @@
 ## -- for use in snakemake workflow
 ## -- usage: snakemake results
 
-## -- perform: 1-2m, 3.4GB for cohort A
+## -- perform: 1-2m, >4GB for cohort A
 
 
 script_name=results.sh

### 2020-12-31 manuscript/main.tex
update figure legends

@@
   }
-  \caption{Differential expression of unannotated human endogenous retroviral loci}
+  \caption{Differential expression of unannotated human endogenous retroviral loci in colorectal transcriptomes}
   \label{fig:dge_res}
@@
   }
-  \caption{Global estimates of proviral expression across tissue phenotypes}
+  \caption{Global estimates of unannotated human endogenous retroviral expression across tissue phenotypes.}
   \label{fig:heat}

### 2020-12-31 manuscript/secs/legends.tex
update figure legends

@@
-\textbf{Figure 1.} Overview of human endogenous retroviral element discovery in colorectal cancer.
-A, Barplot demonstrating
-B, Barplot demonstrating
-C, Histogram demonstrating
-D, Histogram demonstrating
+\textbf{Figure 1.} Summary of unannotated human endogenous retroviral sequences found in colorectal transcriptomes. \\
+a) Barplot demonstrating numbers of human endogenous retroviral elements considered in this study.
+Numbers of elements are shown on the y-axis, and categories of retroviral elements considered are shown on the x-axis. \\
+b) Barplot demonstrating numbers of human endogenous retroviral elements in each of several retroviral groups expressed within one standard deviation of the median of canonical protein coding genes.
+Numbers of elements are shown on the y-axis, and groups of retroviral elements are shown on the x-axis. \\
+c) Histogram demonstrating distributions of expression levels of canonical protein coding genes (i.e.
+GENCODE genes) and detected human endogenous retroviral loci across all samples in the discovery cohort.
+\\
+d) Histogram demonstrating distributions of expression levels of canonical protein coding genes (i.e.
+GENCODE genes) and detected human endogenous retroviral loci across tumor samples in the discovery cohort.
+\\
 
-\textbf{Figure 2.} Unannotated proviral elements demonstrate tissue phenotype specificity in colorectal transcriptomes.
-A, Scatterplot (i.e. volcano plot) demonstrating
-B, Scatterplots (i.e. volcano plot) demonstrating
-C, Boxplots demonstrating
+\textbf{Figure 2.} Differential expression of unannotated human endogenous retroviral loci in colorectal transcriptomes. \\
+a) Scatterplot (i.e.
+volcano plot) demonstrating relationship between $log2(fold change)$ and $-log10(\emph{P})$ for all human endogenous retroviral loci and protein coding genes tested in the discovery cohort with baseline expression set by the healthy phenotype.
+$log2(fold change)$ is shrunken (see Supplementary Methods).
+\emph{P} values are not corrected for multiple hypothesis testing.
+Grey points are either protein coding genes or non-differentially expressed proviral loci.
+Blue points are proviral loci negatively associated with the tumor phenotype.
+Red points are proviral loci positively associated with the tumor phenotype.
+\emph{TUM} = tumor, \emph{HLT} = healthy. \\
+b) Scatterplot (i.e.
+volcano plot) demonstrating relationship between $log2(fold change)$ and $-log10(\emph{P})$ for all human endogenous retroviral loci and protein coding genes tested in the discovery cohort with baseline expression set by the tumor adjacent phenotype.
+$log2(fold change)$ is shrunken (see Supplementary Methods).
+\emph{P} values are not corrected for multiple hypothesis testing.
+Grey points are either protein coding genes or non-differentially expressed proviral loci.
+Blue points are proviral loci negatively associated with the tumor phenotype.
+Red points are proviral loci positively associated with the tumor phenotype.
+\emph{NAT} = normal adjacent to tumor. \\
+c) Boxplots demonstrating expression levels by phenotype of 11 proviral loci positively associated with the tumor phenotype in the discovery cohort (\emph{P} adjusted < 0.05). \\
 
-\textbf{Figure 3.} Global tendency of healthy colorectal transcriptomes to harbor relatively higher expression of unannotated proviral elements.
-A, Heatmap demonstrating
-B, Heatmap demonstrating
+\textbf{Figure 3.} Global estimates of unannotated human endogenous retroviral expression across tissue phenotypes. \\
+Heatmap demonstrating standardized expression levels (i.e. Z-scores) of human endogenous retroviral loci tested in the discovery cohort across all samples in the cohort.
+Proviruses (N = 150) are listed on the y-axis and ordered according to group.
+Group O is an acronym for "Other".
+Samples (N = 833) are listed on the x-axis and ordered according to phenotype.
+Darker panels indicate higher standardized expression.

### 2020-12-31 manuscript/sups/suplegends.tex
update figure legends

\textbf{Figure 1.} Correlations of effect sizes across studies and cohorts. \\
a) Scatterplot demonstrating relationship between estimates of effect sizes (i.e. $log2(fold change)$) of the tumor phenotype on protein coding genes in the discovery cohort of this analysis (i.e. HERV Study) and Cohort A of a prior analysis of the same samples, as reported in \citep{Dampier2020} (i.e. Field Effect Study).
Baseline expression is set by the healthy phenotype.
Only effect sizes on genes tested in both studies are shown.
\emph{R} = Pearson correlation coefficient. \\
b) Scatterplot demonstrating relationship between estimates of effect sizes (i.e. $log2(fold change)$) of the tumor phenotype on protein coding genes in the discovery cohort of this analysis (i.e. HERV Study) and Cohort A of a prior analysis of the same samples, as reported in \citep{Dampier2020} (i.e. Field Effect Study).
Baseline expression is set by the tumor adjacent phenotype.
Only effect sizes on genes tested in both studies are shown. \\
c) Scatterplot demonstrating relationship between estimates of effect sizes (i.e. $log2(fold change)$) of the tumor phenotype on proviral loci in the discovery cohort of this analysis and independent cohort \#1.
Baseline expression is set by the healthy phenotype.
Only effect sizes on loci tested in both cohorts are shown. \\
d) Scatterplot demonstrating relationship between estimates of effect sizes (i.e. $log2(fold change)$) of the tumor phenotype on proviral loci in the discovery cohort of this analysis and independent cohort \#2.
Baseline expression is set by the tumor adjacent phenotype.
Only effect sizes on loci tested in both cohorts are shown.\\

\textbf{Figure 2.} Differential expression of unannotated human endogenous retroviral loci in colorectal transcriptomes using a novel HERV-K reference transcriptome. \\
Boxplots demonstrating expression levels by phenotype of six proviral loci associated with the tumor phenotype (\emph{P} adjusted < 0.05) in samples from the discovery cohort of this analysis analyzed using an alternative proviral reference transcriptome as described in \citep{Grabski2020}.
Five of the six significant associations are negative.

### 2020-12-31 scripts/res_figs.R
change to lowercase for multipanel figure labels

@@ -5,7 +5,7 @@ plot_cor_select <- function(plotlist) {
     cwp <- cowplot::plot_grid(
         plotlist[["TUMvHLT"]], plotlist[["TUMvNAT"]],
         nrow=1, ncol=2,
-        labels=c("A", "B")
+        labels=c("a", "b")
     )
     return(cwp)
 }
@@ -15,7 +15,7 @@ plot_cor_val_select <- function(plotlist) {
     cwp <- cowplot::plot_grid(
         plotlist[[1]], plotlist[[2]],
         nrow=1, ncol=2,
-        labels=c("C", "D")
+        labels=c("c", "d")
     )
     return(cwp)
 }
@@ -147,7 +147,7 @@ write_prelim_expr <- function(plotlist, target) {
         plotlist=plotlist,
         nrow=2,
         ncol=2,
-        labels=c("A", "B", "C", "D")
+        labels=c("a", "b", "c", "d")
     )
     pdf(file=target, width=p_width, height=p_height)
     print(fig)
@@ -186,7 +186,7 @@ plot_volcano_select <- function(plotlist, target) {
         plotlist[["TUMvHLT"]], plotlist[["TUMvNAT"]],
         nrow=1,
         ncol=2,
-        labels=c("A", "B")
+        labels=c("a", "b")
     )
     return(cwp)
 }
@@ -197,7 +197,7 @@ plot_counts_select <- function(plotlist, target) {
         plotlist=plotlist,
         nrow=2,
         ncol=1,
-        labels=c("C", NULL)
+        labels=c("c", NULL)
     )
     return(cwp)
 }

### 2020-12-31 manuscript/sups/suplegends.tex
fix supp legend references

@@
 $log2(fold change)$) of the tumor phenotype on protein coding genes in the discovery cohort of this analysis (i.e.
-HERV Study) and Cohort A of a prior analysis of the same samples, as reported in \citep{Dampier2020} (i.e.
+HERV Study) and Cohort A of a prior analysis of the same samples, as reported in Dampier et al.
+\citep{Dampier2020} (i.e.
 Field Effect Study).
@@
 \textbf{Figure 2.} Differential expression of unannotated human endogenous retroviral loci in colorectal transcriptomes using a novel HERV-K reference transcriptome. \\
-Boxplots demonstrating expression levels by phenotype of six proviral loci associated with the tumor phenotype (\emph{P} adjusted < 0.05) in samples from the discovery cohort of this analysis analyzed using an alternative proviral reference transcriptome as described in \citep{Grabski2020}.
+Boxplots demonstrating expression levels by phenotype of six proviral loci associated with the tumor phenotype (\emph{P} adjusted < 0.05) in samples from the discovery cohort of this analysis analyzed using an alternative proviral reference transcriptome as described in Grabski et al.
+\citep{Grabski2020}.
 Five of the six significant associations are negative.

### 2021-01-01 scripts/xelatex.sh
add pandoc conversion from latex to word for coauthors

@@ -1,4 +1,7 @@
-## this script is experimental
+## this script is meant to be run from snakemake pipeline
+## it will compile a complete pdf including figures, tables, and refs
+## it will also compile a rough docx file for sharing with collaborators
+## sample useage: snakemake manuscript
 
 fig1=$1
 fig2=$2
@@ -21,7 +24,8 @@ cp $spt1 ../manuscript/tabs/spt1.tex
 cp $spt2 ../manuscript/tabs/spt2.tex
 
 printf '%s' "Adding texlive to \$PATH\n"
-export PATH=$PATH:/home/chd5n/apps/texlive/2020/bin/x86_64-linux
+export PATH=$PATH:/home/chd5n/apps/texlive/2020/bin/x86_64-linux; xelatex -v
+export PATH=$PATH:/home/chd5n/apps/pandoc-2.10/bin; pandoc -v
 
 cd ../manuscript/
 rm main.{aux,bbl,blg,log,pdf}
@@ -38,3 +42,5 @@ xelatex supp
 bibtex supp
 xelatex supp
 xelatex supp
+
+pandoc --standalone --output=docx/main.docx --from=latex --to=docx main.tex

### 2021-01-19 manuscript/main.tex
add pandoc citation conversion and fix docx formatting

@@
 
-problems with AMA style file
-https://tex.stackexchange.com/questions/55020/updated-ama-bst-file
-i modified vancouver.bst to get proper author lists
+%% citation issues
+- problems with AMA style file
+- https://tex.stackexchange.com/questions/55020/updated-ama-bst-file
+- i modified vancouver.bst to get proper author lists > myvancouver.bst
+- pandoc conversion to docx requires csl
+- wget -O ama.csl https://github.com/citation-style-language/styles/blob/master/american-medical-association.csl?raw=true
+
+%% word format issues
+- see https://pandoc.org/MANUAL.html, --reference-doc= for DOCX
+- To produce a custom reference.docx, first get a copy of the default reference.docx:
+pandoc --print-default-data-file reference.docx > custom-reference.docx
+Then open custom-reference.docx in Word, modify the styles as you wish, and save the file.
+Modified: Normal, Abstract, Author, Title, Subtitle, Heading 1, Heading 2, Bibliography, Body Text, First Paragraph, Verbatim Char

### 2021-01-19 scripts/xelatex.sh

@@ -43,4 +43,12 @@ bibtex supp
 xelatex supp
 xelatex supp
 
-pandoc --standalone --output=docx/main.docx --from=latex --to=docx main.tex
+pandoc \
+    --standalone \
+    --from=latex \
+    --to=docx \
+    --reference-doc=docx/custom-reference.docx \
+    --bibliography=main.bib \
+    --csl=ama.csl \
+    --output=docx/main.docx \
+    main.tex

### 2021-01-24 scripts/res_characterize.R
major additions: roc, surv, characterization plots

@@ -1,0 +1,286 @@
+## to be sourced in results.R
+
+
+read_charac_inputs <- function(cohort) {
+    target <- paste0(Sys.getenv("rdata_dir"), "dge_", cohort, ".Rda")
+    dgeobj <- load_dgeobj(target)
+    restabs <- fill_restabs()
+    l <- extract_main_hervs(restabs)
+    herv_ids <- intersect(
+        l[["A"]][["res_A_CRCvHLT_hrv"]], l[["A"]][["res_A_CRCvNAT_hrv"]]
+    )
+    names(herv_ids) <- map_cytoband(herv_ids)
+    names(herv_ids) <- gsub("HERV", "", names(herv_ids))
+    names(herv_ids) <- gsub("_", " ", names(herv_ids))
+    ## CTLA4 not among genes tested
+    # markers <- c("ENSG00000120217", "ENSG00000188389")
+    # names(markers) <- c("CD274", "PDCD1")
+    markers <- c(
+        "ENSG00000145649", "ENSG00000100453", "ENSG00000100450",
+        "ENSG00000113088", "ENSG00000197540", "ENSG00000180644"
+    )
+    names(markers) <- c("GZMA", "GZMB", "GZMH", "GZMK", "GZMM", "PRF1")
+    return(setNames(
+        list(herv_ids, dgeobj, markers),
+        c("herv_ids", "dgeobj", "markers")
+    ))
+}
+
+
+prep_charac_df <- function(herv_ids, dgeobj, markers, feature) {
+    i <- rownames(dgeobj$adj) %in% c(herv_ids)
+    j <- colData(dgeobj$dds)$phenotype == "CRC"
+    if (feature %in% c("msi_status", "stage")) {
+        j1 <- !is.na(colData(dgeobj$dds)[ , feature])
+        j <- j & j1
+    } else if (feature == "immunomarker") {
+        i <- rownames(dgeobj$adj) %in% c(herv_ids, markers)
+    } else if (feature == "age") {
+        j <- !is.na(colData(dgeobj$dds)$age_days)
+    } else {
+        warning("feature unrecognized, default data.frame provided")
+    }
+    df <- data.frame(dgeobj$adj[i, j])
+    colnames(df) <- colData(dgeobj$dds)$subject_id[j]
+    return(df)
+}
+
+
+explore_charac_markers <- function(herv_ids, markers, df) {
+    ## draft code to determine what to show
+    ## for command line manual execution, not to run
+    l_cor <- list()
[…]

### 2021-01-24 scripts/res_roc.R

@@ -1,0 +1,222 @@
+library(readr)
+library(DESeq2)
+library(ggplot2)
+library(cowplot)
+library(reshape2)
+library(scales)
+library(pROC)
+
+
+source("res_utilities.R")
+source("res_themes.R")
+source("res_val.R")
+
+
+args <- commandArgs(trailingOnly=TRUE)
+cohort <- args[2]
+cat(paste("Cohort:", cohort, "\n"))
+
+
+read_inputs <- function(cohort) {
+    target <- paste0(Sys.getenv("rdata_dir"), "dge_", cohort, ".Rda")
+    dgeobj <- load_dgeobj(target)
+    target <- paste0(Sys.getenv("rdata_dir"), "txi_", cohort, ".Rda")
+    load(target, verbose=TRUE)
+    restabs <- fill_restabs()
+    return(setNames(list(dgeobj, txi, restabs), c("dgeobj", "txi", "restabs")))
+}
+
+
+set_stage <- function(truth) {
+    truth[is.na(truth$stage_r), "stage_r"] <- "nan"
+    truth[truth$phenotype %in% c("HLT", "NAT"), "stage_r"] <- "hlt"
+    l <- list()
+    l[["nan"]] <- NA
+    l[["hlt"]] <- 0
+    l[["i"]] <- 1
+    l[["ii"]] <- 2
+    l[["iii"]] <- 3
+    l[["iii/iv"]] <- 4
+    l[["iv"]] <- 4
+    truth$stage <- unlist(
+        lapply(truth$stage_r, fun <- function(x) {l[[x]]}),
+        use.names=FALSE
+    )
+    return(truth)
+}
+
+
+set_truth <- function(dgeobj) {
+    truth <- data.frame(
+        subject_id=colData(dgeobj$dds)$subject_id,
+        phenotype=colData(dgeobj$dds)$phenotype,
+        sex=colData(dgeobj$dds)$sex,
+        age=colData(dgeobj$dds)$age_days/365.25,
+        stage_r=colData(dgeobj$dds)$stage
+    )
+    truth <- set_stage(truth)
+    return(truth)
+}
+
+
+set_predictor <-function(dgeobj, txi, restabs) {
+    l <- extract_main_hervs(restabs)
+    herv_ids <- intersect(
+        l[["A"]][["res_A_CRCvHLT_hrv"]], l[["A"]][["res_A_CRCvNAT_hrv"]]
+    )
[…]

### 2021-01-24 scripts/res_surv.R

@@ -1,0 +1,201 @@
+library(readr)
+library(DESeq2)
+library(ggplot2)
+library(cowplot)
+library(reshape2)
+library(scales)
+library(pROC)
+library(survival)
+library(survminer)
+
+
+source("res_utilities.R")
+source("res_val.R")
+
+
+read_inputs <- function() {
+    ann <- list()
+    tpm <- list()
+    target <- paste0(Sys.getenv("ann_dir"), "tcga/ann.tsv")
+    ann$tcga <- readr::read_tsv(target)
+    target <- paste0(Sys.getenv("ann_dir"), "tcga/cbioportal.tsv")
+    ann$cbio <- readr::read_tsv(target)
+    target <- paste0(Sys.getenv("ann_dir"), "pheno_A.tsv")
+    ann$a <- readr::read_tsv(target)
+    target <- paste0(Sys.getenv("ann_dir"), "pheno_C.tsv")
+    ann$c <- readr::read_tsv(target)
+    target <- paste0(Sys.getenv("ann_dir"), "pheno_B.tsv")
+    ann$b <- readr::read_tsv(target)
+    target <- paste0(Sys.getenv("rdata_dir"), "txi_A.Rda")
+    load(target, verbose=TRUE)
+    tpm$a <- txi$abundance
+    target <- paste0(Sys.getenv("rdata_dir"), "txi_C.Rda")
+    load(target, verbose=TRUE)
+    tpm$c <- txi$abundance
+    target <- paste0(Sys.getenv("rdata_dir"), "txi_B.Rda")
+    load(target, verbose=TRUE)
+    tpm$b <- txi$abundance
+    restabs <- fill_restabs()
+    return(setNames(list(ann, tpm, restabs), c("ann", "tpm", "restabs")))
+}
+
+
+parse_inputs <- function(ann, tpm, restabs) {
+    ## extract ids of hervs of interest
+    l <- extract_main_hervs(restabs)
+    herv_ids <- intersect(
+        l[["A"]][["res_A_CRCvHLT_hrv"]], l[["A"]][["res_A_CRCvNAT_hrv"]]
+    )
+    ## extract tpms of hervs of interest
+    tpm <- lapply(tpm, fun <- function(x) {
+        x[match(herv_ids, rownames(x)), ]
+    })
+    ## make herv names easy to read
+    tpm <- lapply(tpm, fun <- function(x) {
+        rownames(x) <- map_cytoband(herv_ids)
+        rownames(x) <- gsub("HERV", "", rownames(x))
[…]

### 2021-01-24 scripts/res_themes.R

@@ -5,7 +5,6 @@ ggp_theme_default <- theme(
     panel.background=element_rect(fill="white"),
     panel.grid.major=element_line(color="white"),
     panel.grid.minor=element_line(color="white"),
-    plot.background=element_rect(fill="white"),
     plot.margin=margin(t=1, r=1, b=1, l=1, unit="lines"),
     plot.title=element_text(size=12, face="bold", hjust=0.5),
     plot.subtitle=element_text(size=10, face="plain", hjust=0.5),
@@ -81,6 +80,19 @@ ggp_theme_heat <- ggp_theme_default +
         axis.line.y.left=element_blank(),
         legend.position="top",
         legend.title=element_text(size=9, face="plain")
+    )
+
+
+ggp_theme_charac <- ggp_theme_default +
+    theme(
+        panel.border=element_rect(colour="black", size=1.0, fill=NA),
+        plot.margin=margin(t=0.5, r=0.5, b=0.5, l=0.5, unit="lines"),
+        axis.text.x=element_text(size=8, face="plain"),
+        axis.text.y=element_text(size=8, face="plain"),
+        strip.text=element_text(colour="white"),
+        axis.line.x.bottom=element_blank(),
+        axis.line.y.left=element_blank(),
+        legend.position="none"
     )
 
 

### 2021-01-24 scripts/results.R

@@ -15,10 +15,8 @@
 ## load dgeobj into list and herv-specific results into dataframe
 ## create boxplots showing top herv hits
 
-## try to highlight samples with MSS
-
-## try showing alignments of top hit along with sequence and predictied protein
-## try showing possible HLA binding sites in predicted protein
+## correlation with age, stage, msi status, immunomarkers
+## tumor vs non tumor roc, survival vs expression level
 
 
 BiocManager::install()
@@ -34,13 +32,6 @@ library(EnsDb.Hsapiens.v86)
 library(scales)
 
 
-args <- commandArgs(trailingOnly=TRUE)
-cohort <- args[2]
-validate <- as.numeric(args[3])
-cat(paste("Cohort:", cohort, "\n"))
-cat(paste("Validate:", validate, "\n"))
-
-
 source("res_utilities.R")
 source("res_themes.R")
 source("res_corr.R")
@@ -50,6 +41,14 @@ source("res_topherv.R")
 source("res_infla.R")
 source("res_val.R")
 source("res_figs.R")
+source("res_characterize.R")
+
+
+args <- commandArgs(trailingOnly=TRUE)
+cohort <- args[2]
+validate <- as.numeric(args[3])
+cat(paste("Cohort:", cohort, "\n"))
+cat(paste("Validate:", validate, "\n"))
 
 
 correlation_analysis <- function() {
@@ -103,6 +102,7 @@ topherv_analysis <- function(dgeobj) {
 
 
 main <- function() {
+    cat("Running main() ...\n")
     if (cohort == "A") {
         correlation_analysis()
         dgeobj <- herv_element_summary()
@@ -133,6 +133,7 @@ inflation_tests <- function() {
 
 
 validation_tests <- function() {
+    cat("Running validation_tests() ...\n")
     restabs <- fill_restabs()
 
     pl <- val_corr_analysis(restabs)
@@ -162,10 +163,53 @@ make_figures <- function() {
 }
 
 
+characterization_tests <- function() {
+    cat("Running characterization_tests() ...\n")
+    l <- read_charac_inputs(cohort)
+    herv_levels <- c(
+        "H Xp22.32", "IP10F 9q13", "H 13q33.3", "K9 13q12.11", "IP10F 9p11.2",
+        "H 20p11.23", "1 I 3p22.3", "L 9q21.11", "H 5q15", "E a 7q22.1",
[…]

### 2021-01-24 scripts/xelatex.sh

@@ -10,8 +10,14 @@ tab1=$4
 tab2=$5
 spf1=$6
 spf2=$7
-spt1=$8
-spt2=$9
+spf3=$8
+spf4=$9
+spf5=${10}
+spf6=${11}
+spf7=${12}
+spf8=${13}
+spt1=${14}
+spt2=${15}
 
 cp $fig1 ../manuscript/figs/fig1.pdf
 cp $fig2 ../manuscript/figs/fig2.pdf
@@ -20,6 +26,12 @@ cp $tab1 ../manuscript/tabs/tab1.tex
 cp $tab2 ../manuscript/tabs/tab2.tex
 cp $spf1 ../manuscript/figs/spf1.pdf
 cp $spf2 ../manuscript/figs/spf2.pdf
+cp $spf3 ../manuscript/figs/spf3.pdf
+cp $spf4 ../manuscript/figs/spf4.pdf
+cp $spf5 ../manuscript/figs/spf5.pdf
+cp $spf6 ../manuscript/figs/spf6.pdf
+cp $spf7 ../manuscript/figs/spf7.pdf
+cp $spf8 ../manuscript/figs/spf8.pdf
 cp $spt1 ../manuscript/tabs/spt1.tex
 cp $spt2 ../manuscript/tabs/spt2.tex
 
@@ -29,9 +41,7 @@ export PATH=$PATH:/home/chd5n/apps/pandoc-2.10/bin; pandoc -v
 
 cd ../manuscript/
 rm main.{aux,bbl,blg,log,pdf}
-rm ._main.pdf
 rm supp.{aux,bbl,blg,log,pdf}
-rm ._supp.pdf
 
 xelatex main
 bibtex main

### 2021-01-24 manuscript/main.tex
first set of revisions to manuscript, excluding MAD

@@
-\documentclass[11pt,letterpaper]{article}
+\documentclass[11pt,letterpaper]{article}  %% 3500 words for ASO research article
 
@@
     %\title{Potential Novel Immunotarget Revealed in Mega Analysis of Colorectal Transcriptomes}
-    \title{Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer}
-
-    \author[1]{Christopher H. Dampier, MD}
-    \author[2]{David F. Grabski, MD, PhD}
-    \author[1]{Matthew Devall, PhD}
-    \author[3]{Virginia Diez-Obrero, BS}
-    \author[3]{Ferran Moratalla-Navarro, MS}
-    \author[4]{David Rekosh, PhD}
-    \author[4]{Marie-Louise Hammarskjold, PhD}
-    \author[5]{Sara K. Rasmussen, MD, PhD}
-    \author[3]{Victor Moreno, MD, PhD}
-    \author[1,*]{Graham Casey, PhD}
-
-    \affil[1]{Center for Public Health Genomics, University of Virginia, Charlottesville, Virginia, USA}
-    \affil[2]{Department of Surgery, University of Virginia, Charlottesville, Virginia, USA}
-    \affil[3]{Catalan Institute of Oncology, Barcelona, Spain}
-    \affil[4]{Department of Microbiology, Immunology and Cancer Biology, University of Virginia, Charlottesville, Virginia, USA}
-    \affil[5]{Department of Surgery, Seattle Children's, Seattle, Washington, USA}
-    \affil[*]{Correspondence: Graham Casey, PhD, Center for Public Health Genomics, MSB Room 3238, Department of Public Health Sciences, University of Virginia, P.O.
-Box 800717, Charlottesville, VA 22908-0717, gc8r@virginia.edu}
+    % \title{Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer}
+    \input{secs/title.tex}
 
@@
                 \newline
-                \normalsize\@author
+                \small\@author
             \end{flushleft}
@@
 \maketitle
+\small\textbf{Running Header}: HERV Transcripts in Colorectal Tumors \\
+\small\textbf{Disclosure Statement}: No author has any commercial interest in the subject of study.
 
@@
 \section*{Synopsis}
-\input{secs/synopsis.tex}
+\input{secs/synopsis.tex}  %% 1-3 sentences, 40 words max
 

### 2021-01-24 manuscript/secs/abstract.tex
first set of revisions to manuscript, excluding MAD

--- 2020-12-29 manuscript/secs/abstract.tex
@@
 \textbf{Background}:
-Microsatellite unstable, mismatch repair deficient metastatic cancers can be treated with immune checkpoint inhibitors, but new therapeutic approaches are needed to extend the benefits of immune modulators to the majority of patients with microsatellite stable tumors.
-Cancer specific over expression of human endogenous retroviral elements could be exploited to enhance immune infiltration of microsatellite stable colorectal cancer and render these tumors more susceptible to immune checkpoint inhibition.
-To identify novel targets for immune mediated cancer therapies, we tested for cancer specific over expression of endogenous retroviral transcripts in a large cohort of colorectal transcriptomes.
+Transcripts from novel genomic features such as endogenous retroviruses have potential utility as biomarkers or therapeutic targets, and their characterization could help improve detection and treatment of colorectal cancer.
+To identify uncharacterized transcripts for possible diagnostic and therapeutic applications, we examined cancer specific over expression of unannotated endogenous retroviral mRNA in a large cohort of colorectal transcriptomes.
 
 \textbf{Methods}:
-A recently curated RNA sequencing dataset composed of 1,139 human primary colorectal tissue samples divided into three cohorts was re-processed using a novel annotation of repeat element transcripts to quantify levels of repeat element expression in colorectal transcriptomes.
-A total of NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples were profiled with short read sequencing.
-Transcripts were quantified with Salmon using the GENCODE v36 transcriptome augmented to include unannotated exons from long terminal repeat elements classified as human endogenous retroviruses (HERVs) in the Dfam database.
-Associations between transcript expression and tissue phenotype were tested using DESeq2 in the R statistical language.
+Bulk RNA sequencing reads from human primary colorectal tissue, including 474 healthy, 133 tumor adjacent normal, and 532 tumor samples, were obtained from public sources and from our own data and re-processed using a custom reference transcriptome.
+Transcripts were quantified using Salmon with GENCODE transcripts combined with novel endogenous retroviral transcripts from a recently described cancer transcriptome.
+Associations between transcript expression and tissue phenotype were determined using \emph{DESeq2} in the R statistical programming language.
 
 \textbf{Results}:
-Out of 5,025 unannotated HERV genes measured in 834 samples, including NUMBER healthy, NUMBER tumor-adjacent normal, and NUMBER tumor samples, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
-Over half of the 150 HERV genes with moderate expression were in the H (26) or K (51) classes.
-Tumor specific over expression was observed in 11 of the 150 HERV genes tested.
+Of 5,025 unannotated proviral loci queried in our discovery analysis, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
+Tumor specific over expression was observed in 11 of the 150 loci tested (\emph{P} adjusted < 0.05), and seven were validated in an independent cohort.
 
-\textbf{Conclusions}: The conclusions go here.
+\textbf{Conclusions}:
+In the largest unbiased assessment of unannotated human endogenous retroviral transcript expression in colorectal tissue to date, we identified over 100 proviral loci expressed at levels similar to canonical protein coding genes and seven that could be novel biomarkers of colorectal cancer or targets of immunotherapy.

### 2021-01-24 manuscript/secs/acknowledge.tex
first set of revisions to manuscript, excluding MAD

@@
 GTEx was supported by the Common Fund of the Office of the Director of the National Institutes of Health and by NCI, NHGRI, NHLBI, NIDA, NIMH, and NINDS.
-The data used for the analyses described in this manuscript were obtained from dbGaP accession number phs000424.v7.p2 on2020-07-15.
+The data used for the analyses described in this manuscript were obtained from dbGaP accession number phs000424.v7.p2 on 2020-07-15.
 This research was supported by the Genomics Core Facility of the CWRU School of Medicine's Genetics and Genome Sciences Department.
-Funding for this work was provided by the following NIH Grants: R01 NIH/NCI CA143237 and CA201407 (G.C.), T32 5T32CA163177-07 (Craig Slingluff, MD).
-The authors are grateful to Dr Craig Slingluff from the Department of Surgery at the University of Virginia, whose T32 training grant supported CHD and DFG.
+Funding for this work was provided by the following grants: NIH/NCI R01 CA143237 (GC), NIH/NCI R01 CA201407 (GC), NIH/NCI R01 CA206275 (DR and MLH), NIH/NCI T32 5T32CA163177-07 (Craig Slingluff, MD), Agency for Management of University and Research Grants of the Catalan Government grant 2017SGR723 (VM), Instituto de Salud Carlos III (co-funded by European Regional Development Fund "A way to build Europe" grants PI14-00613 and PI17-00092), and Spanish Association Against Cancer Scientific Foundation grant GCTRA18022MORE (VM).
+The authors are grateful to CERCA Programme, Generalitat de Catalunya for institutional support.
+The authors are also grateful to Dr Craig Slingluff from the Department of Surgery at the University of Virginia, whose T32 training grant provided salary support for CHD and DFG.
+Partial salary support for DR and MLH was provided by the Charles H. Ross Jr and Myles H. Thaler Professorship endowments at the University of Virginia.

### 2021-01-24 manuscript/secs/discussion.tex
first set of revisions to manuscript, excluding MAD

@@
-In this study, we performed the largest unbiased assessment of transcribed but unannotated human endogenous retroviral elements in colorectal tissue to date.
-We found several unique proviral loci with tumor specific over expression and many other loci with relatively high expression in healthy samples.
-Human endogenous retroviral elements are normally transcriptionally repressed repetitive sequences that are often de-repressed in epigenetically altered cancer cells \citep{Babaian2016}.
+In this study, we performed the largest unbiased assessment of transcribed, previously unannotated human endogenous retroviral elements in colorectal tissue to date.
+We found several unique proviral loci with tumor specific over expression and other loci with relatively high expression in healthy samples.
+Human endogenous retroviral elements are often transcriptionally repressed repetitive sequences that are frequently de-repressed in epigenetically altered cancer cells \citep{Babaian2016}.
 HERV transcripts have been shown to engage pattern recognition receptors and activate interferon responses in cancer cell lines \citep{Chiappinelli2015, Roulois2015} and have been associated with tumor infiltration by T regulatory cells \citep{Desai2017}, response to immune checkpoint blockade in urothelial carcinoma, and survival in subjects with CRC \citep{Solovyov2018}.
-Our results are important because the tumor specific HERV sequences we identified could be used to distinguish tumor from non-tumor tissue for screening and surveillance and to predict or even modify response to immunotherapies.
+Our results are important because the tumor specific HERV sequences we identified could be used to distinguish tumor from non-tumor tissue in screening and surveillance and could be used to predict or even modify response to immunotherapies.
 
@@
 Furthermore, our analysis was not restricted to a particular group of proviruses.
-The unbiased nature of our assessment provides important context for the community of researchers selecting genomic features to develop as biomarkers and immunotargets for CRC.
+The unbiased nature of our assessment provides important context for selecting genomic features to develop as potential biomarkers or immunotargets for CRC.
 
 We note that the tumor specific provirus we mapped to the high resolution cytoband Xp22.32 appears to be the same HERV-H provirus previously mapped to the lower resolution cytoband Xp22.3 and found to be over expressed in CRC \citep{Wentzensen2007}.
-Remarkably, this sequence was subsequently found to encode an \emph{env} protein including at least three immunogenic epitopes \citep{Mullins2012}.
-The epitopes were successfully used to sensitize human cytotoxic T cells in culture, and the sensitized lymphocytes were shown to selectively lyse CRC cell lines expressing the HERV Xp22.3 provirus.
+Remarkably, this sequence was subsequently found to encode an \emph{env} protein that contains at least three immunogenic epitopes \citep{Mullins2012}.
+These epitopes were successfully used to sensitize human cytotoxic T cells in culture, and the sensitized lymphocytes were shown to selectively lyse CRC cell lines expressing the HERV Xp22.3 provirus.
 Other HERV loci identified in our analysis may encode peptides with similar immunogenic potential.
-Identification of protein sequences that could be translated from tumor specific proviral transcripts is therefore an important goal of our future work.
+Identification of protein sequences that could be translated from tumor specific proviral transcripts is therefore an important future goal.
 
@@
 HERVs have been shown to have rare functional roles in healthy cells \citep{Rote2004} and could conceivably have a role in maintaining healthy colorectal epithelium, but this is speculative.
-Over expression in our healthy samples could also result from technical (i.e. sequencing protocols) or demographic (i.e. age) differences between phenotypes.
-Latent factor analysis is expected to adjust for multiple confounders, including subject age and sequencing artifacts, but the adjustment may be insufficient.
-Although surprising, the detection of high levels of HERV expression in healthy samples reduced the likelihood that using a reference HERV transcriptome assembled exclusively from contigs of cancer cells had biased our results.
+Over expression in healthy tissue may be a characteristic of the subset of HERV transcripts selected for analysis.
+Selected transcripts were previously unannotated and understudied, so their patterns of expression are largely unknown.
+Notwithstanding the biological implications of high levels of HERV expression in healthy samples, the bioinformatic implications were reassuring.
+Transcripts of repeat elements included in this study were assembled exclusively from contigs of cancer tissue \citep{Attig2019}, which might be expected to favor the detection of widespread tumor specific over expression.
+The over expression of HERVs in healthy tissue reduced the likelihood of this bias.
 
 The major limitation of this study was its reliance on heterogeneous RNA-seq datasets and quantifications of short reads of repetitive sequences that are infamously challenging to estimate accurately.
+Expression differences may have resulted from technical or demographic differences between phenotypes.
 To address these limitations, we built a complex bioinformatic pipeline based on state of the art software.
@@
 We used a published, alignment-free, k-mer matching quantification algorithm to probabilistically estimate transcript abundance of repetitive sequences that may otherwise have been excluded from analysis due to multiple possible mappings \citep{Patro2017}.
+We used latent factor analysis to adjust for multiple confounders, including subject age and sequencing artifacts across studies \citep{sva}.
 We used actively developed and updated statistical packages designed specifically for high dimensional genomics analyses to perform association tests \citep{Love2014, sva}.

### 2021-01-24 manuscript/secs/introduction.tex
first set of revisions to manuscript, excluding MAD

--- 2020-12-29 manuscript/secs/introduction.tex
@@
 Colorectal cancer (CRC) is a common malignancy in the United States.
-Early detection and effective treatment result in a 5-year relative survival of 90\% for patients diagnosed with local disease.
+Early detection and effective treatment result in a 5-year relative survival of 90\% for patients diagnosed with local disease \citep{SEER2020}.
 However, early detection relies on screening strategies in which nearly a third of adults do not participate as recommended \citep{CDC2019}.
@@
 
-Screening and treatment strategies have begun to rely on genomic biomarkers for prediction and prognosis, but novel genomic features with better sensitivity and specificity are under active investigation \citep{Imperiale2014, Sveen2020}.
-Among emerging treatment strategies, immune checkpoint inhibition is effective against metastatic tumors with mismatch repair defficiency and high microsatellite instability \citep{Le2015, Overman2018}.
-Unfortunately, only 4\% of metastatic CRCs display high microsatellite instability \citep{Ganesh2019}.
-Identification of additional classes of tumor with potential susceptibilty to immunotherapy and development of treatments to improve tumor immunogenicity are required to help extend the benefit of checkpoint blockade.
-Endogenous retroviral elements expressed in cancers may offer a solution.
+Screening and treatment strategies rely increasingly on genomic biomarkers for prediction and prognosis, and identification of novel genomic features with better sensitivity and specificity are under active investigation \citep{Imperiale2014, Sveen2020}.
+Among emerging treatment strategies, immune checkpoint inhibition is effective against metastatic tumors with mismatch repair deficiency and high microsatellite instability \citep{Le2015, Overman2018}, but only 4\% of metastatic CRCs display high microsatellite instability \citep{Ganesh2019}.
+Identification of additional classes of tumor with potential susceptibilty to immunotherapy and development of treatments to improve tumor immunogenicity would help extend the benefit of checkpoint blockade.
+Endogenous retroviral elements expressed in cancers may be of value in these efforts.
 
-Human endogenous retroviruses (HERVs) are proviral elements integrated into the human genome under conditions that render them incompetent of replication or infection.
-Residual HERV elements in contemporary human genomes are the consequence of retroviral infections of germ cells millions of years ago.
-Since then, and like the rest of the human genome, HERV elements have accumulated point mutations and segmental deletions.
-These alterations in HERV sequences disrupt assembly of functional viral machinery.
-Nevertheless, some HERV sequences can still be transcribed and translated, with some protein products even processed as endogenous antigens and presented on major histocompatibility complex class I molecules \citep{Boller1997, Rooney2015}.
-In healthy tissues, DNA methylation and histone modification can silence transcription of HERV proviruses, but epigenetic modifications in the context of oncogenesis can de-repress HERV transcription and occasionally lead to translation of neoantigens \citep{Menendez2004, Wiesner2015}.
+Human endogenous retroviruses (HERVs) are DNA elements previously integrated into the human genome.
+They are the consequence of retroviral infections of germ cells and subsequent rounds of endogenous retrotransposition.
+Over time, most HERV elements have accumulated numerous point mutations and segmental deletions.
+These alterations often disrupt viral functions, making the genomes incompetent to retrotranspose or express proteins \citep{Babaian2016}.
+Nevertheless, some HERV sequences are still capable of being transcribed and translated, and protein products from these mRNAs can be processed as endogenous antigens and presented on major histocompatibility complex class I molecules \citep{Boller1997, Mullins2012, Rooney2015}.
+In healthy tissues, DNA methylation and histone modification often silence HERV transcription, but epigenetic modifications in the context of cancer can lead to de-repression of HERV transcription and occasionally to translation of neoantigens \citep{Babaian2016, Menendez2004, Wiesner2015}.
 
-The HERV transcriptome is a promising source of biomarkers for cancer detection and treatment because of the putative cancer specificity of HERV mRNA expression.
+The HERV derived transcriptome is a promising source of novel genomic features for cancer detection and treatment because of the putative cancer specificity of HERV mRNA expression and its association with immune modulation \citep{Chiappinelli2015, Roulois2015, Desai2017, Solovyov2018}.
 A major challenge to the development of HERVs as cancer biomarkers is the absence of consensus on how to define and identify HERV elements.
-Proviruses are generally absent from human gene and transcript annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
-Proviral genomes are flanked by repetitive elements that often cannot be mapped with short read sequencing.
-Furthermore, HERVs are transposable elements that have generated many similar proviral integrations throughout the human genome.
+Proviral sequences are frequently absent from human gene and transcript annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
+Their repetitive sequences often result in bioinformatic exclusion from analysis.
+Proviral genomes are also frequently flanked by repetitive elements that often cannot be mapped using short read sequencing.
+Furthermore, HERVs were originally transposable elements that generated many similar proviral integrations throughout the human genome.
 The multitude of similar sequences across the genome impairs accurate high throughout read alignment.
-As a result, the genomic locations and transcriptional abundances of most HERVs are not known with confidence.
+As a result, the genomic locations of origin and transcriptional abundance of most HERV mRNAs remain largely unknown.
 
@@
 We hypothesized that HERV proviral elements would be preferentially transcribed in tumor samples compared to healthy and tumor adjacent, histologically normal samples.
-We tested this hypothesis in a large cohort of healthy, tumor adjacent, and tumor RNA sequencing data and found several HERV proviruses over-expressed in tumors but many that had greater expression in healthy tissue.
+We tested this hypothesis in a curated dataset of healthy, tumor adjacent, and tumor RNA sequencing (RNA-seq) reads and found several HERV proviruses over expressed in tumors and others that showed greater expression in healthy tissue.

### 2021-01-24 manuscript/secs/methods.tex
first set of revisions to manuscript, excluding MAD

@@
 Samples were originally gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a recently generated dataset of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
-A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al. \citep{Dampier2020}.
+A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al. (2020) \citep{Dampier2020}.
 RNA-seq reads for all samples were stored and analyzed in FASTQ format.
@@
 Previously unannotated HERV genes were selected for the current study in multiple steps.
-A complete gene transfer format (GTF) \citep{GTF} annotation of the repeat element inclusive cancer transcriptome was downloaded from Attig et al. \citep{Attig2019}.
+A complete gene transfer format (GTF) \citep{GTF} annotation of the repeat element inclusive cancer transcriptome was downloaded from Attig et al. (2019) \citep{Attig2019}.
 The annotation was parsed with a custom Python script to produce a new annotation in BED12 \citep{BED12} format.
-Transcripts were selected based on three criteria: i) overlap of at least one exon with a long terminal repeat element, ii) classification of at least one exon as HERV derived, and iii) abscence of any exon overlapping an annotated GENCODE gene.
+Transcripts were selected based on three criteria: i) overlap of at least one exon with a long terminal repeat element, ii) classification of at least one exon as HERV derived, and iii) absence of any exon overlapping an annotated GENCODE gene.
 All transcripts labeled with the same repeat element identifier (i.e. repeat element class along with genomic coordinates) were considered isoforms of the same HERV locus.

### 2021-01-24 manuscript/secs/results.tex
first set of revisions to manuscript, excluding MAD

@@
 
-To identify HERV proviruses with potential to function as protein coding genes and succeed as biomarkers or immunotargets, we filtered HERV loci by expression level relative to GENCODE protein coding genes.
+To identify HERV proviruses with potential to express proteins and function as possible biomarkers or immunotargets, we filtered HERV loci by expression level relative to GENCODE protein coding genes.
 Of the 1,513 unique loci detected, 150 were expressed at levels within one standard deviation of the median of canonical protein coding genes (Figure \ref{fig:pre_sum}A).
@@
 \subsection*{Proviruses associated with colorectal cancer}
-To test our hypothesis that HERV proviral elements would be preferentially transcribed in tumor samples compared to healthy and tumor adjacent samples, we regressed gene and proviral counts on tissue phenotype.
+To test our hypothesis that some HERV proviral elements would be preferentially transcribed in tumor samples compared to healthy and tumor adjacent samples, we regressed gene and proviral counts on tissue phenotype.
 Confounding factors (e.g.
@@
 
-To verify this surprising result, we tested for differential expression in the discovery cohort using an alternative HERV transcriptome recently compiled by manual curation of published but nevertheless poorly annotated HERV-K (HML-2) sequences \citep{Grabski2020}.
+To verify this result, we tested for differential expression in the discovery cohort using an alternative HERV transcriptome recently compiled by manual curation of published HERV-K (HML-2) sequences \citep{Grabski2020}.
 The alternative transcriptome included 91 proviruses.

### 2021-01-24 manuscript/secs/title.tex
first set of revisions to manuscript, excluding MAD

\title{Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer}

\author[1,2]{Christopher H. Dampier, MD}
\author[3,4]{David F. Grabski, MD}
\author[1,2]{Matthew Devall, PhD}
\author[5]{Virginia Diez-Obrero, BS}
\author[5]{Ferran Moratalla-Navarro, MS}
\author[4,6]{David Rekosh, PhD}
\author[4,6]{Marie-Louise Hammarskjold, MD, PhD}
\author[7]{Sara K. Rasmussen, MD, PhD}
\author[5,8,9,10]{Victor Moreno, MD, PhD}
\author[1,2,*]{Graham Casey, PhD}

\affil[1]{Center for Public Health Genomics, University of Virginia, Charlottesville, Virginia, USA}
\affil[2]{Department of Public Health Sciences, University of Virginia, Charlottesville, Virginia, USA}
\affil[3]{Department of Surgery, University of Virginia, Charlottesville, Virginia, USA}
\affil[4]{Thaler Center for HIV and Retrovirus Research, University of Virginia, Charlottesville, Virginia, USA}
\affil[5]{Unit of Biomarkers and Susceptibility, Oncology Data Analytics Program, Catalan Institute of Oncology, Hospitalet de Llobregat, Barcelona, Spain}
\affil[6]{Department of Microbiology, Immunology and Cancer Biology, University of Virginia, Charlottesville, Virginia, USA}
\affil[7]{Department of Surgery, Seattle Children's, Seattle, Washington, USA}
\affil[8]{Colorectal Cancer Group, ONCOBELL Program, Bellvitge Biomedical Research Institute, Hospitalet de Llobregat, Barcelona, Spain}
\affil[9]{Consortium for Biomedical Research in Epidemiology and Public Health, Spain}
\affil[10]{Department of Clinical Sciences, Faculty of Medicine, University of Barcelona, Barcelona, Spain}
\affil[*]{Correspondence: Graham Casey, PhD, Center for Public Health Genomics, MSB Room 3238, Department of Public Health Sciences, University of Virginia, P.O. Box 800717, Charlottesville, VA 22908-0717, gc8r@virginia.edu}

### 2021-01-24 manuscript/supp.tex
first set of revisions to manuscript, excluding MAD

@@
 
-    \title{Supplement to Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer}
-
-    \author[1]{Christopher H. Dampier, MD}
-    \author[2]{David F. Grabski, MD, PhD}
-    \author[1]{Matthew Devall, PhD}
-    \author[3]{Virginia Diez-Obrero, BS}
-    \author[3]{Ferran Moratalla-Navarro, MS}
-    \author[4]{David Rekosh, PhD}
-    \author[4]{Marie-Louise Hammarskjold, PhD}
-    \author[5]{Sara K. Rasmussen, MD, PhD}
-    \author[3]{Victor Moreno, MD, PhD}
-    \author[1]{Graham Casey, PhD}
-
-    \affil[1]{Center for Public Health Genomics, University of Virginia, Charlottesville, Virginia, USA}
-    \affil[2]{Department of Surgery, University of Virginia, Charlottesville, Virginia, USA}
-    \affil[3]{Catalan Institute of Oncology, Barcelona, Spain}
-    \affil[4]{Department of Microbiology, Immunology and Cancer Biology, University of Virginia, Charlottesville, Virginia, USA}
-    \affil[5]{Department of Surgery, Seattle Children's, Seattle, Washington, USA}
+    \input{sups/title.tex}
 
@@
   {
-    \includegraphics[width=0.9\textwidth]{spf1}
+    \includegraphics[width=0.8\textwidth]{spf1}
   }
@@
   {
-    \includegraphics[width=0.9\textwidth]{spf2}
+    \includegraphics[width=1.2\textwidth]{spf2}
+  }
+  \caption{Receiver operating characteristic curves in discovery cohort}
+  \label{fig:rocA}
+\end{figure}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=1.2\textwidth]{spf3}
+  }
+  \caption{Receiver operating characteristic curves in independent cohort}
+  \label{fig:rocC}
+\end{figure}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=0.6\textwidth]{spf4}
+  }
+  \caption{Receiver operating characteristic curves in second independent cohort}
+  \label{fig:rocB}
+\end{figure}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=0.8\textwidth]{spf5}
+  }
+  \caption{Relationship between subject age and HERV expression level}
+  \label{fig:age}
+\end{figure}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=1.2\textwidth]{spf6}
+  }
+  \caption{Relationship between tumor MSI status or stage and HERV expression level}
+  \label{fig:msistage}
+\end{figure}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=0.6\textwidth]{spf7}
+  }
+  \caption{Relationship between cytotoxic T cell markers and HERV expression level}
+  \label{fig:cd8}
+\end{figure}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=0.9\textwidth]{spf8}
   }

### 2021-01-24 manuscript/sups/title.tex
first set of revisions to manuscript, excluding MAD

--- 2021-01-24 manuscript/secs/title.tex
@@
-\title{Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer}
+\title{Supplement to Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer}
 
@@
 \author[5,8,9,10]{Victor Moreno, MD, PhD}
-\author[1,2,*]{Graham Casey, PhD}
+\author[1,2]{Graham Casey, PhD}
 
@@
 \affil[10]{Department of Clinical Sciences, Faculty of Medicine, University of Barcelona, Barcelona, Spain}
-\affil[*]{Correspondence: Graham Casey, PhD, Center for Public Health Genomics, MSB Room 3238, Department of Public Health Sciences, University of Virginia, P.O.
-Box 800717, Charlottesville, VA 22908-0717, gc8r@virginia.edu}

### 2021-01-24 manuscript/secs/abstract.tex
major revisions part ii

@@
 Transcripts from novel genomic features such as endogenous retroviruses have potential utility as biomarkers or therapeutic targets, and their characterization could help improve detection and treatment of colorectal cancer.
-To identify uncharacterized transcripts for possible diagnostic and therapeutic applications, we examined cancer specific over expression of unannotated endogenous retroviral mRNA in a large cohort of colorectal transcriptomes.
+To identify uncharacterized transcripts for possible diagnostic and therapeutic applications, we examined cancer specific over-expression of unannotated endogenous retroviral mRNA in a large cohort of colorectal transcriptomes.
 
@@
 Of 5,025 unannotated proviral loci queried in our discovery analysis, 150 were expressed within one standard deviation of median protein coding gene expression in at least one tissue phenotype.
-Tumor specific over expression was observed in 11 of the 150 loci tested (\emph{P} adjusted < 0.05), and seven were validated in an independent cohort.
+Tumor specific over expression was observed in 11 of the 150 loci tested (\emph{P} adjusted < 0.05), and seven loci were validated in an independent cohort.
 

### 2021-01-24 manuscript/secs/discussion.tex
major revisions part ii

@@
-In this study, we performed the largest unbiased assessment of transcribed, previously unannotated human endogenous retroviral elements in colorectal tissue to date.
-We found several unique proviral loci with tumor specific over expression and other loci with relatively high expression in healthy samples.
+In this study, we performed the largest unbiased assessment of transcribed, unannotated human endogenous retroviral elements in colorectal tissue to date.
+We found several unique proviral loci with tumor specific over-expression and other loci with relatively high expression in healthy samples.
 Human endogenous retroviral elements are often transcriptionally repressed repetitive sequences that are frequently de-repressed in epigenetically altered cancer cells \citep{Babaian2016}.
@@
 Over expression of HERV-H proviruses in CRC has been demonstrated previously with microarray \citep{Perot2012} and quantitative real time polymerase chain reaction \citep{Perot2015}.
-The prominence of HERV-H transcripts in RNA-seq from TCGA CRC samples has also been previously reported \citep{Desai2017, Attig2019}.
+The prominence of HERV-H transcripts in RNA-seq from TCGA COAD samples has also been previously reported \citep{Desai2017, Attig2019}.
 Therefore, our findings are generally consistent with prior work.
@@
 Interpretation of this result is challenging.
-HERVs have been shown to have rare functional roles in healthy cells \citep{Rote2004} and could conceivably have a role in maintaining healthy colorectal epithelium, but this is speculative.
+HERVs can have rare functional roles in healthy cells \citep{Rote2004} and could conceivably have a role in maintaining healthy colorectal epithelium, but this is speculative.
 Over expression in healthy tissue may be a characteristic of the subset of HERV transcripts selected for analysis.
@@
 Notwithstanding the biological implications of high levels of HERV expression in healthy samples, the bioinformatic implications were reassuring.
-Transcripts of repeat elements included in this study were assembled exclusively from contigs of cancer tissue \citep{Attig2019}, which might be expected to favor the detection of widespread tumor specific over expression.
-The over expression of HERVs in healthy tissue reduced the likelihood of this bias.
+Transcripts of repeat elements included in this study were assembled exclusively from contigs of cancer tissue \citep{Attig2019}, which might be expected to favor the detection of widespread tumor specific over-expression.
+The over-expression of HERVs in healthy tissue reduced the likelihood of this bias.
 
-The major limitation of this study was its reliance on heterogeneous RNA-seq datasets and quantifications of short reads of repetitive sequences that are infamously challenging to estimate accurately.
+A major limitation of this study was its reliance on heterogeneous RNA-seq datasets and quantifications of short reads of repetitive sequences that are difficult to estimate accurately.
 Expression differences may have resulted from technical or demographic differences between phenotypes.
-To address these limitations, we built a complex bioinformatic pipeline based on state of the art software.
+An additional limitation is the prominence of TCGA samples in all three cohorts studied, which could compromise independence of expression signals across cohorts despite independence of subjects.
+To address these limitations, we built a sophisticated bioinformatic pipeline based on state-of-the-art software.
 Specifically, our definition of HERV elements was of critical importance.
@@
 We used a published, alignment-free, k-mer matching quantification algorithm to probabilistically estimate transcript abundance of repetitive sequences that may otherwise have been excluded from analysis due to multiple possible mappings \citep{Patro2017}.
-We used latent factor analysis to adjust for multiple confounders, including subject age and sequencing artifacts across studies \citep{sva}.
+We used latent factor analysis to effectively adjust for multiple confounders, including subject age \citep{sva}.
 We used actively developed and updated statistical packages designed specifically for high dimensional genomics analyses to perform association tests \citep{Love2014, sva}.
@@
 Markers and targets for immune modulation are of particular relevance given the emergence of immune checkpoint inhibition as an effective treatment for advanced disease.
-The seven proviruses we identified were consistently and reliably over expressed in colorectal tumors and warrant future investigation.
+The seven proviruses we identified were consistently and reliably over-expressed in colorectal tumors and warrant future investigation.

### 2021-01-24 manuscript/secs/introduction.tex
major revisions part ii

@@
 Proviral sequences are frequently absent from human gene and transcript annotations, and their structural features make them hard to identify in agnostic surveys based on DNA or RNA sequencing \citep{ERVmap2018, Treangen2011}.
-Their repetitive sequences often result in bioinformatic exclusion from analysis.
-Proviral genomes are also frequently flanked by repetitive elements that often cannot be mapped using short read sequencing.
+Proviral genomes are frequently flanked by repetitive elements that often cannot be mapped using short read sequencing.
 Furthermore, HERVs were originally transposable elements that generated many similar proviral integrations throughout the human genome.

### 2021-01-24 manuscript/secs/methods.tex
major revisions part ii

--- 2020-12-31 manuscript/secs/methods.tex
@@
 Bulk RNA-seq from a curated set of 1,139 flash frozen colorectal tissue samples was re-processed and analyzed for the presence of unannotated HERV transcripts \citep{Dampier2020}.
-For discovery of associations between tissue phenotypes and HERV expression, paired end sequencing from 833 colon biopsies was used.
+For discovery of associations between tissue phenotypes and HERV expression, paired-end sequencing from 833 colon biopsies was used.
 Of those biopsies, 462 were from healthy, 62 were from tumor adjacent normal, and 309 were from tumor tissue.
@@
 
-For validation of discovered associations, two independent cohorts were used.
-The first (independent cohort \#1) included single end sequencing from 276 colon biopsies of which 12 were healthy, 56 were tumor adjacent, and 208 were tumor.
-The second (independent cohort \#2) included paired end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
-An important characteristic of the discovery cohort and independent cohort \#1 was the independence of all tumor and tumor adjacent samples.
-To permit valid inference from regular linear models, only one tissue phenotype was used per subject.
+For validation of discovered associations, two cohorts of samples from independent subjects were used.
+The first (independent cohort \#1) included single-end sequencing from 276 colon biopsies of which 12 were healthy, 56 were tumor adjacent, and 208 were tumor.
+The second (independent cohort \#2) included paired-end sequencing from 30 pairs of tumor and matched adjacent normal biopsies (i.e. 15 samples of both tissue phenotypes paired by subject).
+To permit valid inference from regular linear models, only one tissue phenotype was used per subject in the discovery cohort and independent cohort \#1 (i.e.
+all samples were from independent subjects).
 
 Samples were originally gathered from publicly available RNA-seq datasets hosted on the Genomic Data Commons and Sequence Read Archive as well as from a recently generated dataset of healthy mucosal biopsies obtained during screening colonoscopies at the Catalan Institute of Oncology in Barcelona, Spain \citep{Dampier2020, DiezObrero2020}.
-A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al. \citep{Dampier2020}.
-RNA-seq reads for all samples were stored and analyzed in FASTQ format.
-For samples downloaded in BAM format, Biobambam2 v2.0.87 \citep{Tischler2014} was used to convert reads to FASTQ format.
+A complete description of sample selection and cohort construction for the curated dataset is available in Dampier et al.
+(2020) \citep{Dampier2020}, and preliminary bioinformatic processing is described in the Supplementary Methods.
 
 \subsection*{Human endogenous retrovirus annotation}
-HERVs were identified using the genomic coordinates (GRCh38) and feature names of exons and spliced transcripts of HERV elements from a recently published human cancer transcriptome annotated with the GENCODE v24 basic gene model \citep{Frankish2018} and a model of genome-wide repeat elements \citep{Attig2019}.
+HERVs were identified using the genomic coordinates (GRCh38) and feature names of spliced transcripts of HERV elements from a recently published human cancer transcriptome annotated with the GENCODE v24 basic gene model \citep{Frankish2018} and a model of genome-wide repeat elements \citep{Attig2019}.
 The cancer transcriptome was assembled from RNA-seq reads of 24 subjects from each of 32 cancer types obtained from The Cancer Genome Atlas (TCGA), as previously described \citep{Attig2019}.
@@
 The method relies on profile hidden Markov models representing known human repeat families to classify repeat elements throughout the genome.
-
-Previously unannotated HERV genes were selected for the current study in multiple steps.
-A complete gene transfer format (GTF) \citep{GTF} annotation of the repeat element inclusive cancer transcriptome was downloaded from Attig et al. \citep{Attig2019}.
-The annotation was parsed with a custom Python script to produce a new annotation in BED12 \citep{BED12} format.
-Transcripts were selected based on three criteria: i) overlap of at least one exon with a long terminal repeat element, ii) classification of at least one exon as HERV derived, and iii) abscence of any exon overlapping an annotated GENCODE gene.
-All transcripts labeled with the same repeat element identifier (i.e. repeat element class along with genomic coordinates) were considered isoforms of the same HERV locus.
-Genomic coordinates were extracted for all constituent exons of selected transcripts to permit inference of splice sites and exclusion of intronic sequences.
-For display purposes, repeat element identifier genomic coordinates were converted to cytobands based on the cytoBand table downloaded from the UCSC Genome Browser \citep{Furey2003, Navarro2020}.
+Previously unannotated HERV transcripts were selected for the current study in multiple steps described in the Supplementary Methods.
 
@@
 Transcript quantification was performed with Salmon v1.2.1 \citep{Patro2017} in mapping-based mode with the \verb|--gcBias| and \verb|--validateMappings| flags set.
-Inputs to the Salmon \verb|quant| command were RNA-seq reads in FASTQ format and a custom Salmon index.
-To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34  transcript sequences \citep{Frankish2018} and the list of selected HERV transcript sequences.
-GENCODE sequences were downloaded from the GENCODE FTP site \citep{GENCODE-transcripts}.
-HERV sequences were extracted from the GRCh38 reference genome \citep{ENCODE-GRCh38} using the BEDTools \verb|getfasta| command \citep{Quinlan2010} with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts mentioned above.
-Following the recommendation of Salmon's authors \citep{SalmonDecoys}, contig sequences from the GRCh38 reference genome were subsequently added to the target transcript sequences as decoy sequences to generate a decoy-aware transcriptome for indexing.
+Inputs to the Salmon \verb|quant| command were RNA-seq reads in FASTQ format and a custom Salmon index including unannotated HERV transcripts.
+Creation of the Salmon index is described in the Supplementary Methods.
 
-\subsection*{Proviral locus and protein coding gene expression analysis}
+\subsection*{Proviral locus expression analysis}
 Associations between HERV locus expression and tissue phenotype were tested using the \emph{Bioconductor} suite of genomics analysis packages \citep{bioc} in the R statistical programming language \citep{R}.
-Specifically, \emph{tximport} \citep{Soneson2015} was used to extract transcript length and abundance estimates from Salmon output and summarize them by HERV locus and GENCODE gene for input into \emph{DESeq2} \citep{Love2014}.
-Count matrices of HERV loci and GENCODE genes were prefiltered to select loci and genes with a count of at least one read in at least half of samples in each cohort.
-Factors to normalize counts for sequencing depth and transcript length were calculated with \emph{DESeq2}.
-Latent factors of non-biological variation were estimated with \emph{sva} \citep{sva} as described in the Supplementary Methods.
-Loci and genes of interest for association testing were selected in several steps.
+Data processing and analysis steps are described in detail in the Supplementary Methods.
+Briefly, transcript expression estimates were converted to gene counts using \emph{tximport} \citep{Soneson2015}, confounding variables were estimated using \emph{sva} \citep{sva}, and expression levels were modeled using \emph{DESeq2} \citep{Love2014}.
 
-First, counts of all loci and genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \citep{Ritchie2015}.
-Second, adjusted counts of all annotated genes were filtered to retain only protein coding genes.
-Third, the median and standard deviation of protein coding gene expression were calculated separately for each tissue phenotype in each cohort.
-Finally, adjusted counts of HERV loci were filtered to retain only HERVs expressed within a single standard deviation of median protein coding gene expression in at least one tissue phenotype.
+Counts of all HERV loci with sufficient expression were modeled with phenotype and latent factors as explanatory variables.
+Due to missing information for some samples, effects of demographic and technical factors were not modeled directly but instead captured with latent factors as described in the Supplementary Methods.
+Significant associations between locus counts and tissue phenotype were identified by testing whether the absolute value of the effect of phenotype was at least two fold change for a given locus.
+Multiple hypothesis testing correction was performed using a false discovery rate of 5\%.
+All figures were produced with \emph{ggplot2} \citep{Wickham2016} as described in the Supplementary Methods.
 
-Counts of all HERV loci and protein coding genes with sufficient expression were modeled with phenotype and latent factors as explanatory variables using \emph{DESeq2}.
-Significant associations between gene count and tissue phenotype were identified by testing whether the absolute value of the effect of phenotype was at least two fold change for a given gene.
-Multiple hypothesis testing correction was performed using a false discovery rate of 5\%.
-
-All figures were produced with \emph{ggplot2} \citep{Wickham2016} as described in the Supplementary Methods.
+\subsection*{Biomarker performance assessment}
+Expression levels of select HERV loci were used as binary classifiers to distinguish tumor from non-tumor samples.
+Classification was performed by logistic regression using the \verb|glm| function in R.
+Transcript per million abundance estimates from Salmon were used as predictors of tissue phenotype.
+The performance of each classifier was assessed by calculating the area under the receiver operating characteristic curve using the \emph{pROC} \citep{pROC} package.

### 2021-01-24 manuscript/secs/results.tex
major revisions part ii

@@
-\subsection*{Identification of non-canonical human endogenous retroviral transcripts in colorectal transcriptomes}
-To define an inclusive set of non-canonical, transcribed HERV elements against which to compare human colorectal transcriptomes, we started with a recently published model of the human cancer transcriptome assembled \emph{de novo} from TCGA samples and annotated with RepeatMasker using the Dfam database \citep{Attig2019}.
-The complete annotation included over 1 million transcripts, 416,862 of which overlapped long terminal repeat sequences (i.e. the repeat sequences that flank HERV proviruses).
-We aimed to discover expression of novel, cancer-specific HERV sequences, so we selected a subset of transcripts composed of exons without GENCODE annotations, at least one of which was classifed as a HERV element.
-There were 16,336 transcripts from 5,025 unique HERV loci that met our criteria (Figure \ref{fig:pre_sum}A).
+\subsection*{Unannotated human endogenous retroviral transcripts in colorectal transcriptomes}
+To define an inclusive set of unannotated, transcribed HERV elements against which to compare human colorectal transcriptomes, we started with a recently published model of the human cancer transcriptome assembled \emph{de novo} from TCGA samples and annotated with RepeatMasker using the Dfam database \citep{Attig2019}.
+The complete annotation included over 1 million transcripts, 416,862 of which overlapped long terminal repeat sequences (i.e. the repeat sequences that typically flank HERV proviruses).
+We aimed to discover expression of novel, cancer-specific HERV sequences, so we selected a subset of transcripts composed of sequences classified as HERV elements without GENCODE annotations.
+There were 16,336 transcripts from 5,025 unique HERV loci that met our criteria (Figure \ref{fig:pre_sum}a).
 
-To estimate gene and proviral abundance in colorectal tissue, RNA-seq reads from a discovery cohort of 833 colorectal samples were quasi-mapped to a reference transcriptome composed of GENCODE annotated transcripts combined with the selected 16,336 unannotated HERV transcripts (see Methods).
+To estimate gene and proviral abundance in colorectal tissue, RNA-seq reads from a discovery cohort of 833 colorectal samples were quasi-mapped to a reference transcriptome composed of GENCODE annotated transcripts combined with the selected 16,336 unannotated HERV transcripts.
 Phenotypic and demographic characteristics of the discovery cohort are presented in Table \ref{tab:discovery}.
@@
 Transcript abundance estimates were aggregated into gene abundances for transcripts with GENCODE annotations and into locus abundances for HERV transcripts.
-We found 1,513 unique HERV loci with at least a single read detected in at least half of samples measured (Figure \ref{fig:pre_sum}A).
+We found 1,513 unique HERV loci with at least a single read detected in at least half of samples measured (Figure \ref{fig:pre_sum}a).
 These 1,513 proviruses tended to have lower expression compared to levels of mRNA from canonical protein coding genes.
-This trend was observed when all tissue phenotypes were aggregated and when tumor samples were considered separately (Figure \ref{fig:pre_sum}C,D).
+This trend was observed when all tissue phenotypes were aggregated and when tumor samples were considered separately (Figure \ref{fig:pre_sum}c,d).
 
 To identify HERV proviruses with potential to express proteins and function as possible biomarkers or immunotargets, we filtered HERV loci by expression level relative to GENCODE protein coding genes.
-Of the 1,513 unique loci detected, 150 were expressed at levels within one standard deviation of the median of canonical protein coding genes (Figure \ref{fig:pre_sum}A).
+Of the 1,513 unique loci detected, 150 were expressed at levels within one standard deviation of the median of canonical protein coding genes (Figure \ref{fig:pre_sum}a).
 These 150 moderately expressed proviruses were retained for association testing along with GENCODE protein coding genes.
-Approximately half were from HERV group K or H (Figure \ref{fig:pre_sum}B).
+Approximately half were from HERV group K or H (Figure \ref{fig:pre_sum}b).
 
 \subsection*{Proviruses associated with colorectal cancer}
-To test our hypothesis that some HERV proviral elements would be preferentially transcribed in tumor samples compared to healthy and tumor adjacent samples, we regressed gene and proviral counts on tissue phenotype.
-Confounding factors (e.g.
-age) across tissue phenotypes were modeled with latent factor anaylsis (see Methods), and association tests were adjusted by including terms for the latent factors in regression models.
+After verifying the accuracy of our expression estimates for protein coding genes (Supplementary Methods and Supplementary Figure 1a,b), we tested whether some HERV proviral elements were preferentially transcribed in tumor samples compared to healthy and tumor adjacent samples.
+To do so, we regressed proviral counts on tissue phenotype.
+Confounding factors (e.g. age) across tissue phenotypes were modeled with latent factor anaylsis, and association tests were adjusted by including terms for the latent factors in regression models.
 We compared tumor samples to healthy samples and to tumor adjacent samples in separate association tests because we had previously found evidence of a transcriptomic field effect in the setting of CRC that can mask cancer specificity of some genes \citep{Dampier2020}.
@@
 
-Out of 150 HERV loci tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (\emph{P} adjusted < 0.05) (Figure \ref{fig:dge_res}A,B, red points).
+Out of 150 HERV loci tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (\emph{P} adjusted < 0.05) (Figure \ref{fig:dge_res}a,b, red points).
 The larger number of over-expressed genes in the comparison with healthy tissue could be consistent with field effect bias but could also be due to a larger healthy tissue sample size and concomitant increased statistical power.
-We did not investigate this point further because the intersection of the tumor specific proviruses from both comparisons yielded 11 HERV loci clearly unaffected by field effect bias and expressed in a tumor specific pattern (Figure \ref{fig:dge_res}C).
-These 11 HERVs could serve as novel tumor biomarkers or as templates for translation of immunogenic peptides, so we sought to validate their tumor specificity in an independent cohort.
+We did not investigate this point further because the intersection of the tumor specific proviruses from both comparisons yielded 11 HERV loci clearly unaffected by field effect bias and expressed in a tumor specific pattern (Figure \ref{fig:dge_res}c).
+These 11 HERVs could serve as novel tumor biomarkers or as templates for translation of immunogenic peptides, so we sought to further characterize their expression patterns and validate their tumor specificity in an independent cohort.
 
-To verify the consistency of associations between proviral expression and tissue phenotype, we applied the same methods used in the discovery cohort to an independent cohort (cohort \#1) of 276 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 208 tumor samples (Supplementary Table 1).
-Of the 150 HERVs moderately expressed (i.e.
-expressed within a standard deviation of median protein coding expression) in the discovery cohort, 92 were also moderately expressed in the independent cohort.
-Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in the independent cohort (Supplementary Figure 1C).
-Nominally significant (i.e.
-\emph{P} < 0.05) tumor over-expression relative to either healthy or tumor adjacent samples was considered for the purpose of validation due to the small number of healthy samples.
-Seven of the 11 tumor specific HERVs identified in the discovery cohort were validated in the independent cohort (Table \ref{tab:val}).
+First, to confirm the ability of the 11 HERVs to distinguish tumor from non-tumor tissue, we used their expression level to classify samples in the discovery cohort.
+As expected, when tested on the same samples from which their tumor specificity was discovered, most of the HERVs had excellent performance as biomarkers (Supplementary Figure 2).
+Second, to determine whether the age difference across phenotypes was likely to have artificially inflated the apparent tumor specificity of the markers, we measured the correlation of HERV expression level with subject age.
+Although a weak positive correlation was apparent (Supplementary Figure 3a), comparison of test statistics from linear models predicting expression level from age and phenotype showed a negligible effect of age (Supplementary Figure 3b).
+Third, to assess the relationship between HERV expression and tumor characteristics, we plotted HERV expression separately for each category of tumor microsatellite instability status and stage.
+The tumor specific HERVs tended to be higher in microsatellite stable tumors with the notable exceptions of HERV-H Xp22.32 and HERV-H 20p11.23 (Supplementary Figure 4a).
+No trend in expression level relative to tumor stage was apparent (Supplementary Figure 4b).
+Finally, to determine the relationship between HERV expression and cytotoxic activity within tumors, we measured the correlation between HERV expression and the expression of six markers of cytotoxic activity.
+The correlation was negligible for most HERVs, but HERV1-I 3p22.3 had a weakly negative association (Supplementary Figure 5).
 
-To verify accuracy of bioinformatic and statistical methods, two additional validations were performed.
-The present analysis relied on a modified reference transcriptome and an updated mapping and quantification algorithm relative to a prior study of the same samples \citep{Dampier2020}.
-Consistency of relative protein coding gene expression between phenotypes in the present analysis and the prior analysis was tested.
-Pearson correlation coefficients for comparisons between tumor samples and healthy and tumor adjacent samples were 0.95 and 0.97, respectively (Supplementary Figure 1A,B).
-The high concordance of estimates of relative expression between analyses indicated the new bioinformatics techniques employed in the current analysis did not substantially alter estimates of transcript abundance.
+To verify the consistency of associations between proviral expression and tissue phenotype and to validate the utility of tumor specific HERVs as potential biomarkers, we applied the same methods used in the discovery cohort to an independent cohort (cohort \#1) of 276 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 208 tumor samples (Supplementary Table 1).
+Of the 150 HERVs moderately expressed (i.e. expressed within a standard deviation of median protein coding expression) in the discovery cohort, 92 were also moderately expressed in cohort \#1.
+Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in cohort \#1 (Supplementary Figure 1c).
+As tumor classifiers in this cohort, most of the 11 tumor specific HERVs had weaker performance (Supplementary Figure 6).
+However, HERV-H Xp22.32 and HERV1-I 3p22.3 maintained excellent discriminatory ability.
+Seven of the 11 tumor specific HERVs identified in the discovery cohort were validated as consistently tumor specific in cohort \#1 (Table \ref{tab:val}).
+Note that nominally significant (i.e.
+\emph{P} < 0.05) tumor over-expression relative to either healthy or tumor adjacent samples was considered significant for the purpose of validation due to the small number of healthy samples.
 
-Testing in the discovery cohort relied on latent factor analysis to adjust for confounding factors across tissue phenotypes (see Methods).
+To confirm the accuracy of our bioinformatic and statistical methods, additional validation testing was performed on a third set of samples.
+Testing in the discovery cohort relied on latent factor analysis to adjust for confounding factors across tissue phenotypes.
 An alternative approach is to compare pairs of samples matched by subject.
-This approach eliminates the possibility of comparisons between healthy and affected tissue but is expected to control for differences in age, sex, race, and other possible confounders more effectively.
+This approach eliminates the possibility of comparisons between healthy and affected tissue but is expected to control for demographic, technical, and other possible confounders more effectively.
 Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (see Supplementary Table 2 for cohort characteristics).
-Only 48 HERV loci were moderately expressed in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples was highly concordant (Pearson \emph{R} = 0.93) (Supplementary Figure 1).
+Only 48 HERV loci were moderately expressed in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples were highly concordant (Pearson \emph{R} = 0.93) (Supplementary Figure 1d).
 The high concordance of estimates indicated the latent factor analysis employed in the discovery cohort adequately controlled the effect of confounders.
-
-%% TO DO
-%% Discovery cohort without Attig TCGA 48 + 18 (24 COAD tumor, 24 READ tumor, 12 COAD NAT, 6 READ NAT) and GTEx 12
-%% MSI vs MSS
-%% tumors with most HERV overexpression vs lymphocyte infiltration (Shirley Liu)
 
@@
 For comparison, the number of differentially expressed genes with lower expression in healthy and tumor adjacent samples relative to tumor samples was 28 and 14, respectively.
-This result is demonstrated in Figure \ref{fig:dge_res}A,B (blue points).
-A high number of differentially expressed genes with over expression in healthy tissue was also found in independent cohort \#1 (data not shown).
+This result is demonstrated in Figure \ref{fig:dge_res}a,b (blue points).
+A high number of differentially expressed genes with over-expression in healthy tissue was also found in cohort \#1 (data not shown).
 
@@
 A total of six were differentially expressed between healthy and tumor samples (\emph{P} adjusted < 0.05).
-All but one of the six were higher in healthy samples, indicating healthy specific over expression was not an artifact of our customized HERV transcriptome (Supplementary Figure 2).
+All but one of the six were higher in healthy samples, indicating healthy specific over-expression was not an artifact of our custom HERV transcriptome (Supplementary Figure 7).

### 2021-01-24 manuscript/sups/supmethods.tex
major revisions part ii

@@
+\subsection*{Bioinformatic processing of RNA-seq reads}
+RNA-seq reads for all samples were stored and analyzed in FASTQ format.
+Reads from the Genomic Data Commons were downloaded in BAM files that included unaligned reads.
+All reads downloaded in BAM format were converted to FASTQ using Biobambam2 v2.0.87 \citep{Tischler2014}.
+
+\subsection*{Selection of unannotated HERV loci}
+Previously unannotated HERV transcripts were selected as follows.
+A complete gene transfer format (GTF) \citep{GTF} annotation of a repeat element inclusive cancer transcriptome was downloaded from Attig et al. (2019) \citep{Attig2019}.
+The annotation was parsed with a custom Python script to produce a new annotation in BED12 \citep{BED12} format.
+Transcripts were selected based on three criteria: i) at least partial overlap with a long terminal repeat element, ii) classification of at least one interval as HERV derived, and iii) absence of any interval overlapping an annotated GENCODE gene.
+All transcripts labeled with the same repeat element identifier (i.e. repeat element group along with genomic coordinates) were considered isoforms of the same HERV locus.
+Genomic coordinates were extracted for all constituent parts (labeled as exons in the GTF) of selected transcripts to permit inference of splice sites and exclusion of intervening untranscribed sequences.
+Note that we avoid use of the terms \emph{exon} and \emph{intron} in the context of HERV transcripts because the terminology is ambiguous for HERV genomes.
+For display purposes, repeat element identifier genomic coordinates were converted to cytobands based on the cytoBand table downloaded from the UCSC Genome Browser \citep{Furey2003, Navarro2020}.
+
+\subsection*{Creation of custom Salmon index}
+To create the Salmon index, a custom reference transcriptome was built from a list of GENCODE v34 transcript sequences \citep{Frankish2018} and the list of selected HERV transcript sequences.
+GENCODE sequences were downloaded from the GENCODE FTP site \citep{GENCODE-transcripts}.
+HERV sequences were extracted from the GRCh38 reference genome \citep{ENCODE-GRCh38} using the BEDTools \verb|getfasta| command \citep{Quinlan2010} with the \verb|-split| option and the custom BED12 annotation of selected HERV transcripts mentioned above.
+Following recommendations from the authors of Salmon \citep{SalmonDecoys}, contig sequences from the GRCh38 reference genome were subsequently added to the target transcript sequences as decoy sequences to generate a decoy-aware transcriptome for indexing.
+
+\subsection*{Expression analysis}
+To analyze HERV locus expression using \emph{Bioconductor} \citep{bioc}, \emph{tximport} \citep{Soneson2015} was used to extract transcript length and abundance estimates from Salmon output and summarize them by HERV locus and GENCODE gene for input into \emph{DESeq2} \citep{Love2014}.
+Count matrices of HERV loci and GENCODE genes were prefiltered to select loci and genes with a count of at least one read in at least half of samples in each cohort.
+Factors to normalize counts for sequencing depth and transcript length were calculated with \emph{DESeq2}.
+Latent factors of non-biological variation were estimated with \emph{sva} \citep{sva} as described below.
+Loci and genes of interest for association testing were selected in several steps.
+
+First, counts of all loci and genes were transformed with the variance stabilizing transformation of \emph{DESeq2} and adjusted for batch effects with \emph{limma} \citep{Ritchie2015}.
+Second, adjusted counts of all annotated genes were filtered to retain only protein coding genes.
+Third, the median and standard deviation of protein coding gene expression were calculated separately for each tissue phenotype in each cohort.
+Finally, adjusted counts of HERV loci were filtered to retain only HERVs expressed within a single standard deviation of median protein coding gene expression in at least one tissue phenotype.
+
 \subsection*{Latent factor analysis}
-To estimate the confounding influences of latent factors, including demographic variables with missing data, surrogate variable analysis was performed using the \emph{sva} package \citep{sva}.
+To estimate the confounding effects of latent factors in the cohorts of independent subjects (i.e.
+the discovery cohort and independent cohort \#1), surrogate variable analysis was performed using the \emph{sva} package \citep{sva}.
 First, the number of significant latent factors was estimated with the \verb|num.sv| function.
-For this estimation, the Buja Eyuboglu method was used because it estimated a lower, more tractable number of latent factors.
+For this estimation, the Buja Eyuboglu method was used because it estimated a lower, more tractable number of latent factors (i.e.
+three latent factors each for both the discovery cohort and independent cohort \#1).
 The estimation was performed using a full model with two terms, one for the intercept and one for tissue phenotype.
@@
 
-\subsection*{Data vizualization}
+\subsection*{Data visualization}
 Results of high dimensional data analysis were displayed in figures generated with the \emph{ggplot2} package \citep{Wickham2016}.
-Multipanel figures were assembled using the \emph{cowplot} package \cite{Wilke2020}.
+Multi-panel figures were assembled using the \emph{cowplot} package \cite{Wilke2020}.
 
-Expression levels used for vizualization were based on normalized, transformed, and adjusted counts.
-Normalization was performed with the \emph{DESeq2} package \citep{Love2014} using the \verb|estimateSizeFactors| and then \verb|counts| functions.
+Expression levels used for visualization were based on normalized, transformed, and adjusted counts.
+Normalization was performed with the \emph{DESeq2} package \citep{Love2014} using the \verb|estimateSizeFactors| and \verb|counts| functions.
 The \verb|counts| function was used with the \verb|normalized| argument set to TRUE.
@@
 These counts were also overlaid atop boxplots to represent expression levels of tumor specific HERV loci.
-Lastly, the same counts were used to calculate Z-scores to represent expression levels of all HERV loci tested.
+The same counts were used to calculate Z-scores to represent expression levels of all HERV loci tested.
 
-Effect sizes (i.e. $log2(fold change)$) used for vizualization of correlations between different analyses were calculated by \emph{DESeq2} regression models and used without shrinkage.
+Effect sizes (i.e. $log2(fold change)$) used for visualization of correlations between different analyses were calculated by \emph{DESeq2} regression models and used without shrinkage.
 Effect sizes used for volcano plots (i.e. scatter plots of $log2(fold change)$ versus $-log10(\emph{P})$) were shrunken with the \emph{ashr} package \citep{Stephens2020}.
+
+\subsection*{Assessment of expression estimates}
+The present analysis relied on a modified reference transcriptome and an updated mapping and quantification algorithm relative to a prior study of the same samples \citep{Dampier2020}.
+Consistency of relative protein coding gene expression between phenotypes in the present analysis and the prior analysis was tested by comparing fold change estimates.
+Pearson correlation coefficients for comparisons between tumor samples and healthy and tumor adjacent samples were 0.95 and 0.97, respectively (Supplementary Figure 1a,b).
+The high concordance of estimates of relative expression between analyses indicated the new bioinformatics techniques employed in the current analysis did not substantially alter estimates of transcript abundance.

### 2021-01-24 scripts/res_characterize.R

@@ -101,13 +101,13 @@ prep_charac_im2 <- function(herv_ids, markers, df, hid) {
 
 plot_charac_im <- function(df, herv_levels, hid=NULL) {
     if ("gene_name" %in% colnames(df)) {
-        ggp_title <- paste("HERV", hid, "and CD8+ Markers")
+        ggp_title <- paste("HERV", hid, "and Cytotoxic Activity")
         nrow <- 2
         ggp_theme_charac <- ggp_theme_charac +
             theme(strip.text=element_text(colour="white", face="italic"))
     } else {
         df$group <- factor(df$group, levels=herv_levels)
-        ggp_title <- "HERVs and CD8+ Index"
+        ggp_title <- "HERVs and Cytotoxic Activity Index"
         nrow <- 4
     }
     ggp_ylab <- "Marker Expression"

### 2021-01-24 scripts/xelatex.sh

@@ -15,7 +15,6 @@ spf4=$9
 spf5=${10}
 spf6=${11}
 spf7=${12}
-spf8=${13}
 spt1=${14}
 spt2=${15}
 
@@ -31,7 +30,6 @@ cp $spf4 ../manuscript/figs/spf4.pdf
 cp $spf5 ../manuscript/figs/spf5.pdf
 cp $spf6 ../manuscript/figs/spf6.pdf
 cp $spf7 ../manuscript/figs/spf7.pdf
-cp $spf8 ../manuscript/figs/spf8.pdf
 cp $spt1 ../manuscript/tabs/spt1.tex
 cp $spt2 ../manuscript/tabs/spt2.tex
 

### 2021-01-24 scripts/xelatex.sh
bug in xelatex

@@ -15,8 +15,8 @@ spf4=$9
 spf5=${10}
 spf6=${11}
 spf7=${12}
-spt1=${14}
-spt2=${15}
+spt1=${13}
+spt2=${14}
 
 cp $fig1 ../manuscript/figs/fig1.pdf
 cp $fig2 ../manuscript/figs/fig2.pdf

### 2021-01-24 manuscript/secs/results.tex
add supp table with HERV coordinates

@@
 However, HERV-H Xp22.32 and HERV1-I 3p22.3 maintained excellent discriminatory ability.
-Seven of the 11 tumor specific HERVs identified in the discovery cohort were validated as consistently tumor specific in cohort \#1 (Table \ref{tab:val}).
+Seven of the 11 tumor specific HERVs identified in the discovery cohort were validated as consistently tumor specific in cohort \#1 (Table \ref{tab:val} and Supplementary Table 3).
 Note that nominally significant (i.e.

### 2021-01-24 manuscript/tabs/spt3.tex
add supp table with HERV coordinates

% latex table generated in R 4.0.0 by xtable 1.8-4 package
% Sun Jan 24 12:10:28 2021
\begin{table}[ht]
\centering
\caption{Tumor specific HERV proviruses validated in independent cohort} 
\label{tab:valsupp}
\begingroup\footnotesize
\begin{tabular}{lllll}
 {\textbf{HERV ID}} & {\textbf{HERV}} & {\textbf{CHROM}} & {\textbf{START}} & {\textbf{STOP}} \\ 
  \hline
LTR/ERV1$|$HERV1\_I-int\~{}HERV1\_LTRc$|$3$|$34679071$|$34687025 & HERV1\_I 3p22.3 & 3 & 34679071 & 34687025 \\ 
  LTR/ERV1$|$HERVIP10F-int\~{}LTR10A$|$9$|$40746806$|$40752615 & HERVIP10F 9p11.2 & 9 & 40746806 & 40752615 \\ 
  LTR/ERV1$|$HERVIP10F-int\~{}LTR10A$|$9$|$64870413$|$64876219 & HERVIP10F 9q13 & 9 & 64870413 & 64876219 \\ 
  LTR/ERVK$|$HERVK9-int\~{}MER9a2$|$13$|$19626322$|$19631229 & HERVK9 13q12.11 & 13 & 19626322 & 19631229 \\ 
  LTR/ERV1$|$HERVH-int\~{}LTR7Y$|$13$|$109265521$|$109271116 & HERVH 13q33.3 & 13 & 109265521 & 109271116 \\ 
  LTR/ERV1$|$HERVH-int\~{}LTR7B$|$20$|$19752049$|$19756776 & HERVH 20p11.23 & 20 & 19752049 & 19756776 \\ 
  LTR/ERV1$|$HERVH-int\~{}LTR7Y$|$X$|$4540474$|$4546320 & HERVH Xp22.32 & X & 4540474 & 4546320 \\ 
   \hline
\end{tabular}
\endgroup
\end{table}

### 2021-01-24 scripts/res_val.R

@@ -175,27 +175,50 @@ build_val_df <- function(restabs, l) {
     valdf <- valdf[valdf$order, ]
     fields <- c("HERV ID", "Mean", "Log2(FC)", "SE", "Stat", "P", "P Adj", "FC", "CTL", "HERV", "o")
     colnames(valdf) <- fields
-    fields <- c("HERV", "Mean", "FC", "Log2(FC)", "SE", "Stat", "P", "P Adj", "CTL")
-    valdf <- valdf[ , fields]
+    ## main results table
+    fields1 <- c("HERV", "Mean", "FC", "Log2(FC)", "SE", "Stat", "P", "P Adj", "CTL")
+    valdf1 <- valdf[ , fields1]
     for (i in c("Mean", "Log2(FC)", "SE", "Stat", "P", "P Adj")) {
-        valdf[ , i] <- signif(valdf[ , i], 2)
+        valdf1[ , i] <- signif(valdf1[ , i], 2)
     }
-    valdf$HERV <- gsub("HERV", "", valdf$HERV)
-    valdf$HERV <- gsub("_", " ", valdf$HERV)
-    return(valdf)
+    valdf1$HERV <- gsub("HERV", "", valdf1$HERV)
+    valdf1$HERV <- gsub("_", " ", valdf1$HERV)
+    ## supplementary table
+    fields2 <- c("HERV ID", "HERV")
+    valdf2 <- valdf[ , fields2]
+    valdf2$CHROM <- unlist(
+        lapply(strsplit(valdf2[ , "HERV ID"], "[|]"), "[[", 3)
+    )
+    valdf2$START <- unlist(
+        lapply(strsplit(valdf2[ , "HERV ID"], "[|]"), "[[", 4)
+    )
+    valdf2$STOP <- unlist(
+        lapply(strsplit(valdf2[ , "HERV ID"], "[|]"), "[[", 5)
+    )
+    return(setNames(list(valdf1, valdf2), c("main", "supp")))
 }
 
 
 write_valtable <- function(valdf, target) {
     cat("Writing validated results to tex table\n")
+    lab <- "tab:val"
+    dig <- c(0, 0, 0, 0, 2, 2, 2, 2, 2, 0)
+    dis <- c("d", "s", "f", "f", "f", "f", "f", "E", "E", "s")
+    siz <- "small"
+    if ("HERV ID" %in% colnames(valdf)) {
+        lab <- "tab:valsupp"
[…]

### 2021-01-24 scripts/results.R

@@ -143,11 +143,13 @@ validation_tests <- function() {
     write_cor_val(pl, target)
 
     l <- extract_main_hervs(restabs)
-    valdf <- build_val_df(restabs, l)
+    valdfs <- build_val_df(restabs, l)
     target_dir <- Sys.getenv("table_dir")
     check_dir(target_dir)
     target <- paste0(target_dir, "herv-val-table.tex")
-    write_valtable(valdf, target)
+    write_valtable(valdfs$main, target)
+    target <- paste0(target_dir, "herv-val-table-supp.tex")
+    write_valtable(valdfs$supp, target)
     l2 <- stringent_val(l)
     return(l2)
 }

### 2021-01-24 scripts/xelatex.sh

@@ -17,6 +17,7 @@ spf6=${11}
 spf7=${12}
 spt1=${13}
 spt2=${14}
+spt3=${15}
 
 cp $fig1 ../manuscript/figs/fig1.pdf
 cp $fig2 ../manuscript/figs/fig2.pdf
@@ -32,6 +33,7 @@ cp $spf6 ../manuscript/figs/spf6.pdf
 cp $spf7 ../manuscript/figs/spf7.pdf
 cp $spt1 ../manuscript/tabs/spt1.tex
 cp $spt2 ../manuscript/tabs/spt2.tex
+cp $spt3 ../manuscript/tabs/spt3.tex
 
 printf '%s' "Adding texlive to \$PATH\n"
 export PATH=$PATH:/home/chd5n/apps/texlive/2020/bin/x86_64-linux; xelatex -v

### 2021-01-24 manuscript/supp.tex
major revisions part ii
add supp table with HERV coordinates
add supp fig legends

@@
 \input{tabs/spt2.tex}
+\input{tabs/spt3.tex}
 
@@
   {
-    \includegraphics[width=1.2\textwidth]{spf3}
+    \includegraphics[width=0.8\textwidth]{spf3}
+  }
+  \caption{Relationship between subject age and HERV expression in discovery cohort}
+  \label{fig:age}
+\end{figure}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=1.2\textwidth]{spf4}
+  }
+  \caption{Relationship between tumor MSI status or stage and HERV expression in discovery cohort}
+  \label{fig:msistage}
+\end{figure}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=0.6\textwidth]{spf5}
+  }
+  \caption{Relationship between cytotoxic activity and HERV expression in discovery cohort}
+  \label{fig:cd8}
+\end{figure}
+
+\begin{figure}[ht]
+  \makebox[\textwidth][c]
+  {
+    \includegraphics[width=1.2\textwidth]{spf6}
   }
@@
   {
-    \includegraphics[width=0.6\textwidth]{spf4}
+    \includegraphics[width=0.9\textwidth]{spf7}
   }
-  \caption{Receiver operating characteristic curves in second independent cohort}
-  \label{fig:rocB}
-\end{figure}
-
-\begin{figure}[ht]
-  \makebox[\textwidth][c]
-  {
-    \includegraphics[width=0.8\textwidth]{spf5}
-  }
-  \caption{Relationship between subject age and HERV expression level}
-  \label{fig:age}
-\end{figure}
-
-\begin{figure}[ht]
-  \makebox[\textwidth][c]
-  {
-    \includegraphics[width=1.2\textwidth]{spf6}
-  }
-  \caption{Relationship between tumor MSI status or stage and HERV expression level}
-  \label{fig:msistage}
-\end{figure}
-
-\begin{figure}[ht]
-  \makebox[\textwidth][c]
-  {
-    \includegraphics[width=0.6\textwidth]{spf7}
-  }
-  \caption{Relationship between cytotoxic T cell markers and HERV expression level}
-  \label{fig:cd8}
-\end{figure}
-
-\begin{figure}[ht]
-  \makebox[\textwidth][c]
-  {
-    \includegraphics[width=0.9\textwidth]{spf8}
-  }
-  \caption{Boxplots re-demonstrating healthy specific over expression using a novel HERV-K transcriptome}
+  \caption{Boxplots re-demonstrating healthy specific over-expression using a novel HERV-K transcriptome}
   \label{fig:grabski_method}

### 2021-01-24 manuscript/sups/suplegends.tex
add supp fig legends

@@
 HERV Study) and Cohort A of a prior analysis of the same samples, as reported in Dampier et al.
-\citep{Dampier2020} (i.e.
+(2020) \citep{Dampier2020} (i.e.
 Field Effect Study).
 Baseline expression is set by the healthy phenotype.
-Only effect sizes on genes tested in both studies are shown.
-\emph{R} = Pearson correlation coefficient. \\
+Only effect sizes on genes tested in both studies are shown. \\
 b) Scatterplot demonstrating relationship between estimates of effect sizes (i.e.
@@
 HERV Study) and Cohort A of a prior analysis of the same samples, as reported in Dampier et al.
-\citep{Dampier2020} (i.e.
+(2020) \citep{Dampier2020} (i.e.
 Field Effect Study).
@@
 Only effect sizes on loci tested in both cohorts are shown.\\
+\emph{HLT}: healthy, \emph{L2FC}: $log2(fold change)$, \emph{NAT}: normal adjacent to tumor, \emph{R}: Pearson correlation coefficient, \emph{TUM}: tumor. \\
 
-\textbf{Figure 2.} Differential expression of unannotated human endogenous retroviral loci in colorectal transcriptomes using a novel HERV-K reference transcriptome. \\
+\textbf{Figure 2.} Receiver operating characteristic curves in discovery cohort. \\
+a) Receiver operating characteristic curves for the 11 tumor specific HERVs evaluated separately as binary classifiers of healthy and tumor samples. \\
+b) Receiver operating characteristic curves for the 11 tumor specific HERVs evaluated separately as binary classifiers of tumor adjacent and tumor samples. \\
+\emph{AUC}: area under curve. \\
+
+\textbf{Figure 3.} Relationship between subject age and HERV expression in discovery cohort. \\
+a) Scatterplots demonstrating relationship between expression level and subject age for the 11 tumor specific HERVs.
+Points are overlaid with best fit lines (red) from linear models predicting expression level from subject age. \\
+b) Boxplots demonstrating value of test statistics associated with effect sizes of multiple variables from linear models predicting expression level from subject age and tissue phenotype in multiple regression for the 11 tumor specific HERVs.
+Boxes represent interquartile range (IQR), and whiskers extend to values 1.5 times the IQR.
+Boxplots are overlaid with points representing HERV loci. \\
+\emph{NAT}: normal adjacent to tumor, \emph{TUM}: tumor. \\
+
+\textbf{Figure 4.} Relationship between tumor MSI status or stage and HERV expression in discovery cohort. \\
+a) Boxplots demonstrating expression levels by tumor microsatellite instability (MSI) status of 11 proviral loci positively associated with the tumor phenotype.
+Boxes represent interquartile range (IQR), and whiskers extend to values 1.5 times the IQR.
+Boxplots are overlaid with points representing tumor samples. \\
+b) Boxplots demonstrating expression levels by tumor stage of 11 proviral loci positively associated with the tumor phenotype.
+Boxes represent interquartile range (IQR), and whiskers extend to values 1.5 times the IQR.
+Boxplots are overlaid with points representing tumor samples. \\
+\emph{MSI-H}: microsatellite instability high, \emph{MSI-L}: microsatellite instability low, \emph{MSS}: microsatellite stable. \\
+
+\textbf{Figure 5.} Relationship between cytotoxic activity and HERV expression in discovery cohort. \\
+a) Scatterplots demonstrating relationship between mean expression level of six genes associated with cytotoxic activity (i.e. cytotoxicity index) and expression levels of the 11 tumor specific HERVs.
+Points representing tumor samples are overlaid with best fit lines (red) from linear models predicting cytotoxicity index from HERV expression. \\
+b) Scatterplots demonstrating relationship between expression levels of six genes associated with cytotoxic activity (used to calculate the cytotoxicity index in panel \textbf{a}) and expression levels of the 11 tumor specific HERVs.
+Points representing tumor samples are overlaid with best fit lines (red) from linear models predicting cytotoxicity gene expression from HERV expression. \\
+\emph{GZMA}: granzyme A, \emph{GZMB}: granzyme B, \emph{GZMH}: granzyme H, \emph{GZMK}: granzyme K, \emph{GZMM}: granzyme M, \emph{PRF1}: perforin 1. \\
+
+\textbf{Figure 6.} Receiver operating characteristic curves in independent cohort. \\
+a) Receiver operating characteristic curves for the 11 tumor specific HERVs evaluated separately as binary classifiers of healthy and tumor samples. \\
+b) Receiver operating characteristic curves for the 11 tumor specific HERVs evaluated separately as binary classifiers of tumor adjacent and tumor samples. \\
+\emph{AUC}: area under curve. \\
+
+\textbf{Figure 7.} Boxplots re-demonstrating healthy specific over-expression using a novel HERV-K transcriptome. \\
 Boxplots demonstrating expression levels by phenotype of six proviral loci associated with the tumor phenotype (\emph{P} adjusted < 0.05) in samples from the discovery cohort of this analysis analyzed using an alternative proviral reference transcriptome as described in Grabski et al.
-\citep{Grabski2020}.
-Five of the six significant associations are negative.
+(2020) \citep{Grabski2020}.
+Boxes represent interquartile range (IQR), and whiskers extend to values 1.5 times the IQR.
+Boxplots are overlaid with points representing samples from each tissue phenotype. \\

### 2021-01-24 manuscript/secs/results.tex
few minor changes

--- 2021-01-24 manuscript/secs/results.tex
@@
 We did not investigate this point further because the intersection of the tumor specific proviruses from both comparisons yielded 11 HERV loci clearly unaffected by field effect bias and expressed in a tumor specific pattern (Figure \ref{fig:dge_res}c).
-These 11 HERVs could serve as novel tumor biomarkers or as templates for translation of immunogenic peptides, so we sought to further characterize their expression patterns and validate their tumor specificity in an independent cohort.
+These 11 HERVs could serve as novel tumor biomarkers or as templates for translation of immunogenic peptides, so we sought to further characterize their expression patterns in the discovery cohort.
 
-First, to confirm the ability of the 11 HERVs to distinguish tumor from non-tumor tissue, we used their expression level to classify samples in the discovery cohort.
+First, to confirm the ability of the 11 HERVs to distinguish tumor from non-tumor tissue, we used their expression level to classify samples.
 As expected, when tested on the same samples from which their tumor specificity was discovered, most of the HERVs had excellent performance as biomarkers (Supplementary Figure 2).
@@
 Finally, to determine the relationship between HERV expression and cytotoxic activity within tumors, we measured the correlation between HERV expression and the expression of six markers of cytotoxic activity.
-The correlation was negligible for most HERVs, but HERV1-I 3p22.3 had a weakly negative association (Supplementary Figure 5).
+The correlation was minimal for most HERVs, but HERV1-I 3p22.3 had a weakly negative association (Supplementary Figure 5).
 
-To verify the consistency of associations between proviral expression and tissue phenotype and to validate the utility of tumor specific HERVs as potential biomarkers, we applied the same methods used in the discovery cohort to an independent cohort (cohort \#1) of 276 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 208 tumor samples (Supplementary Table 1).
-Of the 150 HERVs moderately expressed (i.e. expressed within a standard deviation of median protein coding expression) in the discovery cohort, 92 were also moderately expressed in cohort \#1.
-Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in cohort \#1 (Supplementary Figure 1c).
+To verify the consistency of associations between proviral expression and tissue phenotype and to validate the utility of tumor specific HERVs as potential biomarkers, we applied the same methods used in the discovery cohort to an independent cohort (independent cohort \#1) of 276 colorectal samples composed of 12 healthy, 56 tumor adjacent, and 208 tumor samples (Supplementary Table 1).
+Of the 150 HERVs moderately expressed (i.e.
+expressed within a standard deviation of median protein coding expression) in the discovery cohort, 92 were also moderately expressed in independent cohort \#1.
+Estimates of relative expression between healthy and tumor samples were similar for the 92 HERVs tested in both cohorts despite the small number of healthy samples in the independent cohort (Supplementary Figure 1c).
 As tumor classifiers in this cohort, most of the 11 tumor specific HERVs had weaker performance (Supplementary Figure 6).
 However, HERV-H Xp22.32 and HERV1-I 3p22.3 maintained excellent discriminatory ability.
-Seven of the 11 tumor specific HERVs identified in the discovery cohort were validated as consistently tumor specific in cohort \#1 (Table \ref{tab:val}).
+Seven of the 11 tumor specific HERVs identified in the discovery cohort were validated as consistently tumor specific in the independent cohort (Table \ref{tab:val} and Supplementary Table 3).
 Note that nominally significant (i.e.
-\emph{P} < 0.05) tumor over-expression relative to either healthy or tumor adjacent samples was considered significant for the purpose of validation due to the small number of healthy samples.
+\emph{P} < 0.05) tumor over-expression relative to either healthy or tumor adjacent samples was considered significant for the purpose of validation due to the small number of healthy samples in the independent cohort.
 
@@
 This approach eliminates the possibility of comparisons between healthy and affected tissue but is expected to control for demographic, technical, and other possible confounders more effectively.
-Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (see Supplementary Table 2 for cohort characteristics).
+Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (independent cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (see Supplementary Table 2 for cohort characteristics).
 Only 48 HERV loci were moderately expressed in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples were highly concordant (Pearson \emph{R} = 0.93) (Supplementary Figure 1d).
@@
 The number of differentially expressed genes (\emph{P} adjusted < 0.05) with higher expression in healthy and tumor adjacent samples relative to tumor samples was 94 and 13, respectively.
-For comparison, the number of differentially expressed genes with lower expression in healthy and tumor adjacent samples relative to tumor samples was 28 and 14, respectively.
+The number of differentially expressed genes with lower expression in healthy and tumor adjacent samples relative to tumor samples was 28 and 14, respectively.
 This result is demonstrated in Figure \ref{fig:dge_res}a,b (blue points).
-A high number of differentially expressed genes with over-expression in healthy tissue was also found in cohort \#1 (data not shown).
+A high number of differentially expressed genes with over-expression in healthy tissue was also found in independent cohort \#1 (data not shown).
 

### 2021-01-25 manuscript/cover.tex
add cover letter, tweak discussion

\documentclass[11pt,letterpaper]{article}  %% 3500 words for ASO research article

    %% load packages
    \usepackage[margin=1in]{geometry}         %% document margins
    \usepackage[utf8]{inputenc}
    \usepackage{setspace}                     %% to get \doublespacing command
    \usepackage{authblk}                      %% to get \affil command
    \usepackage{parskip}                      %% space between paragraphs
    \usepackage{fontspec}                     %% for \setmainfont command, will need xelatex to compile Arial

    %% general document formatting commands
    \pagestyle{plain}                         %% page numbers in middle
    %\linespread{1.67}                        %% baseline linespread = 1.2 (1.2*1.67~2)
    \setmainfont{Arial}                       %% https://latex.org/forum/viewtopic.php?t=25998

\begin{document}

January 24, 2021 \\
\\
Kelly M. McMasters, MD, PhD \\
Editor-in-Chief \\
Annals of Surgical Oncology \\
Professor and Chair, Department of Surgery \\
University of Louisville \\
Louisville, Kentucky \\

Dear Dr. McMasters, \\

We respectfully submit our manuscript, \textbf{Human Endogenous Retroviral Transcripts Show Potential as Novel, Tumor Specific Biomarkers for Colorectal Cancer}, for consideration for publication as an original research article in \emph{Annals of Surgical Oncology}.
This manuscript has not been previously published in any form, although a limited part of the data was presented in a parallel session at the Society of Surgical Oncology Annual Meeting in 2020.
This manuscript is not under consideration for publication elsewhere in any form.

Transcripts from novel genomic features such as endogenous retroviruses have potential utility as biomarkers or therapeutic targets, and their characterization could help improve detection and treatment of colorectal cancer.
To identify uncharacterized transcripts for possible diagnostic and therapeutic applications, we examined cancer specific over-expression of unannotated endogenous retroviral mRNA in a large cohort of colorectal transcriptomes.
We identified over 100 proviral loci expressed at levels similar to canonical protein coding genes and seven that could be novel biomarkers of colorectal cancer or targets of immunotherapy.

We believe the results of our study would be of interest to your readers and would appreciate the opportunity to publish these findings in \emph{Annals of Surgical Oncology}. Thank you for considering our manuscript for publication. \\

Sincerely,\\
\\
\\
Christopher H. Dampier, MD \\
Research Fellow \\
Center for Public Health Genomics \\
University of Virginia \\
Charlottesville, Virginia \\
Email: chd5n@virginia.edu \\
Twitter: @Dampier\_CH \\

Graham Casey, PhD \\
Professor, Department of Public Health Sciences \\
Center for Public Health Genomics \\
University of Virginia \\
Charlottesville, Virginia \\
Email: gc8r@virginia.edu

\end{document}

### 2021-01-25 manuscript/secs/discussion.tex
add cover letter, tweak discussion

@@
 We found several unique proviral loci with tumor specific over-expression and other loci with relatively high expression in healthy samples.
-Human endogenous retroviral elements are often transcriptionally repressed repetitive sequences that are frequently de-repressed in epigenetically altered cancer cells \citep{Babaian2016}.
-HERV transcripts have been shown to engage pattern recognition receptors and activate interferon responses in cancer cell lines \citep{Chiappinelli2015, Roulois2015} and have been associated with tumor infiltration by T regulatory cells \citep{Desai2017}, response to immune checkpoint blockade in urothelial carcinoma, and survival in subjects with CRC \citep{Solovyov2018}.
+Human endogenous retroviral elements are often selectively de-repressed in epigenetically altered cancer cells and may be useful as tumor markers \citep{Babaian2016}.
+HERV transcripts have been shown to engage pattern recognition receptors and activate interferon responses in cancer cell lines \citep{Chiappinelli2015, Roulois2015}, and they have been negatively associated with tumor infiltration by regulatory T cells \citep{Desai2017}.
+HERV transcripts have also been positively associated with response to immune checkpoint blockade for urothelial carcinoma and negatively associated with survival without immunotherapy for CRC \citep{Solovyov2018}.
+Although a consistent relationship between HERV expression and cytotoxic activity was not apparent in our bulk RNA-seq data, we identified a subset of HERV transcripts that performed well as tumor markers and varied across MSI subtypes.
 Our results are important because the tumor specific HERV sequences we identified could be used to distinguish tumor from non-tumor tissue in screening and surveillance and could be used to predict or even modify response to immunotherapies.
@@
 Therefore, our findings are generally consistent with prior work.
-However, we provide new evidence of the reliability of tumor specific over expression by jointly analyzing a large number of samples from different studies.
+However, we provide new evidence of the reliability of tumor specific over-expression by jointly analyzing a large number of samples from different studies.
 Furthermore, our analysis was not restricted to a particular group of proviruses.
@@
 This study also demonstrated widespread expression of a subset of HERVs in healthy colorectal tissue.
-Interpretation of this result is challenging.
 HERVs can have rare functional roles in healthy cells \citep{Rote2004} and could conceivably have a role in maintaining healthy colorectal epithelium, but this is speculative.
-Over expression in healthy tissue may be a characteristic of the subset of HERV transcripts selected for analysis.
-Selected transcripts were previously unannotated and understudied, so their patterns of expression are largely unknown.
-Notwithstanding the biological implications of high levels of HERV expression in healthy samples, the bioinformatic implications were reassuring.
+Alternatively, HERV loci may undergo transcription in healthy tissue without functional consequence.
+Notwithstanding the biological implications of HERV expression in healthy samples, the bioinformatic implications were reassuring.
 Transcripts of repeat elements included in this study were assembled exclusively from contigs of cancer tissue \citep{Attig2019}, which might be expected to favor the detection of widespread tumor specific over-expression.
-The over-expression of HERVs in healthy tissue reduced the likelihood of this bias.
+The over-expression of HERVs in healthy tissue demonstrated the limited extent of this potential bias.
 
@@
 Expression differences may have resulted from technical or demographic differences between phenotypes.
-An additional limitation is the prominence of TCGA samples in all three cohorts studied, which could compromise independence of expression signals across cohorts despite independence of subjects.
+An additional limitation was the prominence of TCGA samples in all three cohorts studied, which could compromise independence of expression signals across cohorts despite independence of subjects.
 To address these limitations, we built a sophisticated bioinformatic pipeline based on state-of-the-art software.

### 2021-01-25 manuscript/secs/results.tex
add cover letter, tweak discussion

@@
 Although a weak positive correlation was apparent (Supplementary Figure 3a), comparison of test statistics from linear models predicting expression level from age and phenotype showed a negligible effect of age (Supplementary Figure 3b).
-Third, to assess the relationship between HERV expression and tumor characteristics, we plotted HERV expression separately for each category of tumor microsatellite instability status and stage.
+Third, to assess the relationship between HERV expression and tumor characteristics, we plotted HERV expression separately for each category of tumor microsatellite instability (MSI) status and stage.
 The tumor specific HERVs tended to be higher in microsatellite stable tumors with the notable exceptions of HERV-H Xp22.32 and HERV-H 20p11.23 (Supplementary Figure 4a).

### 2021-01-25 scripts/xelatex.sh

@@ -53,6 +53,8 @@ bibtex supp
 xelatex supp
 xelatex supp
 
+xelatex cover
+
 pandoc \
     --standalone \
     --from=latex \
@@ -62,3 +64,13 @@ pandoc \
     --csl=ama.csl \
     --output=docx/main.docx \
     main.tex
+
+pandoc \
+    --standalone \
+    --from=latex \
+    --to=docx \
+    --reference-doc=docx/custom-reference.docx \
+    --bibliography=main.bib \
+    --csl=ama.csl \
+    --output=docx/supp.docx \
+    supp.tex

### 2021-01-25 manuscript/secs/methods.tex
add supp table with HERV coordinates
final edits prior to submission

@@
 (2020) \citep{Dampier2020}, and preliminary bioinformatic processing is described in the Supplementary Methods.
+
+Although race was not analyzed in this study, we report that the discovery cohort included 37 Asian, 83 Black, and 660 White subjects, while the race of 53 others was not available.
+Independent cohort \#1 included 12 Asian, 4 Black, and 17 White subjects, while the race of 243 others was not available.
+Independent cohort \#2 included 7 Asian, 1 Black, and 7 White subjects.
+Although no human subjects were recruited during this study, all previous studies from which data was used followed experimental protocols approved by institutional review committees and met the guidelines of relevant governmental agencies.
 
@@
 The performance of each classifier was assessed by calculating the area under the receiver operating characteristic curve using the \emph{pROC} \citep{pROC} package.
+
+All code is available on GitHub (https://github.com/dampierch/herv).

### 2021-01-25 manuscript/secs/results.tex
final edits prior to submission

@@
 Healthy tissue donors were younger, on average, than donors of tumor and tumor adjacent specimens.
-Transcript abundance estimates were aggregated into gene abundances for transcripts with GENCODE annotations and into locus abundances for HERV transcripts.
 We found 1,513 unique HERV loci with at least a single read detected in at least half of samples measured (Figure \ref{fig:pre_sum}a).
@@
 To do so, we regressed proviral counts on tissue phenotype.
-Confounding factors (e.g. age) across tissue phenotypes were modeled with latent factor anaylsis, and association tests were adjusted by including terms for the latent factors in regression models.
+Confounding factors (e.g. age) across tissue phenotypes were modeled with latent factor analysis, and association tests were adjusted by including terms for the latent factors in regression models.
 We compared tumor samples to healthy samples and to tumor adjacent samples in separate association tests because we had previously found evidence of a transcriptomic field effect in the setting of CRC that can mask cancer specificity of some genes \citep{Dampier2020}.
-Without knowledge of whether HERV transcripts are likely to be affected by the transcriptomic field effect, we performed separate association tests and compared the results.
+Without knowledge of whether HERV transcripts are likely to be affected by the field effect, we performed separate association tests and compared the results.
 
-Out of 150 HERV loci tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (\emph{P} adjusted < 0.05) (Figure \ref{fig:dge_res}a,b, red points).
+Out of 150 HERV loci tested, 28 were over-expressed in tumor samples relative to healthy samples, and 14 were over-expressed in tumor samples relative to tumor adjacent samples (\emph{P} adjusted<0.05) (Figure \ref{fig:dge_res}a,b, red points).
 The larger number of over-expressed genes in the comparison with healthy tissue could be consistent with field effect bias but could also be due to a larger healthy tissue sample size and concomitant increased statistical power.
@@
 As expected, when tested on the same samples from which their tumor specificity was discovered, most of the HERVs had excellent performance as biomarkers (Supplementary Figure 2).
-Second, to determine whether the age difference across phenotypes was likely to have artificially inflated the apparent tumor specificity of the markers, we measured the correlation of HERV expression level with subject age.
+Second, to determine whether age differences across phenotypes were likely to have inflated the apparent tumor specificity of the markers, we measured the correlation of HERV expression with subject age.
 Although a weak positive correlation was apparent (Supplementary Figure 3a), comparison of test statistics from linear models predicting expression level from age and phenotype showed a negligible effect of age (Supplementary Figure 3b).
@@
 Note that nominally significant (i.e.
-\emph{P} < 0.05) tumor over-expression relative to either healthy or tumor adjacent samples was considered significant for the purpose of validation due to the small number of healthy samples in the independent cohort.
+\emph{P}<0.05) tumor over-expression relative to either healthy or tumor adjacent samples was considered significant for the purpose of validation due to the small number of healthy samples in the independent cohort.
 
@@
 An alternative approach is to compare pairs of samples matched by subject.
-This approach eliminates the possibility of comparisons between healthy and affected tissue but is expected to control for demographic, technical, and other possible confounders more effectively.
+This approach prevents comparisons between healthy and affected tissue but may control for demographic and technical confounders more effectively.
 Consistency of relative proviral expression between phenotypes in the discovery cohort and a second independent cohort (independent cohort \#2) of 15 tumor and 15 adjacent normal pairs was tested (see Supplementary Table 2 for cohort characteristics).
-Only 48 HERV loci were moderately expressed in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples were highly concordant (Pearson \emph{R} = 0.93) (Supplementary Figure 1d).
+Only 48 HERV loci were moderately expressed in both cohorts, but estimates of their relative expression in tumor adjacent and tumor samples were highly concordant (Pearson \emph{R}=0.93) (Supplementary Figure 1d).
 The high concordance of estimates indicated the latent factor analysis employed in the discovery cohort adequately controlled the effect of confounders.
@@
 However, a surprising finding of our survey of unannotated retroviral transcripts in colorectal transcriptomes was the unexpectedly high level of expression of HERV elements in healthy samples.
-The number of differentially expressed genes (\emph{P} adjusted < 0.05) with higher expression in healthy and tumor adjacent samples relative to tumor samples was 94 and 13, respectively.
-The number of differentially expressed genes with lower expression in healthy and tumor adjacent samples relative to tumor samples was 28 and 14, respectively.
+The numbers of differentially expressed elements (\emph{P} adjusted<0.05) with higher expression in healthy and tumor adjacent samples relative to tumor samples were 94 and 13, respectively.
+The numbers of differentially expressed elements with lower expression in healthy and tumor adjacent samples relative to tumor samples were 28 and 14, respectively.
 This result is demonstrated in Figure \ref{fig:dge_res}a,b (blue points).
-A high number of differentially expressed genes with over-expression in healthy tissue was also found in independent cohort \#1 (data not shown).
+A high number of differentially expressed elements with over-expression in healthy tissue was also found in independent cohort \#1 (data not shown).
 
@@
 The alternative transcriptome included 91 proviruses.
-A total of six were differentially expressed between healthy and tumor samples (\emph{P} adjusted < 0.05).
+A total of six were differentially expressed between healthy and tumor samples (\emph{P} adjusted<0.05).
 All but one of the six were higher in healthy samples, indicating healthy specific over-expression was not an artifact of our custom HERV transcriptome (Supplementary Figure 7).

### 2021-01-25 scripts/race_chk.R

@@ -1,0 +1,17 @@
+library(readr)
+library(xtable)
+
+data_dir <- "/scratch/chd5n/herv/annotations/"
+targets <- list()
+targets[[1]] <- paste0(data_dir, "pheno_A.tsv")
+targets[[2]] <- paste0(data_dir, "pheno_C.tsv")
+targets[[3]] <- paste0(data_dir, "pheno_B.tsv")
+
+dfs <- list()
+for (i in seq_len(3)) {
+    dfs[[i]] <- readr::read_tsv(targets[[i]])
+}
+
+lapply(dfs, fun <- function(df) {
+    summary(factor(df$race))
+})

### 2021-01-30 manuscript/sups/supmethods.tex
fix tabel legends, clarify three sample changes from field effect

@@
 RNA-seq reads for all samples were stored and analyzed in FASTQ format.
-Reads from the Genomic Data Commons were downloaded in BAM files that included unaligned reads.
+Reads from the Genomic Data Commons (GDC) were downloaded in BAM files that included unaligned reads.
 All reads downloaded in BAM format were converted to FASTQ using Biobambam2 v2.0.87 \citep{Tischler2014}.
+
+During preliminary processing of reads from the GDC, two sets of paired-end reads from tumor samples previously included in Cohort A of Dampier et al.
+(2020) \citep{Dampier2020} that would otherwise have been included in the discovery cohort of this study were found to be from formalin fixed, paraffin embedded as opposed to flash frozen tissue.
+The two sets of reads, identified by file UUIDs b7e78979-c177-427d-a627-b5f241f36232 and ff6d6688-c19c-4a7f-8058-4d1bc0249d83 for subjects TCGA-A6-2677 and TCGA-A6-2672, respectively, were excluded from the discovery cohort.
+Single-end reads from flash frozen tissue of the same tumors (file UUIDs 90832632-cf57-463b-9d08-c76975066f56 and f08dc7f4-3cc3-4743-a84e-d586d74af8d1, respectively) were added to independent cohort \#1.
+With the addition of those two tumor samples to independent cohort \#1, and in order to increase the number of tumor adjacent normal samples in the discovery cohort, single-end reads from a tumor sample for subject TCGA-A6-2684 (file UUID ad1cc824-7bbc-4883-96da-55e51d1b6f0d) were removed from independent cohort \#1, and paired-end reads from tumor adjacent normal tissue from the same subject (file UUID 18625fe4-3c19-45d9-9d7c-a295fbf83f2e) were added to the discovery cohort.
+Besides those changes in sample allocation, the cohorts are the same as those analyzed in Dampier et al. (2020) \citep{Dampier2020}.
 

### 2021-01-30 manuscript/tabs/tab_disc.tex
fix tabel legends, clarify three sample changes from field effect

@@
         \hline
-        \multicolumn{12}{l}{* missing data on subset of samples; \emph{NR} signifies not reported}
+        \multicolumn{12}{l}{* missing data on subset of samples; \emph{NR}: not reported; \emph{GDC}: Genomic Data Commons; \emph{SRA}: Sequence Read Archive; \emph{TCGA}: The Cancer Genome Atlas; \emph{HLT}: healthy; \emph{NAT}: normal adjacent to tumor; \emph{TUM}: tumor}
     \end{tabular}

### 2021-01-30 manuscript/tabs/tab_ind1.tex
fix tabel legends, clarify three sample changes from field effect

@@
         \hline
-        \multicolumn{12}{l}{* missing data on subset of samples; \emph{NR} signifies not reported} \\
+        \multicolumn{12}{l}{* missing data on subset of samples; \emph{NR}: not reported; \emph{GDC}: Genomic Data Commons; \emph{SRA}: Sequence Read Archive; \emph{TCGA}: The Cancer Genome Atlas; \emph{HLT}: healthy; \emph{NAT}: normal adjacent to tumor; \emph{TUM}: tumor} \\
     \end{tabular}
